# **Additional Information**
***

## Helpful Links

- [GEO accession GSE154386](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE154386){target="_blank"}
- [Open Case Studies Guide](https://www.opencasestudies.org/OCS_Guide/){target="_blank"}
- [CoGAPS Bioconductor package](https://bioconductor.org/packages/CoGAPS){target="_blank"}
- [PyCoGAPS GitHub repository](https://github.com/FertigLab/pycogaps){target="_blank"}
- [AnnData documentation](https://anndata.readthedocs.io/){target="_blank"}
- [Scanpy documentation](https://scanpy.readthedocs.io/){target="_blank"}

The Open Case Studies teaching approach is described in the OCS model paper [@wright2023openCaseStudies]. Scholarly, methodological, and public-health sources used in this case study are cited in the sections where they support a claim and are collected in the References section below.

## Concepts Covered

- longitudinal single-cell RNA-seq
- PBMC immune-cell composition
- non-negative matrix factorization
- Bayesian matrix factorization
- latent transcriptional programs
- identity-like and activity-like programs
- rank selection
- stability plateau
- interferon-stimulated genes
- pseudobulk directionality
- R and Python reproducibility
- included analysis files for the main lesson

## Advanced Reproduction Setup

The main lesson uses small included files so you can focus on the biological and modeling evidence. You only need the setup below if you want to rebuild large source-data, preprocessing, model, or sweep outputs.

Advanced reproduction needs two local folders:

1. the cloned case-study repository
1. an external reproduction archive containing large files that are not stored in GitHub

Download and extract the external archive into a folder outside the Git repository, such as `$HOME/cs5_reproduction_archive`. Keeping those large files outside the repository helps prevent accidental commits of raw or bulky analysis artifacts.

Use the following commands from your computer's Terminal:

```bash
# Get the case-study repository if you have not already cloned it.
git clone https://github.com/opencasestudies/ocs-bio-temporal.git
cd ocs-bio-temporal

# Point to the downloaded external reproduction archive.
mkdir -p "$HOME/cs5_reproduction_archive"
export CS5_REPRODUCTION_ARCHIVE="$HOME/cs5_reproduction_archive"

# Launch Docker with both local folders mounted.
docker run --platform linux/amd64 --rm -it \
  -p 8787:8787 \
  -e PASSWORD="Password12" \
  -v "$PWD":/home/rstudio/project:rw \
  -v "$CS5_REPRODUCTION_ARCHIVE":/home/rstudio/project/data/external/reproduction_archive:ro \
  othomas2/cs5-cogaps-runtime:0.1.0
```

The first mount sends your cloned repository into the container at `/home/rstudio/project`. The second mount sends the external archive into the container at `/home/rstudio/project/data/external/reproduction_archive`. The archive is mounted read-only (`ro`) so the original downloaded files are not modified.

Regenerated outputs should be written to a writable project folder, such as `/home/rstudio/project/GSE154386`, unless you intentionally mount another output folder.

## Optional Runtime Launcher

The direct Docker command above is the clearest way to see which folders are being mounted. The repository also includes `scripts/run_cs5_runtime.sh`, a launcher script that wraps the same idea.

Use the launcher only after you understand where the repository and external archive live on your computer:

```bash
export CS5_REPRODUCTION_ARCHIVE="$HOME/cs5_reproduction_archive"
bash scripts/run_cs5_runtime.sh
```

The launcher creates expected output folders, mounts the repository, and mounts the external archive when `CS5_REPRODUCTION_ARCHIVE` points to an existing folder.

## Advanced Reproduction Script Map

The scripts below are included so you can rebuild or inspect how the included files were produced. You do not need to run them to complete the main lesson.

::: {.table-responsive}
Rebuild goal | Scripts to inspect or run | What they are for
--- | --- | ---
Prepare source data and construct the discovery set | `scripts/prepare_gse154386_subset_r.R`; `scripts/prepare_gse154386_subset_python.py` | Download or reuse source files, import 10x data, apply preprocessing decisions, and save prepared discovery-set files.
Review rank-selection evidence | `scripts/reproduction/gse154386_make_cogaps_jobs_tsv.py`; `scripts/reproduction/gse154386_cogaps_sweep_array_singleprocess_rhino_light.sbatch`; `scripts/reproduction/gse154386_aggregate_results.py`; `scripts/reproduction/cs5_generate_revised_k_selection_report.py` | Document the sweep setup, summarize candidate ranks, and create rank-selection summaries.
Rerun the selected model | `scripts/reproduction/cs5_run_selected_model_r.R`; `scripts/reproduction/cs5_run_selected_model_python.py` | Refit the selected CoGAPS model with the external prepared input files.
Rebuild interpretation tables | `scripts/reproduction/cs5_build_k10_interpretation_layer.py` | Recreate the subject-timepoint, cell-type, top-gene, and related interpretation tables used later in the case study.
Rebuild directionality evidence | `scripts/reproduction/gse154386_pattern_directionality.py` | Estimate whether selected pattern genes are higher or lower relative to baseline.
Inspect the combined source workflow | `scripts/reproduction/gse154386_sparse_distributed_cogaps.py` | Review the longer script that connects several source-data, model, and summary steps.
:::

The script names preserve some development history. In particular, `gse154386_sparse_distributed_cogaps.py` contains a non-distributed Case Study 5 workflow even though its filename includes `distributed`.

## Advanced Reproduction: CoGAPS Model and Analysis Commands

The **Data Analysis** section shows the selected CoGAPS model call in a compact reproducibility box. Use the commands below only if you want to rebuild the larger modeling and analysis artifacts from the external reproduction archive. They are not required for the main lesson.

### Review or Rerun the Rank Sweep

The rank sweep was used to compare candidate values of `K`, random seeds, and iteration counts before the selected `K = 10` model was interpreted. The main case-study path reads the summarized rank-selection evidence in **Data Import**. The commands below show how the sweep job table and HPC launcher were organized during advanced reproduction.

```{bash}
#| eval: false
python scripts/reproduction/gse154386_make_cogaps_jobs_tsv.py \
  --k-grid 5,6,7,8,9,10,11,12,16,20,24,28,32,36,40 \
  --seeds 1,2,3,4,5 \
  --iters 2000,4000,6000,8000 \
  --out GSE154386/cogaps_sweep_singleprocess_hpc/jobs.tsv
```

```{bash}
#| eval: false
sbatch --array=1-300%15 \
  scripts/reproduction/gse154386_cogaps_sweep_array_singleprocess_rhino_light.sbatch
```

After the sweep finishes, the aggregation and revised K-selection scripts collect the run summaries and write the evidence tables used by the case study.

```{bash}
#| eval: false
python scripts/reproduction/gse154386_aggregate_results.py \
  --outdir GSE154386/cogaps_sweep_singleprocess_hpc \
  --k-grid 5,6,7,8,9,10,11,12,16,20,24,28,32,36,40 \
  --seeds 1,2,3,4,5 \
  --iters 2000,4000,6000,8000

python scripts/reproduction/cs5_generate_revised_k_selection_report.py \
  --sweep-dir GSE154386/cogaps_sweep_singleprocess_hpc \
  --outdir GSE154386/revised_k_selection_K10 \
  --chosen-k 10 \
  --chosen-seed 2
```

### Rerun the Selected `K = 10` Model

The selected model can be rerun from the prepared CoGAPS-ready inputs in the external archive. R and Python use the same selected rank, seed, and iteration count. The R script reads the cells-by-genes discovery object. The Python script reads the genes-by-cells CoGAPS input with `transposeData = False` and also reads the discovery object for metadata and orientation checks.

::: {.panel-tabset}

#### R

```{bash}
#| eval: false
export CS5_REPRODUCTION_ARCHIVE="/path/to/downloaded/reproduction_archive"

docker run --platform linux/amd64 --rm \
  -v "$PWD":/home/rstudio/project:rw \
  -v "$CS5_REPRODUCTION_ARCHIVE":/home/rstudio/project/data/external/reproduction_archive:ro \
  -w /home/rstudio/project \
  othomas2/cs5-cogaps-runtime:0.1.0 \
  Rscript scripts/reproduction/cs5_run_selected_model_r.R \
    --discovery-h5ad /home/rstudio/project/data/external/reproduction_archive/cache/gse154386_experimental_discovery_cells_x_genes.h5ad \
    --outdir /home/rstudio/project/GSE154386/cogaps_r_revised_model_K10_seed2_iter2000 \
    --k 10 \
    --seed 2 \
    --n-iter 2000 \
    --cogaps-threads 4 \
    --blas-threads 1 \
    --output-frequency 100 \
    --checkpoint-interval 500 \
    --n-snapshots 10 \
    --snapshot-phase all \
    --take-pump-samples \
    --use-sparse-opt
```

#### Python

```{bash}
#| eval: false
export CS5_REPRODUCTION_ARCHIVE="/path/to/downloaded/reproduction_archive"

docker run --platform linux/amd64 --rm \
  -v "$PWD":/home/rstudio/project:rw \
  -v "$CS5_REPRODUCTION_ARCHIVE":/home/rstudio/project/data/external/reproduction_archive:ro \
  -w /home/rstudio/project \
  othomas2/cs5-cogaps-runtime:0.1.0 \
  python scripts/reproduction/cs5_run_selected_model_python.py \
    --cogaps-input-h5ad /home/rstudio/project/data/external/reproduction_archive/cache/gse154386_experimental_discovery_genes_x_cells.h5ad \
    --discovery-h5ad /home/rstudio/project/data/external/reproduction_archive/cache/gse154386_experimental_discovery_cells_x_genes.h5ad \
    --source-script /home/rstudio/project/scripts/reproduction/gse154386_sparse_distributed_cogaps.py \
    --outdir /home/rstudio/project/GSE154386/cogaps_python_revised_model_K10_seed2_iter2000 \
    --k 10 \
    --seed 2 \
    --n-iter 2000 \
    --cogaps-threads 4 \
    --blas-threads 1 \
    --output-frequency 100 \
    --checkpoint-interval 500 \
    --n-snapshots 10 \
    --snapshot-phase all \
    --take-pump-samples \
    --use-sparse-opt
```

:::

### Rebuild the Directionality Evidence from Counts

Directionality is rebuilt after the selected model outputs are available. This script asks a different question from CoGAPS: after a pattern has been identified, are the top genes for that pattern higher or lower relative to each subject's `D0` baseline?

The **Data Analysis** section shows the main idea and a compact command excerpt. The full Docker command below is for advanced reproduction, where the large preprocessed expression object is available through the external reproduction archive.

| Argument | Value used here | What it controls |
|---|---:|---|
| `--top-n` | `25` | Tests the top 25 genes from each selected CoGAPS pattern. |
| `--baseline-timepoint` | `D0` | Uses each subject's baseline sample as the matched reference. |
| `--cohort` | `experimental` | Restricts the directionality analysis to the experimental discovery cohort. |
| `--layer` | `counts` | Uses raw counts for pseudobulk expression before CPM and log2 transformation. |
| `--min-cells` | `10` | Requires enough cells in both baseline and contrast groups before computing a contrast. |
| `--log2fc-threshold` | `0.25` | Labels a contrast as up or down only when the absolute log2 fold-change reaches this threshold. |

```{bash}
#| eval: false
export CS5_REPRODUCTION_ARCHIVE="/path/to/downloaded/reproduction_archive"

docker run --platform linux/amd64 --rm \
  -v "$PWD":/home/rstudio/project:rw \
  -v "$CS5_REPRODUCTION_ARCHIVE":/home/rstudio/project/data/external/reproduction_archive:ro \
  -w /home/rstudio/project \
  othomas2/cs5-cogaps-runtime:0.1.0 \
  python scripts/reproduction/gse154386_pattern_directionality.py \
    --sweep-outdir /home/rstudio/project/data/external/reproduction_archive \
    --run-stem cogaps_K10_seed2_iter2000 \
    --outdir /home/rstudio/project/GSE154386/pattern_directionality_revised_model_K10_seed2_iter2000 \
    --top-n 25 \
    --baseline-timepoint D0 \
    --cohort experimental \
    --layer counts \
    --min-cells 10 \
    --log2fc-threshold 0.25
```

The script writes several outputs, each with a different use in the case study:

| Output | Role |
|---|---|
| `group_expression_log2cpm.csv` | Pseudobulk expression summaries by group before contrast summaries are collapsed. |
| `gene_directionality_by_subject.csv` | Subject-level log2 fold-change contrasts for each selected top gene. |
| `gene_directionality_summary.csv` | Gene-level summaries across subjects, including `median_log2fc` and `frac_up`. |
| `pattern_directionality_summary.csv` | Pattern-level summaries across tested top genes, used for the compact Pattern3 evidence row in Data Analysis. |
| `figures/` | SVG heatmaps for all-pattern and cell-type-stratified directionality checks. |

### Inspect Full Analysis Reproduction Scripts

The script map and command blocks above are usually enough. If you want to audit the script bodies inside the rendered case study, open the collapsed blocks below. They are kept here so the core Data Analysis section can stay focused on interpretation.

```{r}
#| label: additional-info-analysis-reproduction-scripts
#| echo: false
#| results: asis
emit_script_detail <- function(path, title, language) {
  full_path <- here::here(path)
  cat("<details><summary>", title, "</summary>\n\n", sep = "")
  cat("```", language, "\n", sep = "")
  cat(readLines(full_path), sep = "\n")
  cat("\n```\n\n</details>\n\n")
}

emit_script_detail(
  "scripts/reproduction/gse154386_make_cogaps_jobs_tsv.py",
  "Sweep job-table generator",
  "python"
)
emit_script_detail(
  "scripts/reproduction/gse154386_cogaps_sweep_array_singleprocess_rhino_light.sbatch",
  "SLURM array sweep launcher",
  "bash"
)
emit_script_detail(
  "scripts/reproduction/gse154386_run_one_singleprocess.py",
  "One-run PyCoGAPS sweep runner",
  "python"
)
emit_script_detail(
  "scripts/reproduction/gse154386_aggregate_results.py",
  "Sweep aggregation script",
  "python"
)
emit_script_detail(
  "scripts/reproduction/cs5_generate_revised_k_selection_report.py",
  "Revised K-selection report generator",
  "python"
)
emit_script_detail(
  "scripts/reproduction/cs5_build_k10_interpretation_layer.py",
  "K = 10 interpretation-layer table builder",
  "python"
)
emit_script_detail(
  "scripts/reproduction/cs5_run_selected_model_r.R",
  "Selected K = 10 R CoGAPS runner",
  "r"
)
emit_script_detail(
  "scripts/reproduction/cs5_run_selected_model_python.py",
  "Selected K = 10 Python CoGAPS runner",
  "python"
)
emit_script_detail(
  "scripts/reproduction/gse154386_pattern_directionality.py",
  "Directionality analysis script",
  "python"
)
```

## Advanced Audit: Full Preprocessing Decision Reference

This table gives the parameter-level preprocessing reference behind the Core Path. You do not need to memorize it for the main lesson. Use it when you want to audit how preprocessing choices define the model input, what each choice affects, and which sensitivity issues would matter if the artifacts were regenerated.

::: {.table-responsive}
Decision | Value used | Biological or analytical reason | What it affects | Limitation or sensitivity
--- | --- | --- | --- | ---
Imported single-cell object | Labeled `SingleCellExperiment` or `AnnData` object with counts and sample metadata | Keeps expression values aligned with subject, cohort, infection day, cell, and gene labels. | Starting point for QC, normalization, annotation, and discovery-set construction. | Incorrect metadata or matrix orientation would compromise downstream biological interpretation.
Minimum genes per cell | `MIN_GENES = 200` | Cells with very few detected genes provide too little transcriptome information for learning coordinated gene-expression programs. | Which cells enter normalization and CoGAPS input preparation. | More or less stringent cutoffs could change rare-cell retention.
Minimum counts per cell | `MIN_COUNTS = 500` | Low-count cells are more likely to reflect poor RNA recovery than interpretable immune activity. | Which cells remain after QC. | Count thresholds are practical filters, not biological boundaries.
Mitochondrial fraction filter | `MAX_PCT_MT = 20` | High mitochondrial signal can indicate stressed or damaged cells, which can create technical stress-like programs. | QC cell retention. | High mitochondrial signal can be technical or biological, so this cutoff should be reported.
Sample-specific upper outliers | 99.5th percentile for high gene count and high total count flags | Extreme count profiles can represent multiplets or sample-specific technical outliers. | Removes cells that may otherwise dominate summaries. | Outlier definitions depend on each sample's distribution.
Minimum cells per gene | `MIN_CELLS_PER_GENE = 10` | Genes detected in very few cells are hard to interpret as stable temporal programs. | Gene set available for HVG selection and CoGAPS. | Very rare but biologically meaningful genes may be excluded.
Normalization | Library-size normalization followed by log transform | Cells are sequenced to different depths; normalization puts expression values on a more comparable scale. | Expression values used for marker scoring, HVG selection, and downstream summaries. | Normalization reduces technical scale differences but does not remove all batch or subject effects.
Highly variable genes | `N_HVGS = 5000` | CoGAPS should focus on genes carrying useful cell-to-cell or time-course variation, not genes that are nearly constant. | CoGAPS input gene space. | Different HVG strategies can change the patterns; the Python and R reproduction scripts use related but not identical implementations.
Timepoint merge | Experimental `D14` and `D15` are merged to `D14/15` | Nearby late experimental days are combined to support a stable late-timepoint summary. | Timepoint labels used in figures and research-question tables. | The merge improves interpretability but removes a one-day distinction.
Broad cell-type annotation | Marker-set scores for broad immune-cell identities | PBMC identity is a major source of expression variation; broad labels help separate identity-like patterns from activity-like programs. | Composition summaries and identity/activity pattern classification. | Broad labels are useful for interpretation but are not fine-grained cell-state annotations.
Discovery cohort | Experimental cohort only | The experimental infection cohort has aligned days, making it the appropriate primary time course for temporal program discovery. | CoGAPS model input for the core analysis. | Natural infection samples are not used as the primary discovery model here.
Discovery-set balancing | Up to `600` cells per sample | Balancing reduces the chance that one sample or timepoint dominates the learned patterns. | Number of discovery cells used for CoGAPS. | The cap and seed define which cells contribute to model fitting and should be reported.
Random seed | `seed = 42` for subset construction | Makes the balanced discovery subset reproducible. | Which cells are selected when a sample has more than 600 cells. | A different seed could select a slightly different subset.
Natural cohort | Reserved as an optional projection target | Keeps the primary lesson focused on the aligned experimental time course. | Optional extension, not required for the main case study. | Projection requires separate validation before being treated as a main result.
:::

## Advanced Reproduction: Source-Import Code Examples

**Data Wrangling and Preprocessing** now shows the connected map from raw {{< glossary "10x Genomics files" display="10x Genomics files" >}} to a labeled {{< glossary "Single-cell object" display="single-cell object" >}}, then from that object to CoGAPS-ready inputs and included interpretation tables. This advanced reproduction note keeps the optional code examples for the first part of that map: source-file import.

You do not need these steps for the main lesson. Use them if you want to audit or rebuild how raw source files become a labeled single-cell object before preprocessing begins.

:::{.disclaimer_box}
::::{.disclaimer_box_header}
Advanced reproduction note
::::
::::{.inner_block}
The short toy chunks in this subsection are examples you can run to understand the import logic. They use small in-memory objects so you can see how source 10x files become a labeled single-cell object without downloading the full GEO archive.

If you want to run the advanced reproduction workflow on the real source files, use the preparation scripts listed in the **Advanced Reproduction Script Map** above. Those scripts include both the import steps in this section and the preprocessing steps summarized in Data Wrangling and Preprocessing. Keeping those steps in executable scripts makes the workflow easier to rerun because the downloaded files, imported objects, quality-control decisions, and saved outputs are handled in one continuous run. Helper chunks marked with `eval: false` below are script excerpts for transparency, not standalone workflow steps.
::::
:::

The key idea is alignment: a raw 10x sample is not one tidy table, but a linked matrix, barcode list, and feature list. In the steps below, you will see how those linked files become a labeled {{< glossary "Single-cell object" display="single-cell object" >}}, with expression values connected to the correct cells, genes, samples, subjects, and infection days. Pay special attention to {{< glossary "Matrix orientation" display="matrix orientation" >}}, because raw 10x matrices are usually stored as features by cells before being converted into the {{< glossary "Cell-by-gene matrix" display="cell-by-gene" >}} view used later.

### Identify Source Files and Sample Metadata

Before expression values can be interpreted biologically, each cell has to be connected back to its sample. In this dataset, the sample name tells you which subject the cell came from, whether that subject was in the experimental or natural infection cohort, and which infection day the sample represents.

The raw GEO archive contains sample-level 10x files. In this context, "sample-level" means that the raw sequencing output is organized one biological sample at a time, rather than as one combined study-wide table. For each sample, the 10x Genomics files are usually a small set of linked files: a sparse expression matrix, a list of cell barcodes, and a list of gene or feature names. These files have to be read together because the matrix alone only contains numbers; the barcode and feature files tell you which cells and genes those numbers refer to.

The goal is to make implicit information explicit. A **variable** stores a value you want to reuse, such as a source URL or file path. A **function** is reusable code with inputs and outputs. Here, the function `parse_sample_key()` takes one sample name as input and returns metadata columns as output. Later, those metadata columns are copied onto every cell from that sample.

### Step 1: Record Where the Raw Archive Comes From

This first short chunk does not download anything. It gives a name to the GEO source URL and to the local file name that would store the source archive. Naming these values makes the advanced reproduction workflow easier to audit because the source location is written once and reused.

::: {.panel-tabset}

#### R

```{r}
#| label: source-archive-r
RAW_TAR_URL <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE154386&format=file"
RAW_TAR <- file.path("data", "external", "GSE154386", "GSE154386_RAW.tar")

data.frame(
  item = c("GEO source archive", "local archive path"),
  value = c(RAW_TAR_URL, RAW_TAR)
) |>
  knitr::kable()
```

#### Python

```{python}
#| label: source-archive-python
from pathlib import Path
import pandas as pd

RAW_TAR_URL = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE154386&format=file"
RAW_TAR = Path("data/external/GSE154386/GSE154386_RAW.tar")

pd.DataFrame({
    "item": ["GEO source archive", "local archive path"],
    "value": [RAW_TAR_URL, str(RAW_TAR)],
})
```

:::

The archive is a source file, not an analysis table. It contains raw 10x files that still need to be connected to sample labels before quality control or modeling can happen.

### Step 2: Inspect the Structure of One Sample Name

The source filenames are not arbitrary. A name such as `GSM4670192_Subject2_D0` can be read as three pieces: a GEO sample identifier, a subject label, and a day label. The experimental samples use labels such as `Subject2`, while natural infection samples use labels such as `Natural1`.

::: {.panel-tabset}

#### R

```{r}
#| label: sample-key-examples-r
sample_key_examples <- data.frame(
  sample_key = c(
    "GSM4670192_Subject2_D0",
    "GSM4670216_Natural1_Dneg3"
  ),
  meaning = c(
    "experimental subject 2 at day 0",
    "natural infection sample 1 at day -3"
  )
)

sample_key_examples |>
  knitr::kable()
```

#### Python

```{python}
#| label: sample-key-examples-python
sample_key_examples = pd.DataFrame({
    "sample_key": [
        "GSM4670192_Subject2_D0",
        "GSM4670216_Natural1_Dneg3",
    ],
    "meaning": [
        "experimental subject 2 at day 0",
        "natural infection sample 1 at day -3",
    ],
})

sample_key_examples
```

:::

The label `D0` means day 0. A label such as `Dneg3` means day -3. The numeric form of the day will matter later because it lets you sort timepoints and summarize pattern scores along the infection time course.

### Step 3: Write a Function That Parses a Sample Name

The next snippet defines `parse_sample_key()`. In programming, a function is useful when the same operation needs to be repeated many times. The input is one sample name. The output is a small set of metadata fields that can be attached to cells from that sample.

The main programming tools are a regular expression and an `if` statement. A regular expression is a pattern used to find pieces of text. Here, it captures the GEO identifier, the subject label, and the day label. The `if` statement then translates the subject label into a cohort label and translates the day label into a number.

::: {.panel-tabset}

#### R

```{r}
#| label: parse-sample-key-r
parse_sample_key <- function(sample_key) {
  match <- regexec(
    "^(GSM[0-9]+)_([^_]+)_(D(?:neg)?[0-9]+)$",
    sample_key,
    perl = TRUE
  )
  parts <- regmatches(sample_key, match)[[1]]
  if (length(parts) == 0) {
    stop("Could not parse sample key: ", sample_key)
  }

  gsm_id <- parts[[2]]
  subject <- parts[[3]]
  tp <- parts[[4]]

  cohort <- if (startsWith(subject, "Subject")) {
    "experimental"
  } else if (startsWith(subject, "Natural")) {
    "natural"
  } else {
    "unknown"
  }

  day <- if (startsWith(tp, "Dneg")) {
    -as.integer(sub("Dneg", "", tp))
  } else {
    as.integer(sub("D", "", tp))
  }

  data.frame(
    gsm_id = gsm_id,
    sample_id = gsm_id,
    subject = subject,
    cohort = cohort,
    timepoint_raw = tp,
    timepoint_display = paste0("D", day),
    day_numeric = day,
    stringsAsFactors = FALSE
  )
}
```

#### Python

```{python}
#| label: parse-sample-key-python
import re

def parse_sample_key(sample_key: str) -> dict:
    match = re.match(r"^(GSM\d+)_([^_]+)_(D(?:neg)?\d+)$", sample_key)
    if not match:
        raise ValueError(f"Could not parse sample key: {sample_key}")

    gsm_id, subject, tp = match.groups()

    if subject.startswith("Subject"):
        cohort = "experimental"
    elif subject.startswith("Natural"):
        cohort = "natural"
    else:
        cohort = "unknown"

    day = -int(tp.replace("Dneg", "")) if tp.startswith("Dneg") else int(tp.replace("D", ""))

    return {
        "gsm_id": gsm_id,
        "sample_id": gsm_id,
        "subject": subject,
        "cohort": cohort,
        "timepoint_raw": tp,
        "timepoint_display": f"D{day}",
        "day_numeric": day,
    }
```

:::

In R, the parsed metadata becomes a data frame. In Python, it becomes a dictionary, which can then be placed into a `pandas` table. In both languages, the goal is the same: every cell must carry the biological labels needed for subject-level and timepoint-level summaries.

### Step 4: Test the Function on Two Sample Names

When you write a function, test it on a small example before using it on every file in a dataset. This is a coding habit, but it is also a biological check. If the day or cohort is parsed incorrectly here, every later time-course summary will be mislabeled.

::: {.panel-tabset}

#### R

```{r}
#| label: parse-sample-key-demo-r
parsed_examples <- do.call(
  rbind,
  lapply(sample_key_examples$sample_key, parse_sample_key)
)

parsed_examples |>
  knitr::kable()
```

#### Python

```{python}
#| label: parse-sample-key-demo-python
parsed_examples = pd.DataFrame([
    parse_sample_key(sample_key)
    for sample_key in sample_key_examples["sample_key"]
])

parsed_examples
```

:::

Read the output from left to right. The `gsm_id` and `sample_id` identify the GEO sample. The `subject` and `cohort` columns tell you which biological group the sample belongs to. The `timepoint_display` column is a readable label for plotting, while `day_numeric` is the numeric version used for sorting and summarizing days.

:::{.think_question}
::::{.think_question_header}
Checkpoint
::::
::::{.inner_block}
If the sample name is `GSM4670198_Subject2_D15`, what should the cohort and numeric day be?

<details>
<summary>Show answer</summary>

The cohort should be `experimental` because the subject label begins with `Subject`. The numeric day should be `15` because the timepoint label is `D15`.

</details>
::::
:::

<details>
<summary>Optional helper: download the GEO archive safely</summary>

The complete preparation scripts include a small download helper. This helper is kept separate from the parsing lesson because downloading is not the main biological idea here. It is still useful for advanced reproduction because it skips an existing archive and avoids treating the tar file as a text table.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
download_if_needed <- function(url, dest) {
  if (file.exists(dest) && file.info(dest)$size > 0) {
    message("[skip] ", dest, " already exists")
    return(invisible(dest))
  }

  download.file(url, destfile = dest, mode = "wb", quiet = FALSE)
  invisible(dest)
}
```

#### Python

```{python}
#| eval: false
import requests

def download(url: str, dest: Path, chunk_mb: int = 16) -> None:
    if dest.exists() and dest.stat().st_size > 0:
        print(f"[skip] {dest} already exists")
        return

    tmp = dest.with_suffix(dest.suffix + ".part")
    with requests.get(url, stream=True, timeout=180) as response:
        response.raise_for_status()
        with open(tmp, "wb") as handle:
            for chunk in response.iter_content(chunk_size=chunk_mb * 1024 * 1024):
                if chunk:
                    handle.write(chunk)
    tmp.rename(dest)
```

:::

</details>

After this step, the workflow has not changed any expression values yet. The important result is that each future cell-level observation can carry subject, cohort, and day labels. Without those labels, the later CoGAPS pattern scores could not be summarized by timepoint or subject.

### Build the Cell-by-Gene Data Object

In the previous step, you learned how a sample name becomes metadata. Now you will see how the raw expression files become a labeled {{< glossary "Cell-by-gene matrix" display="cell-by-gene" >}} object. The goal is not just to read files; the goal is to preserve the relationship between counts, cells, genes, and sample metadata.

Single-cell RNA-seq data do not begin as a tidy table. For each 10x Genomics sample, the data are split into three linked files. None of these files is biologically interpretable on its own.

| 10x file | What it contains | Why it matters |
|---|---|---|
| `matrix.mtx` | The expression count matrix. In raw 10x output, rows are usually features and columns are cells. | Stores the measured RNA counts, but not the cell or gene names. |
| `barcodes.tsv` | Cell barcode labels. | Tells you which captured cell or droplet each matrix column represents. |
| `features.tsv` or `genes.tsv` | Gene IDs, gene symbols, and sometimes feature types. | Tells you which gene or measured feature each matrix row represents. |

This alignment is biologically important. If genes and cells are mislabeled or transposed, marker scoring and CoGAPS interpretation can point to the wrong biological signal. The code below uses a small toy example so you can see the alignment logic without downloading the full GEO archive.

The R and Python workflows use different object classes for the same basic purpose. They are different containers, but they solve the same problem: they keep the count matrix connected to cell and gene annotations.

| Concept | R `SingleCellExperiment` | Python `AnnData` |
|---|---|---|
| Expression matrix | `assays(sce)$counts` | `adata.X` |
| {{< glossary "Cell metadata" display="Cell metadata" >}} | `colData(sce)` | `adata.obs` |
| {{< glossary "Gene metadata" display="Gene metadata" >}} | `rowData(sce)` | `adata.var` |
| Cell names | `colnames(sce)` | `adata.obs_names` |
| Gene names | `rownames(sce)` | `adata.var_names` |

### Step 1: Represent the Three Linked 10x Pieces

In the advanced reproduction script, this step reads `matrix.mtx`, `barcodes.tsv`, and `features.tsv` from disk. Here, the same idea is shown with a small in-memory example. The matrix, barcode table, and feature table are separate objects at first.

::: {.panel-tabset}

#### R

```{r}
#| label: toy-10x-pieces-r
toy_counts_10x <- matrix(
  c(1, 0,
    3, 2,
    0, 4,
    7, 1),
  nrow = 4,
  byrow = TRUE
)

toy_barcodes <- c("AAAC-1", "TTGG-1")

toy_features <- data.frame(
  gene_id = c("ENSG1", "ENSG2", "ENSG3", "AB1"),
  gene_symbol = c("ISG15", "MX1", "CD3D", "TotalSeq_B"),
  feature_type = c(
    "Gene Expression",
    "Gene Expression",
    "Gene Expression",
    "Antibody Capture"
  )
)

data.frame(
  object = c("toy_counts_10x", "toy_barcodes", "toy_features"),
  rows = c(nrow(toy_counts_10x), length(toy_barcodes), nrow(toy_features)),
  columns = c(ncol(toy_counts_10x), NA, ncol(toy_features))
) |>
  knitr::kable()
```

#### Python

```{python}
#| label: toy-10x-pieces-python
import numpy as np

toy_counts_10x = np.array([
    [1, 0],
    [3, 2],
    [0, 4],
    [7, 1],
])

toy_barcodes = ["AAAC-1", "TTGG-1"]

toy_features = pd.DataFrame({
    "gene_id": ["ENSG1", "ENSG2", "ENSG3", "AB1"],
    "gene_symbol": ["ISG15", "MX1", "CD3D", "TotalSeq_B"],
    "feature_type": [
        "Gene Expression",
        "Gene Expression",
        "Gene Expression",
        "Antibody Capture",
    ],
})

pd.DataFrame({
    "object": ["toy_counts_10x", "toy_barcodes", "toy_features"],
    "rows": [toy_counts_10x.shape[0], len(toy_barcodes), len(toy_features)],
    "columns": [toy_counts_10x.shape[1], None, toy_features.shape[1]],
})
```

:::

In this toy example, the count matrix has four feature rows and two cell columns. The feature table also has four rows, and the barcode list has two labels. Those matching dimensions are the first sign that the pieces can be aligned.

### Step 2: Check Orientation and Dimensions

Raw 10x matrices are usually stored as features by cells. That {{< glossary "Matrix orientation" display="matrix orientation" >}} means rows match the feature table and columns match the barcode table. Many downstream analyses are described as cell-by-gene analyses, so the workflow needs to know when a {{< glossary "Transpose" display="transpose" >}} is required. A transpose flips a matrix so rows become columns and columns become rows.

::: {.panel-tabset}

#### R

```{r}
#| label: toy-10x-dim-check-r
data.frame(
  check = c("feature rows match matrix rows", "barcode labels match matrix columns"),
  expected = c(nrow(toy_features), length(toy_barcodes)),
  observed = c(nrow(toy_counts_10x), ncol(toy_counts_10x)),
  passes = c(
    nrow(toy_features) == nrow(toy_counts_10x),
    length(toy_barcodes) == ncol(toy_counts_10x)
  )
) |>
  knitr::kable()
```

#### Python

```{python}
#| label: toy-10x-dim-check-python
pd.DataFrame({
    "check": [
        "feature rows match matrix rows",
        "barcode labels match matrix columns",
    ],
    "expected": [len(toy_features), len(toy_barcodes)],
    "observed": [toy_counts_10x.shape[0], toy_counts_10x.shape[1]],
    "passes": [
        len(toy_features) == toy_counts_10x.shape[0],
        len(toy_barcodes) == toy_counts_10x.shape[1],
    ],
})
```

:::

These checks look simple, but they protect the biological interpretation. If the feature count does not match the matrix rows, gene labels could be attached to the wrong counts. If the barcode count does not match the matrix columns, cell metadata could be attached to the wrong cells.

### Step 3: Keep the Gene-Expression Rows

Some 10x files include feature types that are not RNA gene-expression measurements. This case study models transcript abundance, so the import workflow keeps `Gene Expression` rows and drops other feature types before CoGAPS input preparation.

::: {.panel-tabset}

#### R

```{r}
#| label: toy-gene-expression-filter-r
keep_gene_expression <- toy_features$feature_type == "Gene Expression"

toy_counts_gene_expression <- toy_counts_10x[keep_gene_expression, , drop = FALSE]
toy_features_gene_expression <- toy_features[keep_gene_expression, , drop = FALSE]

data.frame(
  retained_gene_symbols = toy_features_gene_expression$gene_symbol,
  retained_feature_type = toy_features_gene_expression$feature_type
) |>
  knitr::kable()
```

#### Python

```{python}
#| label: toy-gene-expression-filter-python
keep_gene_expression = toy_features["feature_type"].eq("Gene Expression")

toy_counts_gene_expression = toy_counts_10x[keep_gene_expression.to_numpy(), :]
toy_features_gene_expression = toy_features.loc[keep_gene_expression, :].copy()

toy_features_gene_expression.loc[
    :, ["gene_symbol", "feature_type"]
].rename(columns={
    "gene_symbol": "retained_gene_symbols",
    "feature_type": "retained_feature_type",
})
```

:::

The dropped toy feature, `TotalSeq_B`, represents the kind of non-RNA feature that should not be mixed into this RNA expression model. In the real dataset, this filtering step keeps the biological question focused on gene-expression programs.

### Step 4: Name Cells and Genes Before Constructing the Object

Cell barcodes are only guaranteed to be unique within a sample. The same barcode sequence can appear in different samples, so the workflow prefixes each barcode with the sample ID. Gene symbols are easier for humans to interpret than gene IDs, but duplicated gene symbols can occur, so the workflow makes gene names unique.

::: {.panel-tabset}

#### R

```{r}
#| label: toy-name-cells-genes-r
toy_meta <- parse_sample_key("GSM4670192_Subject2_D0")

toy_gene_names <- make.unique(
  toy_features_gene_expression$gene_symbol,
  sep = "_"
)
toy_cell_names <- paste0(toy_meta$sample_id, ":", toy_barcodes)

data.frame(
  cell_name = toy_cell_names,
  sample_id = toy_meta$sample_id,
  subject = toy_meta$subject,
  timepoint = toy_meta$timepoint_display
) |>
  knitr::kable()
```

#### Python

```{python}
#| label: toy-name-cells-genes-python
toy_meta = parse_sample_key("GSM4670192_Subject2_D0")

toy_gene_names = pd.Index(
    toy_features_gene_expression["gene_symbol"]
).astype(str).tolist()
toy_cell_names = [
    f"{toy_meta['sample_id']}:{barcode}"
    for barcode in toy_barcodes
]

pd.DataFrame({
    "cell_name": toy_cell_names,
    "sample_id": toy_meta["sample_id"],
    "subject": toy_meta["subject"],
    "timepoint": toy_meta["timepoint_display"],
})
```

:::

This is where the metadata from the previous subsection starts to matter. The count matrix has measurements, but the cell names and metadata make those measurements traceable to a subject and infection day.

### Step 5: Build a Labeled Cell-by-Gene Representation

The raw 10x matrix is features by cells. For a {{< glossary "Cell-by-gene matrix" display="cell-by-gene" >}} view, transpose the retained matrix so cells are rows and genes are columns. The small table below shows what the aligned expression data look like after naming cells and genes.

::: {.panel-tabset}

#### R

```{r}
#| label: toy-cell-by-gene-r
toy_cell_by_gene <- t(toy_counts_gene_expression)
rownames(toy_cell_by_gene) <- toy_cell_names
colnames(toy_cell_by_gene) <- toy_gene_names

toy_cell_by_gene_display <- data.frame(
  cell_name = rownames(toy_cell_by_gene),
  as.data.frame(toy_cell_by_gene),
  check.names = FALSE
)

toy_cell_by_gene_display |>
  knitr::kable()
```

#### Python

```{python}
#| label: toy-cell-by-gene-python
toy_cell_by_gene = pd.DataFrame(
    toy_counts_gene_expression.T,
    index=toy_cell_names,
    columns=toy_gene_names,
)

toy_cell_by_gene.reset_index().rename(columns={"index": "cell_name"})
```

:::

In the advanced workflow, this aligned matrix is stored in a {{< glossary "Single-cell object" display="single-cell container" >}} rather than as a plain table. In R, the container is {{< glossary "SingleCellExperiment" display="SingleCellExperiment" >}}; in Python, it is {{< glossary "AnnData" display="AnnData" >}}. The container keeps the expression matrix, cell metadata, and gene metadata together when later steps filter cells, select genes, score markers, or save outputs.

<details>
<summary>Optional: construct the full single-cell container</summary>

The code below shows the final construction step used in the complete preparation scripts. It is shown for transparency, but the key concept is the same one you saw in the toy table: counts, cell labels, gene labels, and metadata must stay aligned.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
SingleCellExperiment::SingleCellExperiment(
  assays = list(counts = mat),
  rowData = S4Vectors::DataFrame(gene_id = gene_ids[keep]),
  colData = S4Vectors::DataFrame(meta[rep(1L, ncol(mat)), , drop = FALSE])
)
```

#### Python

```{python}
#| eval: false
from anndata import AnnData

adata = AnnData(X=matrix.T.tocsr())
adata.obs_names = [
    f"{meta['sample_id']}:{barcode}"
    for barcode in barcodes
]
adata.var_names = pd.Index([
    gene
    for gene, keep_gene in zip(gene_symbols, keep)
    if keep_gene
])
adata.var_names_make_unique()
```

:::

</details>

:::{.think_question}
::::{.think_question_header}
Checkpoint
::::
::::{.inner_block}
If a raw 10x matrix has 20,000 feature rows and 800 cell columns, how many barcode labels should there be?

<details>
<summary>Show answer</summary>

There should be 800 barcode labels because barcodes label cells, and cells are columns in the raw 10x matrix before it is transposed into a cell-by-gene view.

</details>
::::
:::

After this step, the source-file import note has produced a labeled single-cell object. Biologically, that means every expression measurement can be traced back to a cell, a gene, and a sample. In code, later filtering and summarizing can use the same object instead of passing separate matrices and metadata tables around by hand.

<details>
<summary>Optional: file-reading helper functions used above</summary>

The complete preparation scripts use small helper functions to read compressed 10x matrix and table files. These helpers are file-reading utilities, not biological decisions. They are shown here for transparency, but the important concept is that the matrix, barcode table, and feature table are read together and checked before they are aligned into one labeled single-cell object.

The `parse_sample_key()` function is defined in the previous subsection. The helpers below handle the remaining file-reading steps.

**R**

```{r}
#| eval: false
open_text_file <- function(path) {
  # 10x files are often gzip-compressed. This helper opens either a
  # compressed or uncompressed text file.
  if (grepl("\\.gz$", path)) {
    gzfile(path, open = "rt")
  } else {
    file(path, open = "rt")
  }
}

open_binary_file <- function(path) {
  # Matrix Market files are read as binary connections.
  if (grepl("\\.gz$", path)) {
    gzfile(path, open = "rb")
  } else {
    file(path, open = "rb")
  }
}

read_table_gz <- function(path) {
  # Read a barcode or feature table with no header row.
  con <- open_text_file(path)
  on.exit(close(con), add = TRUE)

  readr::read_tsv(
    con,
    col_names = FALSE,
    show_col_types = FALSE
  )
}

read_mtx_gz <- function(path) {
  # Read a sparse Matrix Market expression matrix.
  con <- open_binary_file(path)
  on.exit(close(con), add = TRUE)

  Matrix::readMM(con)
}
```

**Python**

```{python}
#| eval: false
import gzip
import pandas as pd
from scipy import sparse
from scipy.io import mmread

def read_text_table(path):
    # Read a barcode or feature table with no header row.
    opener = gzip.open if path.suffix == ".gz" else open
    with opener(path, "rt") as handle:
        return pd.read_csv(handle, sep="\t", header=None)

def read_mtx(path):
    # Read a sparse Matrix Market expression matrix.
    opener = gzip.open if path.suffix == ".gz" else open
    with opener(path, "rb") as handle:
        matrix = mmread(handle)
    return sparse.csr_matrix(matrix)
```

</details>

The Data Wrangling section connects this source-import step to preprocessing decisions, CoGAPS-ready model inputs, and the small interpretation tables used in the core path. The complete executable preparation scripts are listed in the **Advanced Reproduction Script Map** above.

## Advanced Software Reference

The main setup section lists the packages used for the small included files. The table below adds software used when rebuilding larger workflow outputs.

::: {.table-responsive}
Package or tool | Language | Main lesson or rebuild use
--- | --- | ---
[`here`](https://here.r-lib.org/){target="_blank"} | R | Build project-relative paths.
[`readr`](https://readr.tidyverse.org/){target="_blank"} | R | Read CSV tables.
[`dplyr`](https://dplyr.tidyverse.org/){target="_blank"} | R | Select, filter, summarize, and arrange tables.
[`tidyr`](https://tidyr.tidyverse.org/){target="_blank"} | R | Reshape data between long and wide forms.
[`ggplot2`](https://ggplot2.tidyverse.org/){target="_blank"} | R | Build visualizations.
[`knitr`](https://yihui.org/knitr/){target="_blank"} | R | Render tables in the case-study page.
[`jsonlite`](https://jeroen.r-universe.dev/jsonlite){target="_blank"} | R | Read JSON manifests and diagnostics.
[`reticulate`](https://rstudio.github.io/reticulate/){target="_blank"} | R | Connect R to the Python environment inside Docker when needed.
[`pandas`](https://pandas.pydata.org/){target="_blank"} | Python | Read, filter, and summarize CSV tables.
[`matplotlib`](https://matplotlib.org/){target="_blank"} | Python | Build Python visualizations.
[`CoGAPS`](https://bioconductor.org/packages/CoGAPS){target="_blank"} | R | Rerun CoGAPS models during advanced reproduction.
[`PyCoGAPS`](https://github.com/FertigLab/pycogaps){target="_blank"} | Python | Rerun CoGAPS models during advanced reproduction.
[`scanpy`](https://scanpy.readthedocs.io/){target="_blank"} | Python | Preprocess and inspect single-cell objects.
[`anndata`](https://anndata.readthedocs.io/){target="_blank"} | Python | Read and write H5AD single-cell objects.
Docker | command line | Run the same software environment across computers.
:::

## Setup Troubleshooting

If setup does not work the first time, use the checks below to locate the issue.

::: {.table-responsive}
Problem | What to check | What to try
--- | --- | ---
Docker command is not found | Docker is not installed or Docker Desktop is not running. | Install or open Docker Desktop, then rerun the command from your Terminal.
`http://localhost:8787` does not open | The container may not be running, or another process may already be using port `8787`. | Check the Terminal where Docker is running. If the port is busy, change the left side of the port mapping, such as `-p 8790:8787`, and open `http://localhost:8790`.
RStudio Server opens but project files are missing | The repository may not have been mounted from the repository root. | Stop Docker, `cd` into the cloned `ocs-bio-temporal` folder, and relaunch Docker.
The readiness check cannot find `data/processed` | The repository may be incomplete or you may be in the wrong folder. | Confirm that `/home/rstudio/project` contains `index.qmd`, `data/`, and `scripts/`.
The external archive is not visible | This matters only for advanced reproduction. The archive folder may not be mounted. | Set `CS5_REPRODUCTION_ARCHIVE` to the extracted archive folder and relaunch Docker with the two-mount command.
A package is missing outside Docker | The case study is designed to use the Docker image. | Run the code inside Docker rather than trying to install all packages locally.
:::

***
