# **Data Analysis**
***

The figures in **Data Visualization** nominated `Pattern3` as the strongest temporal candidate in the selected `K = 10` model. They also showed why the interpretation is not automatic. A {{< glossary "Cell score" display="CoGAPS score" >}} can rise across infection days, but the same score can also be shaped by broad PBMC identity, subject structure, and the nonnegative way CoGAPS represents patterns.

This section is where the case study stops looking for interesting shapes and starts testing what those shapes can support. You will carry forward the `K = 10`, `seed = 2`, `nIterations = 2000` model selected in **Data Import**. Then you will ask four questions:

1. Is the selected CoGAPS run technically appropriate to interpret?
1. Which {{< glossary "Latent transcriptional program" display="latent programs" >}} vary across the experimental infection time course?
1. Which programs look {{< glossary "Identity-like pattern" display="identity-like" >}}, {{< glossary "Activity-like pattern" display="activity-like" >}}, or {{< glossary "Mixed pattern" display="mixed" >}}?
1. What evidence supports interpreting one program as an {{< glossary "Interferon-associated response" display="interferon-associated late acute response" >}}?

The analysis still teaches CoGAPS, not just the biology. You will inspect how a fitted CoGAPS model produces two linked matrices: {{< glossary "Gene loading" display="gene loadings" >}} and {{< glossary "Cell score" display="cell scores" >}}. The biological interpretation comes from connecting those matrices back to cells, genes, subjects, timepoints, broad PBMC identities, and directionality evidence.

:::{.reproducibility_box}
::::{.reproducibility_box_header}
How the selected CoGAPS model was run
::::
::::{.inner_block}
You do not need to rerun CoGAPS to complete the main lesson. The included files already contain the {{< glossary "Selected-model outputs" display="selected model outputs" >}}. The excerpts below show the core model call so you can see exactly how the prepared CoGAPS input became the selected `K = 10` model. Full Docker commands, sweep commands, and script bodies are in **Additional Information -> Advanced Reproduction: CoGAPS Model and Analysis Commands**.

**R runner excerpt**

```{r}
#| label: analysis-selected-cogaps-call-r
#| eval: false
sce <- zellkonverter::readH5AD(discovery_h5ad, reader = "R")
data_matrix <- SummarizedExperiment::assay(sce, "X")

params <- CogapsParams(
  nPatterns = 10,
  nIterations = 2000,
  seed = 2,
  sparseOptimization = TRUE
)
params@takePumpSamples <- TRUE

result <- CoGAPS(
  data_matrix,
  params,
  nThreads = 4,
  messages = TRUE,
  outputFrequency = 100,
  checkpointOutFile = "cogaps_K10_seed2_iter2000.checkpoint.out",
  checkpointInterval = 500,
  nSnapshots = 10,
  snapshotPhase = "all",
  asynchronousUpdates = TRUE
)
```

**Python runner excerpt**

```{python}
#| label: analysis-selected-cogaps-call-python
#| eval: false
cogaps_input = ad.read_h5ad(cogaps_input_h5ad)

params = CoParams(adata=cogaps_input)
setParams(
    params,
    {
        "nPatterns": 10,
        "nIterations": 2000,
        "seed": 2,
        "useSparseOptimization": True,
        "transposeData": False,
        "nThreads": 4,
        "takePumpSamples": True,
    },
)

snapshot_frequency = 200  # 10 requested snapshots across 2,000 iterations.

result = CoGAPS(
    cogaps_input,
    params,
    nThreads=4,
    outputFrequency=100,
    checkpointOutFile="cogaps_K10_seed2_iter2000.checkpoint.out",
    checkpointInterval=500,
    asynchronousUpdates=True,
    nSnapshots=snapshot_frequency,
    snapshotPhase="all",
)
```

The R script reads the cells-by-genes discovery object and passes its expression matrix to CoGAPS. The Python script reads the genes-by-cells CoGAPS input and sets `transposeData = False`, which keeps the {{< glossary "Matrix orientation" display="matrix orientation" >}} explicit. Both runners write the same kinds of teaching artifacts: gene loadings, cell scores, top genes, pattern summaries, trace files, diagnostics, and metrics.
::::
:::

## Question 1 Carry-Forward: Is the Selected Model Appropriate to Interpret?

Data Import already answered the rank-selection question: the case study uses `K = 10` as the main model because it stays on the stability plateau and resolves an activity-like interferon-candidate temporal program. `K = 5` remains part of the story, but as a compact {{< glossary "Comparator model" display="sensitivity comparator" >}} rather than the main model.

That decision is necessary, but it is not sufficient. Before interpreting any pattern, check that the selected run has the expected settings and enough {{< glossary "Model-fit diagnostic" display="diagnostic evidence" >}} to be treated as a usable model fit.

::: {.panel-tabset}

### R

```{r}
#| label: analysis-selected-run-settings-r
r_metrics <- jsonlite::fromJSON(
  here::here("data", "processed", "selected_model_k10", "r",
             "cogaps_K10_seed2_iter2000.metrics.json")
)

data.frame(
  setting = c(
    "implementation",
    "K",
    "seed",
    "iterations",
    "CoGAPS threads",
    "sparse optimization",
    "trace output frequency",
    "checkpoint interval",
    "diagnostic snapshots",
    "runtime minutes"
  ),
  value = c(
    r_metrics$language,
    r_metrics$K,
    r_metrics$seed,
    r_metrics$n_iter,
    r_metrics$nThreads,
    r_metrics$sparseOptimization,
    r_metrics$outputFrequency,
    r_metrics$checkpointInterval,
    r_metrics$nSnapshots,
    round(r_metrics$runtime_sec / 60, 1)
  )
) |>
  knitr::kable()
```

### Python

```{python}
#| label: analysis-selected-run-settings-python
with open(
    "data/processed/selected_model_k10/python/"
    "cogaps_K10_seed2_iter2000.metrics.json"
) as f:
    py_metrics = json.load(f)

pd.DataFrame({
    "setting": [
        "implementation",
        "K",
        "seed",
        "iterations",
        "CoGAPS threads",
        "sparse optimization",
        "trace output frequency",
        "checkpoint interval",
        "diagnostic snapshots",
        "runtime minutes",
    ],
    "value": [
        py_metrics["language"],
        py_metrics["K"],
        2,
        py_metrics["n_iter"],
        py_metrics["cogaps_threads"],
        py_metrics["use_sparse_optimization"],
        py_metrics["outputFrequency"],
        py_metrics["checkpointInterval"],
        py_metrics["nSnapshots"],
        round(py_metrics["runtime_sec"] / 60, 1),
    ],
})
```

:::

These settings connect the selected run back to the rank decision from Data Import. The model has ten patterns, uses the representative seed and iteration count, records diagnostic traces every 100 iterations, and saves checkpoint and snapshot information. The run time is also instructive: even this selected model took about 24 minutes in the four-thread Docker setup used to create the included files. That is why the learner path reads the saved outputs instead of rerunning CoGAPS during the lesson.

The trace gives a second check. {{< glossary "Chi-square diagnostic" display="Chi-square" >}} is not a biological result; it is a model-fit diagnostic. A large early drop followed by small late movement is consistent with a run that moved away from a poor initial state and then changed more slowly.

::: {.panel-tabset}

### R

```{r}
#| label: analysis-diagnostic-trace-r
r_trace <- readr::read_csv(
  here::here("data", "processed", "selected_model_k10", "r",
             "cogaps_K10_seed2_iter2000.trace.csv"),
  show_col_types = FALSE
)

ggplot(r_trace, aes(x = trace_index, y = chisq)) +
  geom_line(linewidth = 0.8, color = "#2E74B5") +
  geom_point(size = 1.4, color = "#1F4D78") +
  labs(
    x = "Trace point",
    y = "Chi-square",
    title = "Selected K = 10 CoGAPS diagnostic trace"
  ) +
  theme_minimal(base_size = 11)
```

### Python

```{python}
#| label: analysis-diagnostic-trace-python
py_trace = pd.read_csv(
    "data/processed/selected_model_k10/python/"
    "cogaps_K10_seed2_iter2000.trace.csv"
)

fig, ax = plt.subplots(figsize=(7, 4))
ax.plot(py_trace["trace_index"], py_trace["chisq"], marker="o", color="#2E74B5")
ax.set_xlabel("Trace point")
ax.set_ylabel("Chi-square")
ax.set_title("Selected K = 10 CoGAPS diagnostic trace")
fig.tight_layout()
plt.show()
```

:::

::: {.panel-tabset}

### R

```{r}
#| label: analysis-diagnostic-summary-r
r_diag <- jsonlite::fromJSON(
  here::here("data", "processed", "selected_model_k10", "r",
             "cogaps_K10_seed2_iter2000.diagnostics.json")
)

last_five <- tail(r_trace$chisq, 5)

data.frame(
  diagnostic = c(
    "first trace chi-square",
    "final trace chi-square",
    "chi-square reduction",
    "final-five relative range",
    "trace points",
    "total updates"
  ),
  value = c(
    round(dplyr::first(r_trace$chisq) / 1e6, 1),
    round(dplyr::last(r_trace$chisq) / 1e6, 1),
    paste0(round(
      100 * (dplyr::first(r_trace$chisq) - dplyr::last(r_trace$chisq)) /
        dplyr::first(r_trace$chisq), 1
    ), "%"),
    paste0(round(
      100 * (max(last_five) - min(last_five)) / dplyr::last(r_trace$chisq), 4
    ), "%"),
    r_diag$chisqHistory_length,
    r_diag$totalUpdates
  )
) |>
  knitr::kable()
```

### Python

```{python}
#| label: analysis-diagnostic-summary-python
with open(
    "data/processed/selected_model_k10/python/"
    "cogaps_K10_seed2_iter2000.diagnostics.json"
) as f:
    py_diag = json.load(f)

last_five_py = py_trace["chisq"].tail(5)

pd.DataFrame({
    "diagnostic": [
        "first trace chi-square",
        "final trace chi-square",
        "chi-square reduction",
        "final-five relative range",
        "trace points",
        "total updates",
    ],
    "value": [
        round(py_trace["chisq"].iloc[0] / 1e6, 1),
        round(py_trace["chisq"].iloc[-1] / 1e6, 1),
        f"{100 * (py_trace['chisq'].iloc[0] - py_trace['chisq'].iloc[-1]) / py_trace['chisq'].iloc[0]:.1f}%",
        f"{100 * (last_five_py.max() - last_five_py.min()) / py_trace['chisq'].iloc[-1]:.4f}%",
        py_diag["chisqHistory_length"],
        py_diag["totalUpdates"],
    ],
})
```

:::

The diagnostics support using the selected model for interpretation. Chi-square drops from about 149.3 million to about 124.0 million, and the last five recorded trace points vary by only about 0.005 percent of the final value. That does not prove a unique global optimum, and it does not assign meaning to `Pattern3`. It tells you that the selected run behaved sensibly enough for the biological checks that follow.

## From CoGAPS Output to Biological Evidence

A fitted CoGAPS model does not directly say "this is a temporal immune program." It gives you two matrices that have to be read together.

The **{{< glossary "Gene loading" display="gene-loading matrix" >}}** tells you which genes help define each pattern. A high loading for `ISG15` in `Pattern3`, for example, means `ISG15` contributes strongly to that pattern. The **{{< glossary "Cell score" display="cell-score matrix" >}}** tells you how strongly each cell uses each pattern. A high `Pattern3` score in one cell means that the cell's expression profile contains more of that learned pattern.

Those two matrices become biological evidence only after they are connected back to metadata. Cell scores are joined to subject, infection day, and broad PBMC identity. Gene loadings are ranked into {{< glossary "Top genes" display="top-gene lists" >}}. The interpretation tables used below then combine timing, cell identity, {{< glossary "IFN-score correlation" display="IFN-score correlation" >}}, top genes, and directionality summaries.

The short example below rebuilds one part of that chain from the included selected-model output: it joins the selected cell scores to metadata, then summarizes `Pattern3` by subject and timepoint. The preview focuses on baseline, the late acute window, and recovery so the temporal contrast is easy to read.

::: {.panel-tabset}

### R

```{r}
#| label: analysis-cogaps-output-lineage-r
selected_cells <- readr::read_csv(
  here::here("data", "processed", "selected_model_k10", "r",
             "cogaps_K10_seed2_iter2000.discovery_cells_with_patterns.csv"),
  show_col_types = FALSE
)

subject_time_pattern3 <- selected_cells |>
  group_by(subject, timepoint_merged, day_merged_numeric) |>
  summarize(
    n_cells = n(),
    mean_pattern3_score = mean(Pattern3, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(subject, day_merged_numeric)

subject_time_pattern3 |>
  filter(timepoint_merged %in% c("D0", "D10", "D14/15", "D28")) |>
  select(subject, timepoint_merged, n_cells, mean_pattern3_score) |>
  knitr::kable(digits = 3)
```

### Python

```{python}
#| label: analysis-cogaps-output-lineage-python
selected_cells_py = pd.read_csv(
    "data/processed/selected_model_k10/r/"
    "cogaps_K10_seed2_iter2000.discovery_cells_with_patterns.csv",
    index_col=0,
)

subject_time_pattern3_py = (
    selected_cells_py
    .groupby(["subject", "timepoint_merged", "day_merged_numeric"], as_index=False)
    .agg(
        n_cells=("subject", "size"),
        mean_pattern3_score=("Pattern3", "mean"),
    )
    .sort_values(["subject", "day_merged_numeric"])
)

subject_time_pattern3_py.loc[
    subject_time_pattern3_py["timepoint_merged"].isin(["D0", "D10", "D14/15", "D28"]),
    ["subject", "timepoint_merged", "n_cells", "mean_pattern3_score"],
]
```

:::

Read this table as a change in evidence unit. Each row now represents one subject at one infection day, not one cell. The `n_cells` column shows how many discovery-set cells contributed to that subject-timepoint mean, and `mean_pattern3_score` summarizes how strongly those cells used `Pattern3`.

That grouping matters because the temporal claim should not depend on treating thousands of cells as thousands of independent people. In this preview, each subject has a much higher `Pattern3` mean at `D10` and `D14/15` than at baseline or recovery. The next table asks whether that late acute rise makes `Pattern3` stand out when all selected-model patterns are ranked together.

## Question 2: Which Latent Programs Vary Over Time?

The first biological check asks which patterns move across infection days. Data Visualization made `Pattern3` look important; now the table tests that impression with subject-level summaries and a {{< glossary "Temporal effect size" display="temporal effect size" >}}.

`eta_timepoint` is the table's timepoint effect-size column: it measures how much of a pattern's score variation is associated with infection day. It is not a p-value, and it is not the only evidence you need. It is a useful way to rank temporal candidates before assigning a biological label.

::: {.panel-tabset}

### R

```{r}
#| label: analysis-temporal-programs-r
pattern_annotation <- readr::read_csv(
  here::here("data", "processed", "interpretation",
             "pattern_annotation_table.csv"),
  show_col_types = FALSE
)

temporal_view <- pattern_annotation |>
  select(pattern, observed_label, pattern_class, eta_timepoint,
         subject_peak_timepoint, subject_mean_D0, subject_mean_D10,
         subject_mean_D14_15, subject_mean_D28) |>
  arrange(desc(eta_timepoint)) |>
  head(6)

temporal_view |>
  knitr::kable(digits = 3)
```

### Python

```{python}
#| label: analysis-temporal-programs-python
pattern_annotation_py = pd.read_csv(
    "data/processed/interpretation/pattern_annotation_table.csv"
)

temporal_view_py = (
    pattern_annotation_py
    .loc[:, [
        "pattern", "observed_label", "pattern_class", "eta_timepoint",
        "subject_peak_timepoint", "subject_mean_D0", "subject_mean_D10",
        "subject_mean_D14_15", "subject_mean_D28",
    ]]
    .sort_values("eta_timepoint", ascending=False)
    .head(6)
)

temporal_view_py
```

:::

The table makes the visual story sharper. `Pattern3` has the largest temporal effect by a wide margin (`eta_timepoint` about 0.454). Its subject-level mean is low at `D0`, rises strongly at `D10`, remains high at `D14/15`, and returns near baseline by `D28`. `Pattern9`, which looked like a weaker late-rising candidate in the figures, has a much smaller temporal effect and is still classified as identity-like in the interpretation table.

At this point, the answer to Question 2 is cautious but clear: the selected `K = 10` model contains one dominant temporal program, `Pattern3`, plus smaller time-varying patterns that are more strongly tied to PBMC identity or composition.

## Question 3: Which Programs Reflect Identity, Activity, or Both?

PBMC data mix many immune-cell types. A pattern can look prominent because it separates T cells from monocytes, because it rises during infection, or because both are true. The next check compares two effect sizes: one for infection day and one for broad PBMC identity.

The table keeps those two comparisons separate. `eta_timepoint` is the infection-day effect, while `eta_broad_cell_type` is the broad-cell-type effect; both summarize score variation and should be read as guides rather than definitive labels.

The classification rule is deliberately simple. A pattern is called {{< glossary "Activity-like pattern" display="activity-like" >}} when the timepoint effect is more than 1.5 times the broad-cell-type effect. It is called {{< glossary "Identity-like pattern" display="identity-like" >}} when the broad-cell-type effect is more than 1.5 times the timepoint effect. Anything in between is {{< glossary "Mixed pattern" display="mixed" >}}. This is a guide for interpretation, not a biological truth label.

::: {.panel-tabset}

### R

```{r}
#| label: analysis-identity-activity-r
identity_activity_view <- pattern_annotation |>
  mutate(
    recomputed_class = case_when(
      eta_timepoint > 1.5 * eta_broad_cell_type ~ "activity-like",
      eta_broad_cell_type > 1.5 * eta_timepoint ~ "identity-like",
      TRUE ~ "mixed"
    ),
    time_to_cell_ratio = eta_timepoint / eta_broad_cell_type
  ) |>
  select(pattern, observed_label, pattern_class, recomputed_class,
         eta_timepoint, eta_broad_cell_type, time_to_cell_ratio,
         dominant_broad_cell_type) |>
  arrange(pattern)

identity_activity_view |>
  knitr::kable(digits = 3)
```

### Python

```{python}
#| label: analysis-identity-activity-python
import numpy as np

identity_activity_view_py = pattern_annotation_py.copy()
identity_activity_view_py["recomputed_class"] = np.select(
    [
        identity_activity_view_py["eta_timepoint"]
        > 1.5 * identity_activity_view_py["eta_broad_cell_type"],
        identity_activity_view_py["eta_broad_cell_type"]
        > 1.5 * identity_activity_view_py["eta_timepoint"],
    ],
    ["activity-like", "identity-like"],
    default="mixed",
)
identity_activity_view_py["time_to_cell_ratio"] = (
    identity_activity_view_py["eta_timepoint"]
    / identity_activity_view_py["eta_broad_cell_type"]
)

identity_activity_view_py.loc[:, [
    "pattern", "observed_label", "pattern_class", "recomputed_class",
    "eta_timepoint", "eta_broad_cell_type", "time_to_cell_ratio",
    "dominant_broad_cell_type",
]].sort_values("pattern")
```

:::

Most selected-model patterns are identity-like. That is a useful result, not a distraction: CoGAPS is recovering stable PBMC structure as well as temporal signal. `Pattern7` and `Pattern10`, for example, made sense to display on the UMAP because they act as {{< glossary "Identity anchor" display="identity anchors" >}} for T-cell and B-cell or plasmablast structure.

`Pattern3` is the exception that carries the main temporal claim. Its timepoint effect is about 2.76 times its broad-cell-type effect, so it is the clearest activity-like program in the selected model. The dominant broad-cell-type column still matters: `Pattern3` is not floating outside cell identity. It appears within a PBMC context, especially monocyte-rich regions, while changing strongly across infection days.

## Question 4: What Supports an IFN-Associated Late Acute Interpretation?

The interferon-associated interpretation needs more than a pattern number and a trajectory. A strong candidate should combine timing, gene identity, IFN-score correlation, and directionality. Those evidence types answer different parts of the claim.

First, inspect the genes that define the pattern. A learner does not need to know every immune gene symbol by memory, so the next chunk marks whether the top `Pattern3` genes overlap a small interferon-related reference list.

::: {.panel-tabset}

### R

```{r}
#| label: analysis-ifn-top-genes-r
r_top_genes <- readr::read_csv(
  here::here("data", "processed", "selected_model_k10", "r",
             "cogaps_K10_seed2_iter2000.top_genes.csv"),
  show_col_types = FALSE
)

ifn_reference_genes <- c(
  "LY6E", "ISG15", "IFITM1", "IFI6", "MX1", "IRF7",
  "IFI44L", "IFIT3", "OAS1", "OASL", "IFIT1",
  "IFIT2", "IFITM3", "ISG20", "TRIM22"
)

r_top_genes |>
  filter(pattern == "Pattern3") |>
  arrange(rank) |>
  slice_head(n = 15) |>
  mutate(ifn_reference_gene = gene %in% ifn_reference_genes) |>
  select(rank, gene, weight, ifn_reference_gene) |>
  knitr::kable(digits = 3)
```

### Python

```{python}
#| label: analysis-ifn-top-genes-python
py_top_genes = pd.read_csv(
    "data/processed/selected_model_k10/python/"
    "cogaps_K10_seed2_iter2000.top_genes.csv"
)

ifn_reference_genes = [
    "LY6E", "ISG15", "IFITM1", "IFI6", "MX1", "IRF7",
    "IFI44L", "IFIT3", "OAS1", "OASL", "IFIT1",
    "IFIT2", "IFITM3", "ISG20", "TRIM22",
]

(
    py_top_genes
    .loc[py_top_genes["pattern"].eq("Pattern3"),
         ["pattern", "rank", "gene", "weight"]]
    .sort_values("rank")
    .head(15)
    .assign(ifn_reference_gene=lambda df: df["gene"].isin(ifn_reference_genes))
    .loc[:, ["rank", "gene", "weight", "ifn_reference_gene"]]
)
```

:::

Several of the top `Pattern3` genes are useful interferon anchors, including `LY6E`, `ISG15`, `IFITM1`, `IFI6`, `MX1`, `IRF7`, and `IFI44L`. That supports an interferon-associated reading, but gene identity alone is not enough. The next view brings together the pattern-level evidence: temporal effect, broad-cell-type effect, IFN-score correlation, interferon top-gene overlap, peak timepoint, and directionality at the late acute days.

In this table, `spearman_ifn_score_rho` is the {{< glossary "Spearman correlation" display="Spearman correlation" >}} between the CoGAPS pattern score and an interferon-related program score. Positive values mean cells with higher pattern scores also tend to have higher interferon-related scores.

::: {.panel-tabset}

### R

```{r}
#| label: analysis-ifn-program-evidence-r
directionality_for_pattern3 <- readr::read_csv(
  here::here("data", "processed", "directionality", "tables",
             "pattern_directionality_summary.csv"),
  show_col_types = FALSE
) |>
  filter(pattern == "Pattern3",
         scope == "all_cells",
         stratum == "all_cells",
         timepoint_merged %in% c("D10", "D14/15")) |>
  mutate(timepoint_key = if_else(timepoint_merged == "D14/15",
                                 "D14_15", timepoint_merged)) |>
  select(pattern, timepoint_key,
         median_gene_log2fc, frac_genes_up,
         pattern_direction_consensus) |>
  tidyr::pivot_wider(
    names_from = timepoint_key,
    values_from = c(median_gene_log2fc, frac_genes_up,
                    pattern_direction_consensus)
  )

pattern_annotation |>
  filter(candidate_ifn_pattern) |>
  select(pattern, observed_label, eta_timepoint, eta_broad_cell_type,
         spearman_ifn_score_rho, ifn_top_gene_overlap_top15,
         subject_peak_timepoint) |>
  left_join(directionality_for_pattern3, by = "pattern") |>
  knitr::kable(digits = 3)
```

### Python

```{python}
#| label: analysis-ifn-program-evidence-python
directionality_for_pattern3_py = pd.read_csv(
    "data/processed/directionality/tables/"
    "pattern_directionality_summary.csv"
)

directionality_for_pattern3_py = (
    directionality_for_pattern3_py
    .loc[
        directionality_for_pattern3_py["pattern"].eq("Pattern3")
        & directionality_for_pattern3_py["scope"].eq("all_cells")
        & directionality_for_pattern3_py["stratum"].eq("all_cells")
        & directionality_for_pattern3_py["timepoint_merged"].isin(["D10", "D14/15"]),
        [
            "pattern", "timepoint_merged", "median_gene_log2fc",
            "frac_genes_up", "pattern_direction_consensus",
        ],
    ]
    .assign(
        timepoint_key=lambda x: x["timepoint_merged"].replace({"D14/15": "D14_15"})
    )
    .pivot(
        index="pattern",
        columns="timepoint_key",
        values=[
            "median_gene_log2fc", "frac_genes_up",
            "pattern_direction_consensus",
        ],
    )
)
directionality_for_pattern3_py.columns = [
    f"{value}_{timepoint}"
    for value, timepoint in directionality_for_pattern3_py.columns
]
directionality_for_pattern3_py = directionality_for_pattern3_py.reset_index()

candidate_ifn_mask_py = (
    pattern_annotation_py["candidate_ifn_pattern"]
    .astype(str)
    .str.lower()
    .isin(["true", "1", "yes"])
)

(
    pattern_annotation_py
    .loc[candidate_ifn_mask_py,
         [
             "pattern", "observed_label", "eta_timepoint",
             "eta_broad_cell_type", "spearman_ifn_score_rho",
             "ifn_top_gene_overlap_top15", "subject_peak_timepoint",
         ]]
    .merge(directionality_for_pattern3_py, on="pattern", how="left")
)
```

:::

This row is the strongest support for the main biological interpretation. `Pattern3` is activity-like, has a strong positive IFN-score correlation, includes multiple {{< glossary "Interferon-stimulated genes" display="interferon-stimulated genes" >}} among its top loadings, peaks at `D10`, and remains elevated at `D14/15`. Directionality adds the missing expression-change piece: the top `Pattern3` genes are mostly higher than baseline at the late acute timepoints.

Because CoGAPS is nonnegative, directionality has to be rebuilt outside the model. A high gene loading says that a gene helps define a pattern; it does not say whether that gene increased or decreased after infection. The box below shows how the directionality tables used here were produced.

:::{.reproducibility_box}
::::{.reproducibility_box_header}
How the directionality evidence was rebuilt
::::
::::{.inner_block}

The directionality analysis starts from three files: the preprocessed expression object, the selected model's top-gene table, and the selected model's pattern summary. It tests the top `25` genes for each selected CoGAPS pattern.

The script uses raw counts from the AnnData `counts` layer. For each subject and timepoint, it creates {{< glossary "Pseudobulk directionality" display="pseudobulk expression summaries" >}} by adding counts across cells, converting selected genes to counts per million, and then using `log2(CPM + 1)`. Each post-baseline experimental timepoint is compared with the same subject's `D0` baseline. A gene is labeled higher or lower only when the absolute {{< glossary "Log2 fold-change" display="log2 fold-change" >}} is at least `0.25`, and both the baseline group and the comparison group must contain at least `10` cells.

The same script also builds cell-type-stratified outputs. The main table below uses the `all_cells` summary, while the stratified outputs are available for deeper checks when you want to ask whether directionality is shared across PBMC identities.

```{bash}
#| label: analysis-directionality-rebuild-command
#| eval: false
python scripts/reproduction/gse154386_pattern_directionality.py \
  --sweep-outdir /home/rstudio/project/data/external/reproduction_archive \
  --run-stem cogaps_K10_seed2_iter2000 \
  --outdir /home/rstudio/project/GSE154386/pattern_directionality_revised_model_K10_seed2_iter2000 \
  --top-n 25 \
  --baseline-timepoint D0 \
  --cohort experimental \
  --layer counts \
  --min-cells 10 \
  --log2fc-threshold 0.25
```

The main outputs are `gene_directionality_by_subject.csv`, `gene_directionality_summary.csv`, and `pattern_directionality_summary.csv`. The full Docker command and expanded script body are in **Additional Information -> Advanced Reproduction: CoGAPS Model and Analysis Commands**.

::::
:::

The next table reads the gene-level summary for the leading pattern at `D10`. `median_log2fc` summarizes the subject-level log2 fold-change contrasts for each top gene, and `frac_up` reports the fraction of subject contrasts where that gene is higher than its matched `D0` baseline. Read these columns together: a positive median with a high `frac_up` means the direction is not coming from one subject alone.

::: {.panel-tabset}

### R

```{r}
#| label: analysis-directionality-gene-evidence-r
gene_directionality <- readr::read_csv(
  here::here("data", "processed", "directionality", "tables",
             "gene_directionality_summary.csv"),
  show_col_types = FALSE
)

gene_directionality |>
  filter(pattern == "Pattern3",
         scope == "all_cells",
         stratum == "all_cells",
         timepoint_merged == "D10") |>
  arrange(rank) |>
  select(rank, gene, median_log2fc, frac_up,
         gene_direction_consensus) |>
  head(10) |>
  knitr::kable(digits = 3)
```

### Python

```{python}
#| label: analysis-directionality-gene-evidence-python
gene_directionality_py = pd.read_csv(
    "data/processed/directionality/tables/"
    "gene_directionality_summary.csv"
)

(
    gene_directionality_py
    .loc[
        gene_directionality_py["pattern"].eq("Pattern3")
        & gene_directionality_py["scope"].eq("all_cells")
        & gene_directionality_py["stratum"].eq("all_cells")
        & gene_directionality_py["timepoint_merged"].eq("D10"),
        ["rank", "gene", "median_log2fc", "frac_up",
         "gene_direction_consensus"],
    ]
    .sort_values("rank")
    .head(10)
)
```

:::

The gene-level view shows that the directionality result is not being assigned from one marker. Multiple top `Pattern3` genes show positive median log2 fold-change at `D10`. The pattern-level summary above then condenses those gene-level rows: at `D10`, the median top-gene log2 fold-change is about 1.46 and 88 percent of the tested top genes are up; at `D14/15`, the median is about 1.56 and 96 percent are up.

The answer to Question 4 is therefore evidence-bounded rather than absolute. `Pattern3` is consistent with an IFN-associated late acute response in the experimental discovery subset. The evidence supports that interpretation through timing, activity-like behavior, IFN-linked top genes, IFN-score correlation, and directionality. It does not prove mechanism, identify the cellular source of interferon signaling, or establish clinical causality.

## Sensitivity Check: What Does `K = 5` Compress?

`K = 5` stays in the case study because it teaches an important modeling lesson: the most compact stable rank is not always the best rank for the biological question. If a lower-rank model compresses temporal immune activity into broad identity programs, it may be technically reproducible while still being an {{< glossary "Under-resolved model" display="under-resolved model" >}} for this analysis goal.

Instead of treating `K = 5` as a second analysis path, use it as a {{< glossary "Sensitivity check" display="sensitivity check" >}} for the main claim. Does the compact model recover an interferon-candidate activity-like program comparable to `Pattern3`?

::: {.panel-tabset}

### R

```{r}
#| label: analysis-k5-sensitivity-r
k5_temporal <- readr::read_csv(
  here::here("data", "processed", "comparator_model_k5", "r",
             "RQ1_time_varying_patterns.csv"),
  show_col_types = FALSE
)

k5_top <- k5_temporal |>
  arrange(desc(eta_timepoint)) |>
  slice(1)

cat(
  "The strongest K = 5 temporal effect is ",
  k5_top$pattern,
  " with eta_timepoint = ",
  round(k5_top$eta_timepoint, 3),
  ". It is classified as ",
  k5_top$pattern_class,
  " and candidate_ifn_pattern = ",
  k5_top$candidate_ifn_pattern,
  ". Across all K = 5 patterns, the number of IFN candidates is ",
  sum(k5_temporal$candidate_ifn_pattern),
  ".",
  sep = ""
)
```

### Python

```{python}
#| label: analysis-k5-sensitivity-python
k5_temporal_py = pd.read_csv(
    "data/processed/comparator_model_k5/python/"
    "RQ1_time_varying_patterns.csv"
)

k5_top_py = (
    k5_temporal_py
    .sort_values("eta_timepoint", ascending=False)
    .iloc[0]
)

print(
    "The strongest K = 5 temporal effect is "
    f"{k5_top_py['pattern']} with eta_timepoint = "
    f"{k5_top_py['eta_timepoint']:.3f}. It is classified as "
    f"{k5_top_py['pattern_class']} and candidate_ifn_pattern = "
    f"{k5_top_py['candidate_ifn_pattern']}. Across all K = 5 patterns, "
    f"the number of IFN candidates is "
    f"{int(k5_temporal_py['candidate_ifn_pattern'].sum())}."
)
```

:::

The compact model does not overturn the `K = 10` interpretation. Its strongest temporal effect is small, its strongest patterns are identity-like, and it does not contain an interferon-candidate pattern under the same rule used for the selected model. That comparison is the practical reason `K = 5` remains a comparator: it shows what a stable but compressed model loses.

## What the Evidence Now Supports

The four research questions now line up as one argument.

For Question 1, the case study carries forward the `K = 10` model because Data Import showed that it stays on the stability plateau while resolving the activity-like temporal signal that `K = 5` compresses. The diagnostics in this section add a second layer of confidence: the selected run has the expected settings and a trace that behaves sensibly enough to interpret.

For Question 2, `Pattern3` is the dominant temporal program in the selected model. It rises sharply in the late acute window, peaks at `D10`, remains elevated at `D14/15`, and returns near baseline by `D28`. Other patterns move more modestly across time and are more closely tied to PBMC identity.

For Question 3, most patterns are identity-like. That is exactly what you should expect in PBMC data, where broad immune-cell identity explains a large amount of expression variation. `Pattern3` is the major exception: it is activity-like by the timepoint-versus-cell-type comparison, even though it still appears within recognizable PBMC cell contexts.

For Question 4, the interferon-associated late acute interpretation is supported by {{< glossary "Converging evidence" display="converging evidence" >}} rather than by any single column. `Pattern3` has interferon-linked top genes, a strong IFN-score correlation, late acute timing, and top-gene directionality that is mostly upward relative to baseline. The careful version of the claim is this: in the experimental discovery subset, `Pattern3` is consistent with an interferon-associated late acute activity program.

That wording is intentionally bounded. The analysis supports a coherent interpretation of the selected CoGAPS model, not a causal mechanism. A stronger biological claim would require additional validation, such as cell-type-stratified follow-up, independent cohorts, or the optional projection work described in the homework section.

The Summary now gathers these answers and returns to the larger case-study lesson: good model interpretation depends on both computational checks and biological evidence.

***
