# **What Are the Data?**
***

The source data come from GEO accession GSE154386, associated with Waickman et al. 2021. The study profiled unfractionated PBMCs using a 10x Genomics 5-prime single-cell RNA-seq workflow.

For this case study, the important design distinction is:

- **Experimental infection cohort:** three dengue-naive adults sampled at aligned study days `D0`, `D2`, `D4`, `D6`, `D8`, `D10`, `D14/15`, and `D28`. This is the primary discovery cohort.
- **Natural infection cohort:** two subjects sampled during natural primary infection plus a convalescent reference. This cohort is reserved for optional projection material.

The full raw and preprocessed files are too large for a lightweight GitHub-rendered case study. The repository therefore includes only small teaching artifacts derived from the analysis.

::: {.panel-tabset}

## R

```{r}
#| label: artifact-manifest-summary-r
manifest <- jsonlite::fromJSON(here::here("data", "artifact_manifest.json"))

data.frame(
  artifact_group = c("GitHub-tracked teaching files", "External or local-only files"),
  file_count = c(nrow(manifest$copied), nrow(manifest$skipped)),
  size_mb = c(
    round(sum(manifest$copied$size_bytes) / 1024^2, 2),
    round(sum(manifest$skipped$size_bytes) / 1024^2, 2)
  )
) |>
  knitr::kable()
```

## Python

```{python}
#| label: artifact-manifest-summary-python
import json
import pandas as pd

with open("data/artifact_manifest.json") as f:
    manifest_py = json.load(f)

pd.DataFrame({
    "artifact_group": [
        "GitHub-tracked teaching files",
        "External or local-only files",
    ],
    "file_count": [
        len(manifest_py["copied"]),
        len(manifest_py["skipped"]),
    ],
    "size_mb": [
        round(sum(x["size_bytes"] for x in manifest_py["copied"]) / 1024**2, 2),
        round(sum(x["size_bytes"] for x in manifest_py["skipped"]) / 1024**2, 2),
    ],
})
```

:::

The tracked teaching files include R and Python selected-model outputs, K-selection tables, directionality summaries, R/Python agreement summaries, and case-study figures. The skipped files include large H5ADs, RDS model objects, and checkpoint files. Those are documented in `data/artifact_manifest.json`.

## **Discovery Set**

The core CoGAPS analysis uses a balanced experimental discovery set rather than the entire published single-cell atlas. The discovery set has 4,800 cells and 5,000 highly variable genes. Cells are balanced across experimental samples so that the model is not dominated by a single high-yield sample.

::: {.panel-tabset}

## R

```{r}
#| label: discovery-composition-table-r
composition <- readr::read_csv(
  here::here("data", "processed", "figures", "source_tables",
             "figure_02_discovery_set_composition_source.csv"),
  show_col_types = FALSE
)

composition |>
  group_by(timepoint_merged) |>
  summarize(n_cells = sum(n_cells), .groups = "drop") |>
  knitr::kable()
```

## Python

```{python}
#| label: discovery-composition-table-python
composition_py = pd.read_csv(
    "data/processed/figures/source_tables/"
    "figure_02_discovery_set_composition_source.csv"
)

composition_py.groupby("timepoint_merged", as_index=False)["n_cells"].sum()
```

:::

***
