# **Data Import**
***

In this section, you will import the files that enter the analysis environment. The required case-study path uses small `CSV` and `JSON` files that are already included with the case study. The optional full-reproduction path starts earlier, with large source 10x Genomics files that must be imported into a single-cell object before preprocessing.

A `CSV` file is a rectangular table: rows are observations or summaries, and columns are variables. A `JSON` file stores structured metadata, such as model settings or diagnostic values. In R, the imported files become data frames. In Python, they become tables managed by `pandas`.

If you are completing the required path, you will import the selected-model files needed for the main analysis. If you are reproducing the full workflow locally with the large Zenodo/GEO files, this section also shows where the raw source files enter R or Python. The preprocessing decisions that happen after import are covered in **Data Wrangling and Preprocessing**.

| File family | Example file | What one row means | Why you import it |
|---|---|---|---|
| Pattern summary | `RQ1_time_varying_patterns.csv` | One CoGAPS pattern. | Rank patterns by temporal variation and IFN association. |
| Top-gene table | `cogaps_K10_seed2_iter2000.top_genes.csv` | One gene-pattern entry. | Inspect which genes define each pattern. |
| Run metrics | `cogaps_K10_seed2_iter2000.metrics.json` | One model run summary. | Confirm `K`, seed, iterations, runtime, and diagnostic settings. |
| Diagnostic trace | `cogaps_K10_seed2_iter2000.trace.csv` | One saved trace point during the model run. | Check convergence-related traces before interpretation. |

The larger files needed to rerun preprocessing or CoGAPS from scratch are intentionally external:

- raw GEO download: `GSE154386_RAW.tar`
- preprocessed H5AD cache
- experimental discovery H5ADs
- RDS model objects
- checkpoint files

These files are documented in `data/artifact_manifest.json` and are distributed through Zenodo for full reproduction.

## **Import the Selected-Model Results**

In this subsection, you will turn the selected-model files into named R or Python objects. This is not interpretation yet. It is the setup step that makes later summaries, figures, and diagnostics possible.

The selected-model files come from the `K = 10`, `seed = 2`, `nIterations = 2000` CoGAPS run. You will import four file families: a pattern summary table, a top-gene table, a run-metrics file, and a diagnostic trace. At this stage, the practical question is: did you load the expected files, and do they describe the model you think they describe?

The code below is split into short steps. Each step teaches one programming move: build a path, make a file checklist, read files, inspect table shapes, run sanity checks, and preview the pattern summary.

**Step 1: Point to the selected-model folder.**

The first programming move is to define the folder that contains the files you want to import. In R, `here::here()` builds a path from the project root. In Python, `Path()` from `pathlib` does the same job. Project-relative paths are useful because they work on different computers and inside the Docker runtime.

::: {.panel-tabset}

## R

```{r}
#| label: import-selected-dir-r
r_selected_dir <- here::here(
  "data", "processed", "selected_model_k10", "r"
)

r_selected_dir
```

## Python

```{python}
#| label: import-selected-dir-python
from pathlib import Path
import json
import pandas as pd

py_selected_dir = Path("data/processed/selected_model_k10/python")

py_selected_dir
```

:::

The output is the folder path that the next chunks will use. You do not need to type the full path again each time; you can reuse the named object. In R, that object is called `r_selected_dir`. In Python, it is called `py_selected_dir`.

**Step 2: Make a checklist of expected files.**

Before reading files, make a small checklist. This is a good reproducibility habit: it catches missing files, misspelled file names, and working-directory mistakes before you start interpreting results.

::: {.panel-tabset}

## R

```{r}
#| label: import-file-map-r
r_import_plan <- data.frame(
  file_role = c(
    "pattern summary",
    "top genes",
    "run metrics",
    "diagnostic trace"
  ),
  file_name = c(
    "RQ1_time_varying_patterns.csv",
    "cogaps_K10_seed2_iter2000.top_genes.csv",
    "cogaps_K10_seed2_iter2000.metrics.json",
    "cogaps_K10_seed2_iter2000.trace.csv"
  )
)

r_import_plan$relative_path <- file.path(
  "data", "processed", "selected_model_k10", "r",
  r_import_plan$file_name
)

r_import_plan$file_exists <- file.exists(
  file.path(here::here(), r_import_plan$relative_path)
)

r_import_plan |>
  select(file_role, relative_path, file_exists) |>
  knitr::kable()
```

## Python

```{python}
#| label: import-file-map-python
py_import_plan = pd.DataFrame({
    "file_role": [
        "pattern summary",
        "top genes",
        "run metrics",
        "diagnostic trace",
    ],
    "file_name": [
        "RQ1_time_varying_patterns.csv",
        "cogaps_K10_seed2_iter2000.top_genes.csv",
        "cogaps_K10_seed2_iter2000.metrics.json",
        "cogaps_K10_seed2_iter2000.trace.csv",
    ],
})

py_import_plan["relative_path"] = py_import_plan["file_name"].map(
    lambda name: str(py_selected_dir / name)
)
py_import_plan["file_exists"] = py_import_plan["relative_path"].map(
    lambda path: Path(path).exists()
)

py_import_plan.loc[:, ["file_role", "relative_path", "file_exists"]]
```

:::

The `file_exists` column should be `TRUE` in R or `True` in Python for every required file. If one value is false, the problem is usually a path, file name, or missing-data issue. That is useful to know before you spend time debugging analysis code.

**Step 3: Read the files into memory.**

Now you can import the files. A `CSV` file becomes a rectangular table. A `JSON` file becomes a structured object that stores run-level settings, such as model rank, iterations, and thread count.

The object names are intentionally descriptive. For example, `r_rq1` and `py_rq1` store the pattern-level summary used for the first research question. The top-gene objects store genes ranked by pattern loading, the metrics objects store run settings, and the trace objects store diagnostic values saved during the run.

::: {.panel-tabset}

## R

```{r}
#| label: import-selected-r
r_rq1 <- readr::read_csv(
  file.path(r_selected_dir, "RQ1_time_varying_patterns.csv"),
  show_col_types = FALSE
)

r_top_genes <- readr::read_csv(
  file.path(r_selected_dir, "cogaps_K10_seed2_iter2000.top_genes.csv"),
  show_col_types = FALSE
)

r_model_metrics <- jsonlite::fromJSON(
  file.path(r_selected_dir, "cogaps_K10_seed2_iter2000.metrics.json")
)

r_trace <- readr::read_csv(
  file.path(r_selected_dir, "cogaps_K10_seed2_iter2000.trace.csv"),
  show_col_types = FALSE
)
```

## Python

```{python}
#| label: import-selected-python
py_rq1 = pd.read_csv(py_selected_dir / "RQ1_time_varying_patterns.csv")
py_top_genes = pd.read_csv(
    py_selected_dir / "cogaps_K10_seed2_iter2000.top_genes.csv"
)

with open(py_selected_dir / "cogaps_K10_seed2_iter2000.metrics.json") as f:
    py_model_metrics = json.load(f)

py_trace = pd.read_csv(py_selected_dir / "cogaps_K10_seed2_iter2000.trace.csv")
```

:::

This chunk does not print a table because its main job is to create objects for later use. In R, assignment uses `<-`. In Python, assignment uses `=`. In both languages, the name on the left is the object you will reuse later.

**Step 4: Inspect the imported objects.**

After reading files, inspect their shapes. A table's shape tells you how many rows and columns were imported. This quick check can reveal accidental empty files, wrong file paths, or files that do not have the columns you expected.

::: {.panel-tabset}

## R

```{r}
#| label: import-object-shapes-r
data.frame(
  imported_object = c("r_rq1", "r_top_genes", "r_trace"),
  rows = c(nrow(r_rq1), nrow(r_top_genes), nrow(r_trace)),
  columns = c(ncol(r_rq1), ncol(r_top_genes), ncol(r_trace)),
  first_columns = c(
    paste(head(names(r_rq1), 5), collapse = ", "),
    paste(head(names(r_top_genes), 4), collapse = ", "),
    paste(head(names(r_trace), 4), collapse = ", ")
  )
) |>
  knitr::kable()
```

## Python

```{python}
#| label: import-object-shapes-python
pd.DataFrame({
    "imported_object": ["py_rq1", "py_top_genes", "py_trace"],
    "rows": [len(py_rq1), len(py_top_genes), len(py_trace)],
    "columns": [py_rq1.shape[1], py_top_genes.shape[1], py_trace.shape[1]],
    "first_columns": [
        ", ".join(py_rq1.columns[:5]),
        ", ".join(py_top_genes.columns[:4]),
        ", ".join(py_trace.columns[:4]),
    ],
})
```

:::

Read this output as a basic import audit. The row counts tell you how much information was loaded. The `first_columns` field tells you whether the table looks like the file you intended to read. For example, the pattern summary should begin with pattern-level columns, while the trace file should begin with diagnostic trace columns.

**Step 5: Run sanity checks on the selected model.**

The next check asks whether the imported files describe the selected model used throughout the case study. These checks do not prove that the model is biologically correct. They only confirm that you imported the expected `K = 10`, `seed = 2`, `nIterations = 2000` run and that the main tables contain all ten patterns.

::: {.panel-tabset}

## R

```{r}
#| label: import-sanity-r
data.frame(
  check = c(
    "model rank K",
    "seed",
    "iterations",
    "CoGAPS threads",
    "number of pattern rows",
    "patterns represented in top-gene table",
    "diagnostic trace length"
  ),
  value = c(
    r_model_metrics$K,
    r_model_metrics$seed,
    r_model_metrics$n_iter,
    r_model_metrics$nThreads,
    nrow(r_rq1),
    length(unique(r_top_genes$pattern)),
    nrow(r_trace)
  )
) |>
  knitr::kable()
```

## Python

```{python}
#| label: import-sanity-python
pd.DataFrame({
    "check": [
        "model rank K",
        "seed",
        "iterations",
        "CoGAPS threads",
        "number of pattern rows",
        "patterns represented in top-gene table",
        "diagnostic trace length",
    ],
    "value": [
        py_model_metrics["K"],
        py_model_metrics.get("seed", 2),
        py_model_metrics["n_iter"],
        py_model_metrics["cogaps_threads"],
        len(py_rq1),
        py_top_genes["pattern"].nunique(),
        len(py_trace),
    ],
})
```

:::

The sanity-check output should confirm that the imported files match the selected model. The number of pattern rows should be `10`, and the top-gene table should contain entries for all ten patterns. If those values do not match, stop here and resolve the import problem before continuing.

**Step 6: Preview the pattern summary.**

As a final import check, sort the pattern summary by temporal variation. This preview helps you see why later sections focus on specific patterns, but it is not enough evidence by itself. A temporal score needs to be interpreted alongside cell-type summaries, top genes, model diagnostics, and directionality.

::: {.panel-tabset}

## R

```{r}
#| label: import-preview-r
r_rq1 |>
  select(pattern, eta_timepoint, peak_timepoint,
         spearman_ifn_score_rho, candidate_ifn_pattern) |>
  arrange(desc(eta_timepoint)) |>
  head(5) |>
  knitr::kable(digits = 3)
```

## Python

```{python}
#| label: import-preview-python
py_rq1.loc[
    :, ["pattern", "eta_timepoint", "peak_timepoint",
        "spearman_ifn_score_rho", "candidate_ifn_pattern"]
].sort_values("eta_timepoint", ascending=False).head(5)
```

:::

By the end of this subsection, the selected-model files have become named objects in R or Python. You have also checked that they exist, have the expected shapes, and describe the selected `K = 10` model. The next sections use these imported tables to explore the discovery set and reshape model outputs for interpretation.

## **Optional: Import Source 10x Files for Full Reproduction**

You can skip this subsection if you are completing the required case-study path. It is included so the source of the prepared files is transparent, and so learners who download the large Zenodo/GEO files can see how the raw 10x Genomics files enter the workflow.

The optional import path stops once the source data have become a labeled single-cell object. Quality control, normalization, gene selection, annotation, and discovery-set construction happen later in **Data Wrangling and Preprocessing**.

:::{.callout-note title="How to run the optional full workflow"}
The code chunks in this subsection are excerpts from the complete preparation scripts. They are shown so you can understand how the source 10x files become a labeled single-cell object. They are not intended to be copied and run one chunk at a time.

If you want to run the optional full reproduction workflow, use the complete script shown in **Data Wrangling and Preprocessing**. That one script includes both the import steps in this section and the preprocessing steps in the next section. Keeping those steps in one script makes the workflow easier to rerun because the downloaded files, imported objects, quality-control decisions, and saved outputs are handled in one continuous run.
:::

```{mermaid}
flowchart TB
  classDef source fill:#eaf3ff,stroke:#2d6da3,stroke-width:2px,color:#102a43
  classDef process fill:#f7f9fb,stroke:#6b7c8f,stroke-width:1.5px,color:#102a43
  classDef stage fill:#1f4e79,stroke:#123047,stroke-width:2px,color:#ffffff

  A["Source data<br/><b>GEO GSE154386</b><br/>raw 10x archive"]:::source

  subgraph S1[" "]
    P1["1. Build the single-cell<br/>data object"]:::stage
    B["Parse file names<br/>subject, cohort, day"]:::process
    C["Read 10x triplets<br/>matrix + barcodes + features"]:::process
    D["Attach metadata<br/>cells x genes object"]:::process
  end

  A --> P1 --> B --> C --> D

  style S1 fill:#f8fafc,stroke:#cbd5e1,stroke-width:1px,color:#102a43
```

:::{.definition_box}
::::{.definition_box_header}
Terms to know for optional source-file import
::::
::::{.inner_block}
- **Sample-level 10x files:** Raw 10x Genomics files organized by biological sample. Each sample has linked files for the expression matrix, cell barcodes, and gene or feature names.
- **10x triplet:** A set of three files that must be read together: `matrix.mtx`, `barcodes.tsv`, and `features.tsv` or `genes.tsv`.
- **Cell barcode:** A sequence label used to identify one captured cell or droplet in a single-cell sequencing experiment.
- **Feature:** A measured item in the 10x matrix. In this case study, the import code keeps gene-expression features for the RNA-seq analysis.
- **Cell-by-gene matrix:** A matrix where rows are cells and columns are genes. Each value records the expression measurement for one gene in one cell.
- **Sparse matrix:** A compact matrix format used when most values are zero. Single-cell RNA-seq count matrices are usually sparse because most genes are not detected in most cells.
- **Single-cell object:** A labeled data container, such as `SingleCellExperiment` in R or `AnnData` in Python, that keeps expression values aligned with cell and gene annotations.
- **`SingleCellExperiment`:** An R/Bioconductor object for single-cell data. It stores expression matrices in assays, cell metadata in `colData()`, and gene metadata in `rowData()`.
- **`AnnData`:** A Python object commonly used by Scanpy for single-cell data. It stores the expression matrix in `.X`, cell metadata in `.obs`, and gene metadata in `.var`.
- **Cell metadata:** Information attached to each cell, such as sample ID, subject, cohort, and infection day.
- **Gene metadata:** Information attached to each gene, such as gene symbol, gene ID, or feature type.
::::
:::

Those terms explain why the import code is careful about filenames, matrix orientation, and metadata. A temporal immune-program analysis only works if each expression value is attached to the correct cell, gene, subject, and day.

### **Identify Source Files and Sample Metadata**

Before expression values can be interpreted biologically, each cell has to be connected back to its sample. In this dataset, the sample name tells you which subject the cell came from, whether that subject was in the experimental or natural infection cohort, and which infection day the sample represents.

The raw GEO archive contains sample-level 10x files. In this context, "sample-level" means that the raw sequencing output is organized one biological sample at a time, rather than as one combined study-wide table. For each sample, the 10x Genomics files are usually a small set of linked files: a sparse expression matrix, a list of cell barcodes, and a list of gene or feature names. These files have to be read together because the matrix alone only contains numbers; the barcode and feature files tell you which cells and genes those numbers refer to.

The programming idea is to make implicit information explicit. A **variable** stores a value you want to reuse, such as a source URL or file path. A **function** is reusable code with inputs and outputs. Here, the function `parse_sample_key()` takes one sample name as input and returns metadata columns as output. Later, those metadata columns are copied onto every cell from that sample.

**Step 1: Record where the raw archive comes from.**

The first code step does not download anything. It gives a name to the GEO source URL and to the local file name that would store the source archive. Naming these values makes the optional full workflow easier to audit because the source location is written once and reused.

::: {.panel-tabset}

## R

```{r}
#| label: source-archive-r
RAW_TAR_URL <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE154386&format=file"
RAW_TAR <- file.path("data", "external", "GSE154386", "GSE154386_RAW.tar")

data.frame(
  item = c("GEO source archive", "local archive path"),
  value = c(RAW_TAR_URL, RAW_TAR)
) |>
  knitr::kable()
```

## Python

```{python}
#| label: source-archive-python
from pathlib import Path
import pandas as pd

RAW_TAR_URL = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE154386&format=file"
RAW_TAR = Path("data/external/GSE154386/GSE154386_RAW.tar")

pd.DataFrame({
    "item": ["GEO source archive", "local archive path"],
    "value": [RAW_TAR_URL, str(RAW_TAR)],
})
```

:::

The archive is a source file, not an analysis table. It contains raw 10x files that still need to be connected to sample labels before quality control or modeling can happen.

**Step 2: Inspect the structure of one sample name.**

The source filenames are not arbitrary. A name such as `GSM4670192_Subject2_D0` can be read as three pieces: a GEO sample identifier, a subject label, and a day label. The experimental samples use labels such as `Subject2`, while natural infection samples use labels such as `Natural1`.

::: {.panel-tabset}

## R

```{r}
#| label: sample-key-examples-r
sample_key_examples <- data.frame(
  sample_key = c(
    "GSM4670192_Subject2_D0",
    "GSM4670216_Natural1_Dneg3"
  ),
  meaning = c(
    "experimental subject 2 at day 0",
    "natural infection sample 1 at day -3"
  )
)

sample_key_examples |>
  knitr::kable()
```

## Python

```{python}
#| label: sample-key-examples-python
sample_key_examples = pd.DataFrame({
    "sample_key": [
        "GSM4670192_Subject2_D0",
        "GSM4670216_Natural1_Dneg3",
    ],
    "meaning": [
        "experimental subject 2 at day 0",
        "natural infection sample 1 at day -3",
    ],
})

sample_key_examples
```

:::

The label `D0` means day 0. A label such as `Dneg3` means day -3. The numeric form of the day will matter later because it lets you sort timepoints and summarize pattern scores along the infection time course.

**Step 3: Write a function that parses a sample name.**

The next code defines `parse_sample_key()`. In programming, a function is useful when the same operation needs to be repeated many times. The input is one sample name. The output is a small set of metadata fields that can be attached to cells from that sample.

The main programming tools are a regular expression and an `if` statement. A regular expression is a pattern used to find pieces of text. Here, it captures the GEO identifier, the subject label, and the day label. The `if` statement then translates the subject label into a cohort label and translates the day label into a number.

::: {.panel-tabset}

## R

```{r}
#| label: parse-sample-key-r
parse_sample_key <- function(sample_key) {
  match <- regexec(
    "^(GSM[0-9]+)_([^_]+)_(D(?:neg)?[0-9]+)$",
    sample_key,
    perl = TRUE
  )
  parts <- regmatches(sample_key, match)[[1]]
  if (length(parts) == 0) {
    stop("Could not parse sample key: ", sample_key)
  }

  gsm_id <- parts[[2]]
  subject <- parts[[3]]
  tp <- parts[[4]]

  cohort <- if (startsWith(subject, "Subject")) {
    "experimental"
  } else if (startsWith(subject, "Natural")) {
    "natural"
  } else {
    "unknown"
  }

  day <- if (startsWith(tp, "Dneg")) {
    -as.integer(sub("Dneg", "", tp))
  } else {
    as.integer(sub("D", "", tp))
  }

  data.frame(
    gsm_id = gsm_id,
    sample_id = gsm_id,
    subject = subject,
    cohort = cohort,
    timepoint_raw = tp,
    timepoint_display = paste0("D", day),
    day_numeric = day,
    stringsAsFactors = FALSE
  )
}
```

## Python

```{python}
#| label: parse-sample-key-python
import re

def parse_sample_key(sample_key: str) -> dict:
    match = re.match(r"^(GSM\d+)_([^_]+)_(D(?:neg)?\d+)$", sample_key)
    if not match:
        raise ValueError(f"Could not parse sample key: {sample_key}")

    gsm_id, subject, tp = match.groups()

    if subject.startswith("Subject"):
        cohort = "experimental"
    elif subject.startswith("Natural"):
        cohort = "natural"
    else:
        cohort = "unknown"

    day = -int(tp.replace("Dneg", "")) if tp.startswith("Dneg") else int(tp.replace("D", ""))

    return {
        "gsm_id": gsm_id,
        "sample_id": gsm_id,
        "subject": subject,
        "cohort": cohort,
        "timepoint_raw": tp,
        "timepoint_display": f"D{day}",
        "day_numeric": day,
    }
```

:::

In R, the parsed metadata becomes a data frame. In Python, it becomes a dictionary, which can then be placed into a `pandas` table. In both languages, the goal is the same: every cell must carry the biological labels needed for subject-level and timepoint-level summaries.

**Step 4: Test the function on two sample names.**

When you write a function, test it on a small example before using it on every file in a dataset. This is a programming habit, but it is also a biological check. If the day or cohort is parsed incorrectly here, every later time-course summary will be mislabeled.

::: {.panel-tabset}

## R

```{r}
#| label: parse-sample-key-demo-r
parsed_examples <- do.call(
  rbind,
  lapply(sample_key_examples$sample_key, parse_sample_key)
)

parsed_examples |>
  knitr::kable()
```

## Python

```{python}
#| label: parse-sample-key-demo-python
parsed_examples = pd.DataFrame([
    parse_sample_key(sample_key)
    for sample_key in sample_key_examples["sample_key"]
])

parsed_examples
```

:::

Read the output from left to right. The `gsm_id` and `sample_id` identify the GEO sample. The `subject` and `cohort` columns tell you which biological group the sample belongs to. The `timepoint_display` column is a readable label for plotting, while `day_numeric` is the numeric version used for sorting and summarizing days.

:::{.callout-tip title="Checkpoint"}
If the sample name is `GSM4670198_Subject2_D15`, what should the cohort and numeric day be?

<details>
<summary>Show answer</summary>

The cohort should be `experimental` because the subject label begins with `Subject`. The numeric day should be `15` because the timepoint label is `D15`.

</details>
:::

<details>
<summary>Optional helper: download the GEO archive safely</summary>

The complete preparation scripts include a small download helper. This helper is kept separate from the parsing lesson because downloading is not the main biological idea here. It is still useful for full reproduction because it skips an existing archive and avoids treating the tar file as a text table.

::: {.panel-tabset}

## R

```{r}
#| eval: false
download_if_needed <- function(url, dest) {
  if (file.exists(dest) && file.info(dest)$size > 0) {
    message("[skip] ", dest, " already exists")
    return(invisible(dest))
  }

  download.file(url, destfile = dest, mode = "wb", quiet = FALSE)
  invisible(dest)
}
```

## Python

```{python}
#| eval: false
import requests

def download(url: str, dest: Path, chunk_mb: int = 16) -> None:
    if dest.exists() and dest.stat().st_size > 0:
        print(f"[skip] {dest} already exists")
        return

    tmp = dest.with_suffix(dest.suffix + ".part")
    with requests.get(url, stream=True, timeout=180) as response:
        response.raise_for_status()
        with open(tmp, "wb") as handle:
            for chunk in response.iter_content(chunk_size=chunk_mb * 1024 * 1024):
                if chunk:
                    handle.write(chunk)
    tmp.rename(dest)
```

:::

</details>

After this step, the workflow has not changed any expression values yet. The important result is that each future cell-level observation can carry subject, cohort, and day labels. Without those labels, the later CoGAPS pattern scores could not be summarized by timepoint or subject.

### **Build the Cell-by-Gene Data Object**

In the previous step, you learned how a sample name becomes metadata. Now you will see how the raw expression files become a labeled cell-by-gene object. The goal is not just to read files; the goal is to preserve the relationship between counts, cells, genes, and sample metadata.

Single-cell RNA-seq data do not begin as a tidy table. For each 10x Genomics sample, the data are split into three linked files. None of these files is biologically interpretable on its own.

| 10x file | What it contains | Why it matters |
|---|---|---|
| `matrix.mtx` | The expression count matrix. In raw 10x output, rows are usually features and columns are cells. | Stores the measured RNA counts, but not the cell or gene names. |
| `barcodes.tsv` | Cell barcode labels. | Tells you which captured cell or droplet each matrix column represents. |
| `features.tsv` or `genes.tsv` | Gene IDs, gene symbols, and sometimes feature types. | Tells you which gene or measured feature each matrix row represents. |

This alignment is biologically important. If genes and cells are mislabeled or transposed, marker scoring and CoGAPS interpretation can point to the wrong biological signal. The code below uses a small toy example so you can see the alignment logic without downloading the full GEO archive.

The R and Python workflows use different object classes for the same basic purpose. They are different containers, but they solve the same problem: they keep the count matrix connected to cell and gene annotations.

| Concept | R `SingleCellExperiment` | Python `AnnData` |
|---|---|---|
| Expression matrix | `assays(sce)$counts` | `adata.X` |
| Cell metadata | `colData(sce)` | `adata.obs` |
| Gene metadata | `rowData(sce)` | `adata.var` |
| Cell names | `colnames(sce)` | `adata.obs_names` |
| Gene names | `rownames(sce)` | `adata.var_names` |

**Step 1: Represent the three linked 10x pieces.**

In the full reproduction script, this step reads `matrix.mtx`, `barcodes.tsv`, and `features.tsv` from disk. Here, the same idea is shown with a small in-memory example. The matrix, barcode table, and feature table are separate objects at first.

::: {.panel-tabset}

## R

```{r}
#| label: toy-10x-pieces-r
toy_counts_10x <- matrix(
  c(1, 0,
    3, 2,
    0, 4,
    7, 1),
  nrow = 4,
  byrow = TRUE
)

toy_barcodes <- c("AAAC-1", "TTGG-1")

toy_features <- data.frame(
  gene_id = c("ENSG1", "ENSG2", "ENSG3", "AB1"),
  gene_symbol = c("ISG15", "MX1", "CD3D", "TotalSeq_B"),
  feature_type = c(
    "Gene Expression",
    "Gene Expression",
    "Gene Expression",
    "Antibody Capture"
  )
)

data.frame(
  object = c("toy_counts_10x", "toy_barcodes", "toy_features"),
  rows = c(nrow(toy_counts_10x), length(toy_barcodes), nrow(toy_features)),
  columns = c(ncol(toy_counts_10x), NA, ncol(toy_features))
) |>
  knitr::kable()
```

## Python

```{python}
#| label: toy-10x-pieces-python
import numpy as np

toy_counts_10x = np.array([
    [1, 0],
    [3, 2],
    [0, 4],
    [7, 1],
])

toy_barcodes = ["AAAC-1", "TTGG-1"]

toy_features = pd.DataFrame({
    "gene_id": ["ENSG1", "ENSG2", "ENSG3", "AB1"],
    "gene_symbol": ["ISG15", "MX1", "CD3D", "TotalSeq_B"],
    "feature_type": [
        "Gene Expression",
        "Gene Expression",
        "Gene Expression",
        "Antibody Capture",
    ],
})

pd.DataFrame({
    "object": ["toy_counts_10x", "toy_barcodes", "toy_features"],
    "rows": [toy_counts_10x.shape[0], len(toy_barcodes), len(toy_features)],
    "columns": [toy_counts_10x.shape[1], None, toy_features.shape[1]],
})
```

:::

In this toy example, the count matrix has four feature rows and two cell columns. The feature table also has four rows, and the barcode list has two labels. Those matching dimensions are the first sign that the pieces can be aligned.

**Step 2: Check orientation and dimensions.**

Raw 10x matrices are usually stored as features by cells. That means rows match the feature table and columns match the barcode table. Many downstream analyses are described as cell-by-gene analyses, so the workflow needs to know when a transpose is required. A transpose flips a matrix so rows become columns and columns become rows.

::: {.panel-tabset}

## R

```{r}
#| label: toy-10x-dim-check-r
data.frame(
  check = c("feature rows match matrix rows", "barcode labels match matrix columns"),
  expected = c(nrow(toy_features), length(toy_barcodes)),
  observed = c(nrow(toy_counts_10x), ncol(toy_counts_10x)),
  passes = c(
    nrow(toy_features) == nrow(toy_counts_10x),
    length(toy_barcodes) == ncol(toy_counts_10x)
  )
) |>
  knitr::kable()
```

## Python

```{python}
#| label: toy-10x-dim-check-python
pd.DataFrame({
    "check": [
        "feature rows match matrix rows",
        "barcode labels match matrix columns",
    ],
    "expected": [len(toy_features), len(toy_barcodes)],
    "observed": [toy_counts_10x.shape[0], toy_counts_10x.shape[1]],
    "passes": [
        len(toy_features) == toy_counts_10x.shape[0],
        len(toy_barcodes) == toy_counts_10x.shape[1],
    ],
})
```

:::

These checks look simple, but they protect the biological interpretation. If the feature count does not match the matrix rows, gene labels could be attached to the wrong counts. If the barcode count does not match the matrix columns, cell metadata could be attached to the wrong cells.

**Step 3: Keep the gene-expression rows.**

Some 10x files include feature types that are not RNA gene-expression measurements. This case study models transcript abundance, so the import workflow keeps `Gene Expression` rows and drops other feature types before CoGAPS input preparation.

::: {.panel-tabset}

## R

```{r}
#| label: toy-gene-expression-filter-r
keep_gene_expression <- toy_features$feature_type == "Gene Expression"

toy_counts_gene_expression <- toy_counts_10x[keep_gene_expression, , drop = FALSE]
toy_features_gene_expression <- toy_features[keep_gene_expression, , drop = FALSE]

data.frame(
  retained_gene_symbols = toy_features_gene_expression$gene_symbol,
  retained_feature_type = toy_features_gene_expression$feature_type
) |>
  knitr::kable()
```

## Python

```{python}
#| label: toy-gene-expression-filter-python
keep_gene_expression = toy_features["feature_type"].eq("Gene Expression")

toy_counts_gene_expression = toy_counts_10x[keep_gene_expression.to_numpy(), :]
toy_features_gene_expression = toy_features.loc[keep_gene_expression, :].copy()

toy_features_gene_expression.loc[
    :, ["gene_symbol", "feature_type"]
].rename(columns={
    "gene_symbol": "retained_gene_symbols",
    "feature_type": "retained_feature_type",
})
```

:::

The dropped toy feature, `TotalSeq_B`, represents the kind of non-RNA feature that should not be mixed into this RNA expression model. In the real dataset, this filtering step keeps the biological question focused on gene-expression programs.

**Step 4: Name cells and genes before constructing the object.**

Cell barcodes are only guaranteed to be unique within a sample. The same barcode sequence can appear in different samples, so the workflow prefixes each barcode with the sample ID. Gene symbols are easier for humans to interpret than gene IDs, but duplicated gene symbols can occur, so the workflow makes gene names unique.

::: {.panel-tabset}

## R

```{r}
#| label: toy-name-cells-genes-r
toy_meta <- parse_sample_key("GSM4670192_Subject2_D0")

toy_gene_names <- make.unique(
  toy_features_gene_expression$gene_symbol,
  sep = "_"
)
toy_cell_names <- paste0(toy_meta$sample_id, ":", toy_barcodes)

data.frame(
  cell_name = toy_cell_names,
  sample_id = toy_meta$sample_id,
  subject = toy_meta$subject,
  timepoint = toy_meta$timepoint_display
) |>
  knitr::kable()
```

## Python

```{python}
#| label: toy-name-cells-genes-python
toy_meta = parse_sample_key("GSM4670192_Subject2_D0")

toy_gene_names = pd.Index(
    toy_features_gene_expression["gene_symbol"]
).astype(str).tolist()
toy_cell_names = [
    f"{toy_meta['sample_id']}:{barcode}"
    for barcode in toy_barcodes
]

pd.DataFrame({
    "cell_name": toy_cell_names,
    "sample_id": toy_meta["sample_id"],
    "subject": toy_meta["subject"],
    "timepoint": toy_meta["timepoint_display"],
})
```

:::

This is where the metadata from the previous subsection starts to matter. The count matrix has measurements, but the cell names and metadata make those measurements traceable to a subject and infection day.

**Step 5: Build a labeled cell-by-gene representation.**

The raw 10x matrix is features by cells. For a cell-by-gene view, transpose the retained matrix so cells are rows and genes are columns. The small table below shows what the aligned expression data look like after naming cells and genes.

::: {.panel-tabset}

## R

```{r}
#| label: toy-cell-by-gene-r
toy_cell_by_gene <- t(toy_counts_gene_expression)
rownames(toy_cell_by_gene) <- toy_cell_names
colnames(toy_cell_by_gene) <- toy_gene_names

toy_cell_by_gene_display <- data.frame(
  cell_name = rownames(toy_cell_by_gene),
  as.data.frame(toy_cell_by_gene),
  check.names = FALSE
)

toy_cell_by_gene_display |>
  knitr::kable()
```

## Python

```{python}
#| label: toy-cell-by-gene-python
toy_cell_by_gene = pd.DataFrame(
    toy_counts_gene_expression.T,
    index=toy_cell_names,
    columns=toy_gene_names,
)

toy_cell_by_gene.reset_index().rename(columns={"index": "cell_name"})
```

:::

In the full workflow, this aligned matrix is stored in a single-cell container rather than as a plain table. In R, the container is `SingleCellExperiment`; in Python, it is `AnnData`. The container keeps the expression matrix, cell metadata, and gene metadata together when later steps filter cells, select genes, score markers, or save outputs.

<details>
<summary>Optional: construct the full single-cell container</summary>

The code below shows the final construction step used in the complete preparation scripts. It is shown for transparency, but the key concept is the same one you saw in the toy table: counts, cell labels, gene labels, and metadata must stay aligned.

::: {.panel-tabset}

## R

```{r}
#| eval: false
SingleCellExperiment::SingleCellExperiment(
  assays = list(counts = mat),
  rowData = S4Vectors::DataFrame(gene_id = gene_ids[keep]),
  colData = S4Vectors::DataFrame(meta[rep(1L, ncol(mat)), , drop = FALSE])
)
```

## Python

```{python}
#| eval: false
from anndata import AnnData

adata = AnnData(X=matrix.T.tocsr())
adata.obs_names = [
    f"{meta['sample_id']}:{barcode}"
    for barcode in barcodes
]
adata.var_names = pd.Index([
    gene
    for gene, keep_gene in zip(gene_symbols, keep)
    if keep_gene
])
adata.var_names_make_unique()
```

:::

</details>

:::{.callout-tip title="Checkpoint"}
If a raw 10x matrix has 20,000 feature rows and 800 cell columns, how many barcode labels should there be?

<details>
<summary>Show answer</summary>

There should be 800 barcode labels because barcodes label cells, and cells are columns in the raw 10x matrix before it is transposed into a cell-by-gene view.

</details>
:::

After this step, the optional import path has produced a labeled single-cell object. Biologically, that means every expression measurement can be traced back to a cell, a gene, and a sample. In code, later filtering and summarizing can use the same object instead of passing separate matrices and metadata tables around by hand.

::: {.callout-note collapse="true" title="Optional: file-reading helper functions used above"}
The complete preparation scripts use small helper functions to read compressed 10x matrix and table files. These helpers are file-reading utilities, not biological decisions. They are shown here for transparency, but the important concept is that the matrix, barcode table, and feature table are read together and checked before they are aligned into one labeled single-cell object.

The `parse_sample_key()` function is defined in the previous subsection. The helpers below handle the remaining file-reading steps.

**R**

```{r}
#| eval: false
open_text_file <- function(path) {
  # 10x files are often gzip-compressed. This helper opens either a
  # compressed or uncompressed text file.
  if (grepl("\\.gz$", path)) {
    gzfile(path, open = "rt")
  } else {
    file(path, open = "rt")
  }
}

open_binary_file <- function(path) {
  # Matrix Market files are read as binary connections.
  if (grepl("\\.gz$", path)) {
    gzfile(path, open = "rb")
  } else {
    file(path, open = "rb")
  }
}

read_table_gz <- function(path) {
  # Read a barcode or feature table with no header row.
  con <- open_text_file(path)
  on.exit(close(con), add = TRUE)

  readr::read_tsv(
    con,
    col_names = FALSE,
    show_col_types = FALSE
  )
}

read_mtx_gz <- function(path) {
  # Read a sparse Matrix Market expression matrix.
  con <- open_binary_file(path)
  on.exit(close(con), add = TRUE)

  Matrix::readMM(con)
}
```

**Python**

```{python}
#| eval: false
import gzip
import pandas as pd
from scipy import sparse
from scipy.io import mmread

def read_text_table(path):
    # Read a barcode or feature table with no header row.
    opener = gzip.open if path.suffix == ".gz" else open
    with opener(path, "rt") as handle:
        return pd.read_csv(handle, sep="\t", header=None)

def read_mtx(path):
    # Read a sparse Matrix Market expression matrix.
    opener = gzip.open if path.suffix == ".gz" else open
    with opener(path, "rb") as handle:
        matrix = mmread(handle)
    return sparse.csr_matrix(matrix)
```
:::

The next section uses the small included files to explore the discovery set before you move deeper into preprocessing decisions. The optional full-preparation workflow resumes in **Data Wrangling and Preprocessing**, where the imported single-cell object is filtered, normalized, annotated, and subset for modeling. The complete runnable preparation scripts are provided there.

***
