# **Data Wrangling and Preprocessing**
***

The full preprocessing workflow starts from GEO 10x files, constructs an AnnData object, performs QC, normalizes counts, selects highly variable genes, annotates broad cell types, and builds a balanced experimental discovery set. That workflow is intentionally treated as full reproduction material rather than a required rendered-case-study step.

The required wrangling task is to convert model artifacts into tidy tables for interpretation.

## **Optional Full Preprocessing Workflow: From GEO to Discovery Subset**

The code in this subsection is shown for transparency and reproducibility. It is **not run during the rendered case study** because it downloads and processes large files. Instead, the standard case-study path starts from the small precomputed artifacts imported earlier.

This full preparation workflow starts from the GEO raw 10x files and ends with the large local objects used for CoGAPS:

1. download and extract the GSE154386 raw archive
1. discover per-sample 10x matrix/barcode/feature files
1. parse sample metadata from GEO file names
1. build a combined single-cell object
1. perform QC, normalization, gene filtering, and HVG selection
1. score broad immune-cell marker sets and biology programs
1. annotate broad immune-cell types for teaching summaries
1. build the balanced experimental discovery subset
1. create a natural-cohort projection target for the optional extension

This script intentionally stops before running CoGAPS. The selected CoGAPS model is covered in the analysis section.

Learners do not need to run this code to complete the case study. Instead, use it as an audit trail for the teaching dataset.

```{r}
#| label: preprocessing-inspection-prompts
data.frame(
  thing_to_inspect = c(
    "QC thresholds",
    "Timepoint handling",
    "Gene selection",
    "Marker and program scoring",
    "Discovery-set balancing",
    "Experimental versus natural cohorts",
    "Large-file outputs"
  ),
  guiding_question = c(
    "Which cells and genes are removed before modeling?",
    "Why are experimental D14 and D15 merged into D14/15?",
    "Why does the workflow select 5,000 highly variable genes?",
    "Which marker sets support broad cell-type and IFN-program interpretation?",
    "Why are experimental samples capped at 600 cells per sample?",
    "Which cohort is used for discovery, and which is reserved for optional projection?",
    "Which outputs are too large for GitHub and need an external host?"
  )
) |>
  knitr::kable()
```

::: {.panel-tabset}

## Python

Run this from the repository root inside the case-study Docker image after the large data host is finalized, or locally in an equivalent Python environment:

```{bash}
#| eval: false
python scripts/prepare_gse154386_subset_python.py \
  --workdir data/external/GSE154386 \
  --max-cells-per-sample 600 \
  --n-hvgs 5000 \
  --seed 42
```

```{r}
#| label: show-python-prep-script
#| echo: false
#| results: asis
cat("<details><summary>Complete Python preparation script</summary>\n\n")
cat("```python\n")
cat(readLines(here::here("scripts", "prepare_gse154386_subset_python.py")), sep = "\n")
cat("\n```\n\n</details>\n")
```

## R

Run this from the repository root inside the case-study Docker image after the large data host is finalized, or locally in an equivalent R/Bioconductor environment:

```{bash}
#| eval: false
Rscript scripts/prepare_gse154386_subset_r.R \
  --workdir data/external/GSE154386 \
  --max-cells-per-sample 600 \
  --n-hvgs 5000 \
  --seed 42
```

```{r}
#| label: show-r-prep-script
#| echo: false
#| results: asis
cat("<details><summary>Complete R preparation script</summary>\n\n")
cat("```r\n")
cat(readLines(here::here("scripts", "prepare_gse154386_subset_r.R")), sep = "\n")
cat("\n```\n\n</details>\n")
```

:::

The Python version is closest to the original preprocessing script used during analysis development. The R version mirrors the same preparation logic with `SingleCellExperiment` objects so R-first learners can see how the subset can be prepared without switching languages.

## **From Cell Scores to Subject-Level Summaries**

Cell-level scores are useful, but cells from the same subject and timepoint are not independent people. We therefore summarize pattern usage at the subject-timepoint level.

::: {.panel-tabset}

## R

```{r}
#| label: subject-summary-r
subject_time <- readr::read_csv(
  here::here("data", "processed", "interpretation", "subject_level",
             "subject_time_pattern_means_long.csv"),
  show_col_types = FALSE
)

subject_time |>
  filter(pattern == "Pattern3") |>
  select(subject, timepoint_merged, day_merged_numeric,
         n_cells, mean_pattern_score) |>
  arrange(subject, day_merged_numeric) |>
  head(12) |>
  knitr::kable(digits = 3)
```

## Python

```{python}
#| label: subject-summary-python
subject_time = pd.read_csv(
    "data/processed/interpretation/subject_level/"
    "subject_time_pattern_means_long.csv"
)

subject_time.loc[
    subject_time["pattern"].eq("Pattern3"),
    ["subject", "timepoint_merged", "day_merged_numeric",
     "n_cells", "mean_pattern_score"]
].sort_values(["subject", "day_merged_numeric"]).head(12)
```

:::

This subject-level view is the main evidence layer for temporal claims. It avoids treating thousands of cells as thousands of independent subjects.

## **Classifying Patterns as Identity-Like or Activity-Like**

The case study uses a simple comparison:

- if variation by broad cell type dominates, the pattern is interpreted as identity-like
- if variation by timepoint dominates, the pattern is interpreted as activity-like
- if neither dominates, the pattern is mixed

::: {.panel-tabset}

## R

```{r}
#| label: identity-activity-table-r
rq2 <- readr::read_csv(
  here::here("data", "processed", "interpretation", "rq_tables",
             "RQ2_identity_vs_activity.csv"),
  show_col_types = FALSE
)

rq2 |>
  count(pattern_class) |>
  knitr::kable()
```

## Python

```{python}
#| label: identity-activity-table-python
rq2_py = pd.read_csv(
    "data/processed/interpretation/rq_tables/"
    "RQ2_identity_vs_activity.csv"
)

rq2_py["pattern_class"].value_counts().rename_axis(
    "pattern_class"
).reset_index(name="n")
```

:::

In the selected `K = 10` run, nine patterns are classified as identity-like and one pattern is classified as activity-like. That activity-like pattern is the IFN-associated temporal program.

***
