# **Data Import**
***

In this section, you will import the analysis-ready files used in the required case-study path. These files are small `CSV` and `JSON` files that are already included with the case study.

A `CSV` file is a rectangular table: rows are observations or summaries, and columns are variables. A `JSON` file stores structured metadata, such as model settings or diagnostic values. In R, the imported files become data frames. In Python, they become tables managed by `pandas`.

The raw GEO archive, `GSE154386_RAW.tar`, is not imported here. Downloading that file starts the optional full preprocessing workflow: extract 10x files, build single-cell objects, perform quality control, normalize counts, select genes, annotate broad cell types, and construct the discovery subset. That workflow is covered in **Data Wrangling and Preprocessing**. Here, you will import the selected-model files needed for the main analysis.

| File family | Example file | What one row means | Why you import it |
|---|---|---|---|
| Pattern summary | `RQ1_time_varying_patterns.csv` | One CoGAPS pattern. | Rank patterns by temporal variation and IFN association. |
| Top-gene table | `cogaps_K10_seed2_iter2000.top_genes.csv` | One gene-pattern entry. | Inspect which genes define each pattern. |
| Run metrics | `cogaps_K10_seed2_iter2000.metrics.json` | One model run summary. | Confirm `K`, seed, iterations, runtime, and diagnostic settings. |
| Diagnostic trace | `cogaps_K10_seed2_iter2000.trace.csv` | One saved trace point during the model run. | Check convergence-related traces before interpretation. |

The larger files needed to rerun CoGAPS from scratch are intentionally external:

- raw GEO download: `GSE154386_RAW.tar`
- preprocessed H5AD cache
- experimental discovery H5ADs
- RDS model objects
- checkpoint files

These files are documented in `data/artifact_manifest.json` and are distributed through Zenodo for full reproduction.

## **Import the Selected-Model Results**

Before importing data, build paths relative to the project root. Project-relative paths make the same code work on different computers and inside the Docker runtime.

::: {.panel-tabset}

## R

In R, `here::here()` builds a path from the project root. `readr::read_csv()` reads a CSV table, and `jsonlite::fromJSON()` reads a JSON file.

```{r}
#| label: import-file-map-r
# Build paths to the selected K = 10 R files.
r_selected_dir <- here::here(
  "data", "processed", "selected_model_k10", "r"
)

r_import_plan <- data.frame(
  file_role = c(
    "pattern summary",
    "top genes",
    "run metrics",
    "diagnostic trace"
  ),
  file_name = c(
    "RQ1_time_varying_patterns.csv",
    "cogaps_K10_seed2_iter2000.top_genes.csv",
    "cogaps_K10_seed2_iter2000.metrics.json",
    "cogaps_K10_seed2_iter2000.trace.csv"
  )
)

r_import_plan$relative_path <- file.path(
  "data", "processed", "selected_model_k10", "r",
  r_import_plan$file_name
)

r_import_plan$file_exists <- file.exists(
  file.path(here::here(), r_import_plan$relative_path)
)

r_import_plan |>
  select(file_role, relative_path, file_exists) |>
  knitr::kable()
```

```{r}
#| label: import-selected-r
# Read selected-model tables and metadata.
r_rq1 <- readr::read_csv(
  file.path(r_selected_dir, "RQ1_time_varying_patterns.csv"),
  show_col_types = FALSE
)

r_top_genes <- readr::read_csv(
  file.path(r_selected_dir, "cogaps_K10_seed2_iter2000.top_genes.csv"),
  show_col_types = FALSE
)

r_model_metrics <- jsonlite::fromJSON(
  file.path(r_selected_dir, "cogaps_K10_seed2_iter2000.metrics.json")
)

r_trace <- readr::read_csv(
  file.path(r_selected_dir, "cogaps_K10_seed2_iter2000.trace.csv"),
  show_col_types = FALSE
)

data.frame(
  imported_object = c("r_rq1", "r_top_genes", "r_trace"),
  rows = c(nrow(r_rq1), nrow(r_top_genes), nrow(r_trace)),
  columns = c(ncol(r_rq1), ncol(r_top_genes), ncol(r_trace)),
  first_columns = c(
    paste(head(names(r_rq1), 5), collapse = ", "),
    paste(head(names(r_top_genes), 4), collapse = ", "),
    paste(head(names(r_trace), 4), collapse = ", ")
  )
) |>
  knitr::kable()
```

```{r}
#| label: import-sanity-r
# Confirm that the imported files describe the expected selected model.
data.frame(
  check = c(
    "model rank K",
    "seed",
    "iterations",
    "CoGAPS threads",
    "number of pattern rows",
    "patterns represented in top-gene table",
    "diagnostic trace length"
  ),
  value = c(
    r_model_metrics$K,
    r_model_metrics$seed,
    r_model_metrics$n_iter,
    r_model_metrics$nThreads,
    nrow(r_rq1),
    length(unique(r_top_genes$pattern)),
    nrow(r_trace)
  )
) |>
  knitr::kable()
```

The imported R objects are now available for later sections. For example, this preview shows the most time-varying patterns in the selected model.

```{r}
#| label: import-preview-r
r_rq1 |>
  select(pattern, eta_timepoint, peak_timepoint,
         spearman_ifn_score_rho, candidate_ifn_pattern) |>
  arrange(desc(eta_timepoint)) |>
  head(5) |>
  knitr::kable(digits = 3)
```

## Python

In Python, `pathlib.Path()` builds project-relative paths. `pandas.read_csv()` reads CSV tables, and `json.load()` reads a JSON file.

```{python}
#| label: import-file-map-python
from pathlib import Path
import json
import pandas as pd

py_selected_dir = Path("data/processed/selected_model_k10/python")

py_import_plan = pd.DataFrame({
    "file_role": [
        "pattern summary",
        "top genes",
        "run metrics",
        "diagnostic trace",
    ],
    "file_name": [
        "RQ1_time_varying_patterns.csv",
        "cogaps_K10_seed2_iter2000.top_genes.csv",
        "cogaps_K10_seed2_iter2000.metrics.json",
        "cogaps_K10_seed2_iter2000.trace.csv",
    ],
})

py_import_plan["relative_path"] = py_import_plan["file_name"].map(
    lambda name: str(py_selected_dir / name)
)
py_import_plan["file_exists"] = py_import_plan["relative_path"].map(
    lambda path: Path(path).exists()
)

py_import_plan.loc[:, ["file_role", "relative_path", "file_exists"]]
```

```{python}
#| label: import-selected-python
# Read selected-model tables and metadata.
py_rq1 = pd.read_csv(py_selected_dir / "RQ1_time_varying_patterns.csv")
py_top_genes = pd.read_csv(
    py_selected_dir / "cogaps_K10_seed2_iter2000.top_genes.csv"
)

with open(py_selected_dir / "cogaps_K10_seed2_iter2000.metrics.json") as f:
    py_model_metrics = json.load(f)

py_trace = pd.read_csv(py_selected_dir / "cogaps_K10_seed2_iter2000.trace.csv")

pd.DataFrame({
    "imported_object": ["py_rq1", "py_top_genes", "py_trace"],
    "rows": [len(py_rq1), len(py_top_genes), len(py_trace)],
    "columns": [py_rq1.shape[1], py_top_genes.shape[1], py_trace.shape[1]],
    "first_columns": [
        ", ".join(py_rq1.columns[:5]),
        ", ".join(py_top_genes.columns[:4]),
        ", ".join(py_trace.columns[:4]),
    ],
})
```

```{python}
#| label: import-sanity-python
# Confirm that the imported files describe the expected selected model.
pd.DataFrame({
    "check": [
        "model rank K",
        "seed",
        "iterations",
        "CoGAPS threads",
        "number of pattern rows",
        "patterns represented in top-gene table",
        "diagnostic trace length",
    ],
    "value": [
        py_model_metrics["K"],
        2,
        py_model_metrics["n_iter"],
        py_model_metrics["cogaps_threads"],
        len(py_rq1),
        py_top_genes["pattern"].nunique(),
        len(py_trace),
    ],
})
```

The imported Python objects are now available for later sections. For example, this preview shows the most time-varying patterns in the selected model.

```{python}
#| label: import-preview-python
py_rq1.loc[
    :, ["pattern", "eta_timepoint", "peak_timepoint",
        "spearman_ifn_score_rho", "candidate_ifn_pattern"]
].sort_values("eta_timepoint", ascending=False).head(5)
```

:::

Both tabs import the same kinds of files. You only need to run the language you plan to use, but the two examples follow the same programming pattern: build a path, read a file, inspect the imported object, and run sanity checks.

The sanity-check tables confirm that the selected-model files describe the expected `K = 10`, `seed = 2`, `nIterations = 2000` run. The next sections explore these imported tables before reshaping them for interpretation.

***
