# **Data Wrangling and Preprocessing**
***

Data Import selected the `K = 10` model for the core analysis and loaded the compact selected-model files. Data Exploration checked what those files contain: subjects, infection days, broad cell types, gene information, and pattern-score summaries. This section asks a different question: why are those files safe enough to interpret?

Before CoGAPS can support a biological claim, each cell has to stay connected to its subject, infection day, cohort, quality-control information, and broad immune-cell identity. If that chain breaks, a model pattern can look like dengue-associated biology while actually reflecting damaged cells, sequencing-depth differences, unstable gene selection, or one overrepresented sample.

You will first trace how the imported source files became a CoGAPS-ready experimental discovery input. Then you will use the included selected-model tables to make the smaller subject-level summaries that Visualization and Analysis build on.

:::{.disclaimer_box}
::::{.disclaimer_box_header}
How to use this section
::::
::::{.inner_block}
- **Core path:** read the preprocessing guide, then run the small table-wrangling chunks under **Build Interpretation-Ready Summary Tables**.
- **Reproducibility notes:** use the visible thresholds, matrix-orientation notes, manifest descriptions, and runnable rebuild commands to understand where the included files came from.
- **Additional Information:** use the advanced reproduction map and code explanations there if you want to rebuild or inspect the large preprocessing artifacts in more detail.
::::
:::

:::{.definition_box}
::::{.definition_box_header}
Preprocessing vocabulary
::::
::::{.inner_block}
- {{< glossary "Preprocessing" display="Preprocessing" >}} turns imported single-cell data into a model-ready object whose cells, genes, metadata, and file provenance can be interpreted together.
- {{< glossary "Quality control" display="Quality control" >}} asks which cells and genes are reliable enough to keep.
- {{< glossary "Library-size normalization" display="Library-size normalization" >}} and {{< glossary "Log transformation" display="log transformation" >}} put expression values on a more comparable scale.
- {{< glossary "Highly variable genes" display="Highly variable gene" >}} selection and the {{< glossary "Balanced discovery set" display="balanced discovery subset" >}} define what CoGAPS learns from.
- {{< glossary "Boolean mask" display="Boolean mask" >}} logic is how preprocessing code applies yes/no filters, such as keeping cells that pass gene-count and mitochondrial-signal thresholds.
- {{< glossary "Matrix orientation" display="Matrix orientation" >}} records whether genes or cells are stored as rows, which matters when model outputs are joined back to biology.
::::
:::

Together, these ideas explain why this section spends time on filtering, scaling, gene selection, metadata, and sample balance before showing any new figures. A temporal immune-program analysis only works if the model is learning biological structure rather than technical artifacts or sampling imbalance.

## Preprocessing Decisions as Interpretation Safeguards

Preprocessing is where the case study decides what CoGAPS is allowed to learn from. These choices are not just software settings. They decide which cells are reliable, which genes enter the model, which samples represent the main time-course question, and how much influence each sample can have.

The first safeguard is signal quality. The preparation scripts start from a labeled {{< glossary "SingleCellExperiment" display="SingleCellExperiment" >}} or {{< glossary "AnnData" display="AnnData" >}} object, then keep cells with at least `200` {{< glossary "Detected gene" display="detected genes" >}} and `500` {{< glossary "Total counts" display="total counts" >}}, flag cells above `20%` {{< glossary "Mitochondrial fraction" display="mitochondrial signal" >}}, remove sample-specific upper outliers, and keep genes detected in at least `10` cells. In the code, these keep/remove decisions are implemented with Boolean masks: each cell or gene either passes a condition or it does not. The thresholds are practical filters, not biological boundaries, but they reduce the chance that low-coverage cells, damaged cells, or unstable rare genes drive a pattern.

The second safeguard is comparability. After filtering, cells still differ in sequencing depth, so the workflow applies library-size normalization and a log transformation before choosing `5,000` highly variable genes. That combination gives CoGAPS a tractable expression space with enough variation to learn from. It does not erase every subject, batch, identity, or infection-response effect; instead, it defines the expression scale that later evidence has to be interpreted on.

The third safeguard is biological context. The workflow merges nearby late experimental days `D14` and `D15` as `D14/15`, scores broad PBMC marker sets, and keeps response-program scores available as guideposts. Those annotations help later sections ask whether a pattern is mostly identity-like, activity-like, or mixed. They are not final biological labels, and the `D14/15` merge trades one-day resolution for a more stable late-timepoint comparison.

The final safeguard is a fair discovery set. The experimental cohort is used for the main model because its infection days form the aligned time course in this case study, while the natural infection cohort is reserved for optional projection work. Within that experimental cohort, the workflow samples up to `600` cells per sample using `seed = 42`. That cap keeps one large sample from controlling the factorization, and the seed makes the selected subset rebuildable.

:::{.think_question}
::::{.think_question_header}
Checkpoint
::::
::::{.inner_block}
Which preprocessing choice would you revisit if a pattern looked like PBMC identity, sample imbalance, or technical stress?

<details>
<summary>Show answer</summary>

Different interpretations point back to different safeguards. A pattern that mainly separates PBMC identities should send you back to the marker-score and broad-cell-type context. A pattern dominated by one subject or one timepoint points back to the balancing rule and random seed. A pattern that looks like damaged cells or poor RNA recovery points back to the quality-control filters. Normalization and highly variable gene selection sit in the middle because they shape which expression differences are available for any of these patterns to learn.

</details>
::::
:::

For the full parameter-level reference, see the preprocessing decision reference in **Additional Information**. The shorter explanation here is the part you need to carry forward: these choices do not prove a biological program, but they define the evidence that can be interpreted later.

## From Source Files to CoGAPS-Ready Discovery Input

The safeguards above become the workflow below. The first part of the workflow is source-file import: raw 10x files from GEO are read together with sample names so expression counts can be connected to cells, genes, subjects, cohorts, and infection days. That import step creates a labeled {{< glossary "Single-cell object" display="single-cell object" >}}, which then becomes the starting point for quality control, normalization, gene selection, annotation, discovery-set construction, and CoGAPS input saving.

This map is provenance for the included files. You do not need to download the raw GEO archive or rebuild the large single-cell objects for the core path. The purpose here is to see how the small tables used later can be traced back to source files, preprocessing decisions, and a saved manifest.

Read the diagram from top to bottom as a connected evidence chain. At each step, ask what biological comparison would be affected if that step were missing or mislabeled.

```{mermaid}
flowchart TB
  classDef source fill:#eaf3ff,stroke:#2d6da3,stroke-width:2px,color:#102a43
  classDef process fill:#f7f9fb,stroke:#6b7c8f,stroke-width:1.5px,color:#102a43
  classDef decision fill:#fff4df,stroke:#b36b00,stroke-width:2px,color:#102a43
  classDef output fill:#e9f7ef,stroke:#2f855a,stroke-width:2px,color:#102a43
  classDef note fill:#f3eefc,stroke:#6b46c1,stroke-width:1.5px,color:#102a43
  classDef stage fill:#1f4e79,stroke:#123047,stroke-width:2px,color:#ffffff

  A["Source data<br/><b>GEO GSE154386</b><br/>raw 10x archive"]:::source

  subgraph S0[" "]
    P0["1. Import source files<br/>and labels"]:::stage
    B["Parse sample names<br/>subject, cohort, day"]:::process
    C["Read 10x triplets<br/>matrix + barcodes + features"]:::process
    D["Create labeled<br/>single-cell object"]:::source
  end

  subgraph S1[" "]
    P1["2. Apply preprocessing<br/>decisions"]:::stage
    E{{"QC filters<br/>cells and genes"}}:::decision
    F["Normalize counts<br/>library-size + log scale"]:::process
    G{{"Select HVGs<br/>5,000 genes"}}:::decision
    H["Score marker/program sets<br/>PBMC identity + response-program scores"]:::process
  end

  subgraph S2[" "]
    P2["3. Define the case-study<br/>discovery set"]:::stage
    I{{"Assign broad PBMC identity"}}:::decision
    J{{"Use experimental cohort<br/>for the main time course"}}:::decision
    K["Balance samples<br/>up to 600 cells each"]:::process
  end

  subgraph S3[" "]
    P3["4. Save outputs<br/>for lesson and reproduction use"]:::stage
    L["Optional reproduction:<br/>CoGAPS-ready model input files<br/>and preprocessing manifest"]:::output
    M["Included lesson tables:<br/>interpretation summaries<br/>used after model fitting"]:::output
    N["Optional extension:<br/>natural-cohort projection target"]:::note
  end

  A --> P0 --> B --> C --> D --> P1 --> E --> F --> G --> H --> P2 --> I --> J --> K --> P3
  P3 --> M
  P3 --> L
  P3 --> N
  J -. "not used for the primary discovery model" .-> N

  style S0 fill:#f8fafc,stroke:#cbd5e1,stroke-width:1px,color:#102a43
  style S1 fill:#f8fafc,stroke:#cbd5e1,stroke-width:1px,color:#102a43
  style S2 fill:#f8fafc,stroke:#cbd5e1,stroke-width:1px,color:#102a43
  style S3 fill:#f8fafc,stroke:#cbd5e1,stroke-width:1px,color:#102a43
```

<details style="overflow-wrap: anywhere;">
<summary>Show the files represented by the final graph boxes</summary>

**Source import inputs**

- GEO accession `GSE154386`
- Raw 10x archive with sample-level matrix, barcode, and feature files
- Sample names that encode subject, cohort, and infection day

**Optional reproduction model-input files**

- `gse154386_experimental_discovery_genes_x_cells.rds`
- `gse154386_experimental_discovery_genes_x_cells.h5ad`
- `gse154386_experimental_discovery_cells_x_genes.h5ad`
- `preprocessing_manifest.json`

**Core included analysis tables**

- `subject_time_pattern_means_long.csv`
- `subject_time_celltype_pattern_means_long.csv`
- `pattern_annotation_table.csv`
- `RQ1_temporal_programs.csv`
- `RQ2_identity_vs_activity.csv`
- `RQ3_ifn_associated_program.csv`

**Optional projection target files**

- `gse154386_natural_projection_target.rds`
- `gse154386_natural_projection_target.h5ad`

</details>

## Preprocessing Guide: How the Discovery Input Is Created

This guide summarizes how the included CoGAPS-ready artifacts were prepared. You do not need to run these preparation steps for the core path. They are here so the small tables you use later have a visible origin: each file came from a labeled single-cell object, a documented set of preprocessing decisions, and a saved manifest.

:::{.reproducibility_box}
::::{.reproducibility_box_header}
Reproducibility note
::::
::::{.inner_block}
The complete preparation scripts live in the repository and are listed in **Additional Information -> Advanced Reproduction Script Map**. Use that map if you want to rebuild the large preprocessing artifacts from source files. Additional Information also includes code explanations for the preprocessing steps, so the main lesson can stay focused on what the prepared inputs mean.

These scripts download or reuse the GEO source archive, import the 10x files into a labeled single-cell object, apply the preprocessing decisions summarized below, construct the balanced experimental discovery set, and save orientation-specific matrices plus `preprocessing_manifest.json`. They prepare the inputs needed before a selected CoGAPS model can be run. They do not run CoGAPS itself, and they do not rerun the rank sweep assessed in Data Import.

```{bash}
#| eval: false
Rscript scripts/prepare_gse154386_subset_r.R \
  --workdir data/external/GSE154386 \
  --max-cells-per-sample 600 \
  --n-hvgs 5000 \
  --seed 42
```

```{bash}
#| eval: false
python scripts/prepare_gse154386_subset_python.py \
  --workdir data/external/GSE154386 \
  --max-cells-per-sample 600 \
  --n-hvgs 5000 \
  --seed 42
```
::::
:::

Use the guide below as a narrated preparation pass. The details matter because every later visualization depends on the same chain: reliable cells and genes, comparable expression values, documented biological context, a balanced experimental discovery set, and correctly oriented model-input files.

### Quality Control Protects the Input Signal

Quality control is not about making the dataset look cleaner for its own sake. It protects the latent-factor analysis from learning patterns that mostly reflect low RNA recovery, damaged cells, or extreme count profiles. In a PBMC infection study, that matters because a technical stress pattern can be mistaken for an immune-response program if it is not filtered or at least reported.

The filtered object keeps cells with enough detected genes and total counts, removes high mitochondrial-signal cells and sample-specific upper outliers, and removes genes detected in too few cells. Those keep/remove steps are where Boolean masks appear in the preparation code: a cell that passes the mask stays in the object, and a cell that fails the mask is excluded from the CoGAPS input.

### Normalization and Highly Variable Genes Define the Expression Space

After QC, cells still differ in sequencing depth. {{< glossary "Library-size normalization" display="Library-size normalization" >}} and {{< glossary "Log transformation" display="log transformation" >}} make expression values more comparable before marker scoring and variable-gene selection. {{< glossary "Highly variable genes" display="Highly variable gene" >}} selection then focuses the model on genes that carry useful variation across cells and timepoints.

The CoGAPS input is built from normalized, log-transformed expression values and a documented set of `5,000` variable genes. This gene space can include infection-response genes, but it can also emphasize stable PBMC identity differences. That is why later sections compare pattern scores with timepoint summaries, broad cell-type evidence, top genes, and directionality rather than treating a single score as a conclusion.

### Marker and Program Scores Add Biological Guideposts

CoGAPS learns patterns from the expression matrix, but numbered patterns do not explain themselves. Marker and program scores add biological guideposts before you interpret the model. A {{< glossary "Marker score" display="marker score" >}} asks whether each cell expresses genes associated with broad PBMC identities, such as T cells, monocytes, natural killer cells, B cells, plasmablasts, dendritic cells, or neutrophils. A {{< glossary "Program score" display="program score" >}} uses the same scoring idea for biological activity, such as interferon-associated response.

These scores are useful precisely because they are modest. A high `T_cell_score` or `Monocyte_score` can suggest a broad immune identity, but it is still an annotation aid rather than a definitive cell-type call. A high `ifn_program_score` can flag antiviral-response activity, but it does not prove that a CoGAPS pattern is caused by interferon signaling. Later sections use the scores as guideposts beside CoGAPS cell scores, top genes, subject trajectories, and model diagnostics.

### The Discovery Set Focuses the Main Time Course

The experimental cohort is the main discovery cohort because its timepoints are aligned across subjects. The natural infection cohort is biologically interesting, but it is less controlled for the primary temporal discovery question and is better handled as optional projection material.

Within the experimental cohort, the workflow samples up to `600` cells per sample using `seed = 42`. That rule is a modeling decision: it keeps one large sample from dominating the factorization and makes the selected subset reproducible.

The selected model therefore learns from an experimental discovery matrix with documented timepoint labels, broad PBMC identities, and sample-level balancing. It is not trained on every cell from every cohort in the source study, and that distinction should travel with any biological claim.

### The Saved Files Must Be CoGAPS-Ready

At this point, the preprocessing workflow has produced the discovery data that CoGAPS needs for model fitting. The object is no longer just an imported single-cell dataset. It has been filtered, normalized, restricted to highly variable genes, limited to the experimental cohort, labeled with biological metadata, and balanced so that no single sample dominates the model.

This is the right place to pause before visualization or analysis. Visualization and interpretation depend on CoGAPS outputs, but CoGAPS can only be interpreted biologically if its input matrix is well defined.

By the time the files are saved, the discovery input has a few properties you should expect to carry forward:

- low-quality cells, technical outliers, and unstable rare genes have been filtered;
- expression values are normalized, log-transformed, checked for non-negativity, and restricted to informative genes;
- each cell remains linked to subject, sample, cohort, infection day, merged timepoint, and broad PBMC identity;
- the main CoGAPS discovery matrix uses the experimental cohort only;
- each sample contributes up to the same number of cells;
- orientation-specific files record which dimension contains genes and which contains cells;
- `preprocessing_manifest.json` records shapes, parameters, seed, and output paths.

The key distinction is between the biological annotation context and the matrix passed to CoGAPS. The annotation context keeps cells connected to sample, subject, cohort, infection day, merged timepoint, and broad PBMC identity. The CoGAPS input matrix is the expression matrix used for factorization, where the model learns {{< glossary "Gene loading" display="gene loadings" >}} and {{< glossary "Cell score" display="cell scores" >}}.

{{< glossary "Matrix orientation" display="Matrix orientation" >}} tells CoGAPS which dimension contains genes and which dimension contains cells. If the orientation is wrong, gene loadings and cell scores can be swapped or mislabeled. The R preparation script saves a genes-by-cells `SingleCellExperiment` discovery object. The Python preparation script saves one cells-by-genes `AnnData` object for metadata joins and one genes-by-cells `AnnData` object for PyCoGAPS; the selected Python runner uses the genes-by-cells file with `transposeData = False`.

:::{.think_question}
::::{.think_question_header}
Checkpoint
::::
::::{.inner_block}
Why does the preprocessing guide spend so much time on metadata and matrix orientation before you look at the biological interpretation?

<details>
<summary>Show answer</summary>

CoGAPS patterns are interpretable only after the model output is joined back to the right biological context. Metadata tells you which subject, day, cohort, and broad PBMC identity each cell belongs to. Matrix orientation tells you which model weights belong to genes and which belong to cells. If either layer is wrong, the later figures can look polished while pointing to the wrong biological evidence.

</details>
::::
:::

## Build Interpretation-Ready Summary Tables

From this point forward, you use the small files already included in `data/processed`. You are no longer preprocessing raw single-cell data. Instead, you are revisiting {{< glossary "Selected-model outputs" display="selected-model outputs" >}} that Data Exploration introduced and reshaping them into tables that support later biological interpretation. The chunks in this part are part of the core path and should run with the included files.

The small wrangling work has three products. First, you read the subject-timepoint-pattern table so temporal evidence stays tied to participants and sampled days. Next, you reshape the same evidence into an all-pattern time-course view that is easier to scan before plotting. Finally, you preview the identity/activity evidence table so the analysis can ask whether a pattern looks activity-like, identity-like, or mixed.

### Step 1: Read Subject-Timepoint Pattern Summaries

{{< glossary "Cell score" display="Cell-level CoGAPS scores" >}} are useful, but cells from the same subject and timepoint are not independent people. If you treated thousands of cells as thousands of independent biological replicates, the analysis would exaggerate the amount of independent evidence. The {{< glossary "Subject-timepoint summary" display="subject-timepoint summary" >}} is the better evidence unit for temporal claims in this case study.

The first chunk rereads a prepared {{< glossary "Long format" display="long table" >}} that you saw structurally in Data Exploration. Here the purpose is different: each row, representing one subject at one infection day for one CoGAPS pattern, becomes the starting point for derived summaries.

::: {.panel-tabset}

#### R

```{r}
#| label: subject-time-read-r
subject_time <- readr::read_csv(
  here::here("data", "processed", "interpretation", "subject_level",
             "subject_time_pattern_means_long.csv"),
  show_col_types = FALSE
)
```

#### Python

```{python}
#| label: subject-time-read-python
subject_time = pd.read_csv(
    "data/processed/interpretation/subject_level/"
    "subject_time_pattern_means_long.csv"
)
```

:::

Now inspect the subject-timepoint summaries without choosing one biological candidate yet. The output keeps `n_cells` visible so you can see how many cells contributed to each subject-timepoint-pattern summary.

::: {.panel-tabset}

#### R

```{r}
#| label: subject-summary-r
subject_time |>
  mutate(pattern_number = as.integer(sub("Pattern", "", pattern))) |>
  select(subject, timepoint_merged, day_merged_numeric,
         pattern, pattern_number, n_cells, mean_pattern_score) |>
  arrange(subject, day_merged_numeric, pattern_number) |>
  select(-pattern_number) |>
  head(12) |>
  knitr::kable(digits = 3)
```

#### Python

```{python}
#| label: subject-summary-python
(
    subject_time
    .assign(pattern_number=lambda df: (
        df["pattern"].str.replace("Pattern", "", regex=False).astype(int)
    ))
    .loc[:, ["subject", "timepoint_merged", "day_merged_numeric",
             "pattern", "pattern_number", "n_cells", "mean_pattern_score"]]
    .sort_values(["subject", "day_merged_numeric", "pattern_number"])
    .drop(columns="pattern_number")
    .head(12)
)
```

:::

This table shows why the case study often works at the subject-timepoint level: it keeps the time course visible while avoiding the mistake of treating every cell as an independent subject.

### Step 2: Reshape All Patterns into a Readable Time Course

The next chunk summarizes the subject-timepoint table across subjects, then reshapes the result so patterns become rows and infection days become columns. You have already seen reshaping in earlier sections, so focus here on why the display changes: the {{< glossary "Wide table" display="wide table" >}} makes the selected model's time-course structure easier to scan without assigning a final biological label.

::: {.panel-tabset}

#### R

```{r}
#| label: all-pattern-subject-wide-r
timepoint_order <- c("D0", "D2", "D4", "D6", "D8", "D10", "D14/15", "D28")

all_pattern_timecourse <- subject_time |>
  mutate(
    timepoint_merged = factor(timepoint_merged, levels = timepoint_order),
    pattern_number = as.integer(sub("Pattern", "", pattern))
  ) |>
  group_by(pattern, pattern_number, timepoint_merged, day_merged_numeric) |>
  summarize(
    n_subjects = n_distinct(subject),
    cells_summarized = sum(n_cells),
    mean_subject_score = mean(mean_pattern_score),
    .groups = "drop"
  ) |>
  arrange(pattern_number, day_merged_numeric)

all_pattern_timecourse_wide <- all_pattern_timecourse |>
  select(pattern, pattern_number, timepoint_merged, mean_subject_score) |>
  tidyr::pivot_wider(
    names_from = timepoint_merged,
    values_from = mean_subject_score
  ) |>
  arrange(pattern_number) |>
  select(-pattern_number)

all_pattern_timecourse_wide |>
  knitr::kable(digits = 3)
```

#### Python

```{python}
#| label: all-pattern-subject-wide-python
py_timepoint_order = ["D0", "D2", "D4", "D6", "D8", "D10", "D14/15", "D28"]

all_pattern_timecourse_py = (
    subject_time
    .assign(pattern_number=lambda df: (
        df["pattern"].str.replace("Pattern", "", regex=False).astype(int)
    ))
    .groupby(
        ["pattern", "pattern_number", "timepoint_merged", "day_merged_numeric"],
        as_index=False
    )
    .agg(
        n_subjects=("subject", "nunique"),
        cells_summarized=("n_cells", "sum"),
        mean_subject_score=("mean_pattern_score", "mean"),
    )
    .sort_values(["pattern_number", "day_merged_numeric"])
)

all_pattern_timecourse_wide_py = (
    all_pattern_timecourse_py
    .pivot(index=["pattern", "pattern_number"],
           columns="timepoint_merged",
           values="mean_subject_score")
    .loc[:, py_timepoint_order]
    .reset_index()
    .sort_values("pattern_number")
    .drop(columns="pattern_number")
)

all_pattern_timecourse_wide_py.round(3)
```

:::

The {{< glossary "Wide table" display="wide view" >}} is a wrangling product, not a conclusion. It prepares a compact view of all selected-model patterns so the visualization and analysis sections can decide which patterns deserve biological interpretation.

:::{.think_question}
::::{.think_question_header}
Checkpoint
::::
::::{.inner_block}
For a temporal claim, what is the evidence unit: one cell, one subject, or one subject at one timepoint? Why does that choice matter for how confident the figure should look?

<details>
<summary>Show answer</summary>

For the temporal claims in this case study, the evidence unit is a subject at one timepoint. Individual cells are useful measurements, but cells from the same subject and day are not independent biological replicates. Summarizing by subject-timepoint keeps the repeated-subject design visible and prevents the analysis from looking more precise than the study design supports.

</details>
::::
:::

### Step 3: Prepare the Identity/Activity Table

The final wrangling task prepares the table used for the identity/activity research question. This table asks whether a pattern changes mostly across infection days, mostly across broad immune-cell identities, or across both. That distinction matters because a pattern can look temporal if the mix of PBMC cell types changes over time, even if the pattern itself is mostly a cell-identity signal.

The heuristic compares two summaries. `eta_timepoint` measures how much pattern variation is associated with infection day. `eta_broad_cell_type` measures how much pattern variation is associated with broad PBMC identity. A pattern with stronger timepoint variation is a better candidate for dynamic infection activity, while a pattern with stronger cell-type variation is more likely to represent stable PBMC identity or composition.

The summary table already stores this classification for each selected `K = 10` pattern. Here you inspect the prepared table so the later analysis can focus on interpretation rather than table construction. The file name begins with `RQ2` because this table belongs to the identity/activity question in the analysis files; Data Import has already handled the model-rank question.

::: {.panel-tabset}

#### R

```{r}
#| label: identity-activity-table-r
rq2 <- readr::read_csv(
  here::here("data", "processed", "interpretation", "rq_tables",
             "RQ2_identity_vs_activity.csv"),
  show_col_types = FALSE
)
```

#### Python

```{python}
#| label: identity-activity-table-python
rq2_py = pd.read_csv(
    "data/processed/interpretation/rq_tables/"
    "RQ2_identity_vs_activity.csv"
)
```

:::

First, count how many selected-model patterns fall into each broad interpretation class. This is a quick orientation step, not a biological conclusion.

::: {.panel-tabset}

#### R

```{r}
#| label: identity-activity-count-r
rq2 |>
  count(pattern_class) |>
  knitr::kable()
```

#### Python

```{python}
#| label: identity-activity-count-python
rq2_py["pattern_class"].value_counts().rename_axis(
    "pattern_class"
).reset_index(name="n")
```

:::

Next, rank patterns by the timepoint effect size while keeping their cell-type evidence and peak timepoint visible. This preview prepares the identity/activity question without yet assigning a final biological label.

::: {.panel-tabset}

#### R

```{r}
#| label: identity-activity-preview-r
rq2 |>
  select(pattern, eta_timepoint, eta_broad_cell_type,
         pattern_class, dominant_broad_cell_type,
         subject_peak_timepoint) |>
  arrange(desc(eta_timepoint)) |>
  head(6) |>
  knitr::kable(digits = 3)
```

#### Python

```{python}
#| label: identity-activity-preview-python
(
    rq2_py
    .loc[:, ["pattern", "eta_timepoint", "eta_broad_cell_type",
             "pattern_class", "dominant_broad_cell_type",
             "subject_peak_timepoint"]]
    .sort_values("eta_timepoint", ascending=False)
    .head(6)
)
```

:::

In the selected `K = 10` run, most patterns are classified as identity-like. One pattern is classified as activity-like by this heuristic, which flags it for later visual and analytic follow-up. Treat that label as a guide, not a conclusion: top genes, subject trajectories, directionality, and model diagnostics still have to support the biological interpretation.

:::{.think_question}
::::{.think_question_header}
Checkpoint
::::
::::{.inner_block}
If a pattern has high cell-type variation but low timepoint variation, what kind of biological story would be most cautious? If a pattern has high timepoint variation, what additional evidence would you want before calling it an infection-response program?

<details>
<summary>Show answer</summary>

A pattern with high cell-type variation but low timepoint variation should be described cautiously as identity-like or composition-associated, not as a dynamic infection-response program. A pattern with high timepoint variation is a stronger candidate for infection-response activity, but it still needs supporting evidence. You would want to inspect top genes, subject-level trajectories, directionality relative to baseline, and model diagnostics before assigning a biological label.

</details>
::::
:::

Before moving from preprocessing to visualization, keep the following interpretation safeguards in view.

:::{.definition_box}
::::{.definition_box_header}
What to carry forward
::::
::::{.inner_block}
- The preprocessing choices define the discovery set that the selected CoGAPS model learns from.
- The runnable summaries use subject-timepoint evidence rather than treating every cell as an independent replicate.
- Marker and program scores are guideposts. They help guide interpretation, but they do not replace top genes, trajectories, directionality, or diagnostics.
- Matrix orientation and metadata joins have to stay correct, because gene loadings and cell scores only become biological evidence after they are connected back to the right genes, cells, subjects, and days.
::::
:::

By the end of this section, you have followed the path from source files to a CoGAPS-ready experimental discovery input, then from cell-level CoGAPS scores to subject-timepoint summaries. The next section uses those summaries to visualize temporal structure, PBMC identity structure, and candidate patterns that Analysis will test more formally.

***
