# **Data Wrangling and Preprocessing**
***

The previous section showed two import paths: the standard path that reads the small selected-model files, and the optional source-file path that turns large 10x Genomics files into a labeled single-cell object. This section starts after that import step. The question now is how an imported single-cell object becomes a trustworthy discovery set for temporal CoGAPS analysis.

Before CoGAPS can discover temporal immune programs, every cell needs to stay connected to its subject, infection day, cohort, quality-control information, and broad immune-cell identity. Without that structure, a model can still find patterns, but those patterns may reflect damaged cells, sequencing-depth differences, unstable gene selection, or one overrepresented sample rather than dengue-associated biology.

This section shows how the case-study discovery set was produced and how the selected CoGAPS cell scores are reshaped for interpretation. The goal is not to memorize every preprocessing command. The goal is to understand how each decision protects the later biological claim that a pattern reflects temporal infection-response activity rather than technical noise, PBMC composition, or sampling imbalance.

This section separates two kinds of work:

- **Full reproduction path:** start from the imported single-cell object, perform quality control, normalize counts, select highly variable genes, annotate broad immune-cell types, and construct the discovery subset.
- **Standard path:** use the small selected-model files in `data/processed` to summarize CoGAPS scores by subject, timepoint, and broad cell type.

You will see both kinds of work in this section. First, you will read the preprocessing workflow in pieces so the source of the analysis files is transparent. Then you will return to the included selected-model files and practice the wrangling moves used for the rest of the case study.

:::{.disclaimer_box}
::::{.disclaimer_box_header}
How to use this section
::::
::::{.inner_block}
If you are following the **standard path**, read the full-reproduction preprocessing code as provenance and begin running code again at **Standard Path: Return to the Case-Study Files**. If you are following the **full reproduction path**, use the complete scripts shown later in this section rather than copying each excerpt one by one. The K-selection sweep is not rerun here; it remains precomputed evidence that you will inspect in **Data Analysis**.
::::
:::

## From Imported Object to the Discovery Set

The optional import step produced a labeled single-cell object: expression counts are connected to cells, genes, subjects, cohorts, and infection days. The workflow below begins from that object and ends with the analysis-ready discovery set used for CoGAPS interpretation. Read the diagram from top to bottom as a map of decisions. At each step, ask what biological comparison would be affected if that decision changed.

```{mermaid}
flowchart TB
  classDef source fill:#eaf3ff,stroke:#2d6da3,stroke-width:2px,color:#102a43
  classDef process fill:#f7f9fb,stroke:#6b7c8f,stroke-width:1.5px,color:#102a43
  classDef decision fill:#fff4df,stroke:#b36b00,stroke-width:2px,color:#102a43
  classDef output fill:#e9f7ef,stroke:#2f855a,stroke-width:2px,color:#102a43
  classDef note fill:#f3eefc,stroke:#6b46c1,stroke-width:1.5px,color:#102a43
  classDef stage fill:#1f4e79,stroke:#123047,stroke-width:2px,color:#ffffff

  A["Imported single-cell object<br/>counts + subject/cohort/day metadata"]:::source

  subgraph S1[" "]
    P1["1. Apply preprocessing<br/>decisions"]:::stage
    E{{"QC filters<br/>cells and genes"}}:::decision
    F["Normalize counts<br/>library-size + log scale"]:::process
    G{{"Select HVGs<br/>5,000 genes"}}:::decision
    H["Score marker/program sets<br/>PBMC identity + IFN clues"]:::process
  end

  subgraph S2[" "]
    P2["2. Define the case-study<br/>discovery set"]:::stage
    I{{"Assign broad PBMC identity"}}:::decision
    J{{"Use experimental cohort<br/>for the main time course"}}:::decision
    K["Balance samples<br/>up to 600 cells each"]:::process
  end

  subgraph S3[" "]
    P3["3. Create files<br/>used later"]:::stage
    L["Large reproduction files<br/>H5AD/RDS + manifests"]:::output
    M["Small analysis tables<br/>composition, scores, RQ tables"]:::output
    N["Optional natural cohort target<br/>reserved for projection"]:::note
  end

  A --> P1 --> E --> F --> G --> H --> P2 --> I --> J --> K --> P3
  P3 --> M
  P3 --> L
  P3 --> N
  J -. "not part of standard path" .-> N

  style S1 fill:#f8fafc,stroke:#cbd5e1,stroke-width:1px,color:#102a43
  style S2 fill:#f8fafc,stroke:#cbd5e1,stroke-width:1px,color:#102a43
  style S3 fill:#f8fafc,stroke:#cbd5e1,stroke-width:1px,color:#102a43
```

:::{.definition_box}
::::{.definition_box_header}
Terms to know before reading the preprocessing code
::::
::::{.inner_block}
- **Quality control:** Filtering and diagnostic checks used to remove cells or genes that are unlikely to support reliable biological interpretation.
- **Boolean mask:** A vector of `TRUE`/`FALSE` values in R or `True`/`False` values in Python used to keep or remove rows, columns, cells, or genes.
- **Library-size normalization:** A scaling step that reduces differences caused by cells having different total sequencing counts.
- **Log transformation:** A transformation that compresses large expression values so comparisons are less dominated by the most abundant transcripts.
- **Highly variable genes:** Genes selected because they vary substantially across cells or samples and are therefore informative for learning latent patterns.
- **Balanced discovery subset:** The experimental subset used for CoGAPS, with up to 600 cells per sample so that one sample does not dominate the learned patterns.
::::
:::

Those terms are the vocabulary of the preprocessing workflow. They also explain why the code is careful about filtering, scaling, gene selection, and sample balance: a temporal immune-program analysis only works if the model is learning biological structure rather than technical artifacts or sampling imbalance.

## Preprocessing Decisions That Shape Interpretation

Every row in the table below is an analysis decision, not just a software setting. These choices define which cells are considered reliable, which genes are available for factorization, which timepoints are compared, and which cohort is used for temporal discovery.

Three decisions deserve special attention as you read. Quality control affects whether technical stress can masquerade as immune activation. Highly variable gene selection affects which biological signals CoGAPS is allowed to learn. Sample balancing affects whether the model is dominated by one subject or day. These decisions are practical, but they are also biological because they shape what the later patterns can mean.

| Decision | Value used | Biological or analytical reason | What it affects | Limitation or sensitivity |
|---|---|---|---|---|
| Imported single-cell object | Labeled `SingleCellExperiment` or `AnnData` object with counts and sample metadata | Keeps expression values aligned with subject, cohort, infection day, cell, and gene labels. | Starting point for QC, normalization, annotation, and discovery-set construction. | Incorrect metadata or matrix orientation would compromise downstream biological interpretation. |
| Minimum genes per cell | `MIN_GENES = 200` | Cells with very few detected genes provide too little transcriptome information for learning coordinated gene-expression programs. | Which cells enter normalization and CoGAPS input preparation. | More or less stringent cutoffs could change rare-cell retention. |
| Minimum counts per cell | `MIN_COUNTS = 500` | Low-count cells are more likely to reflect poor RNA recovery than interpretable immune activity. | Which cells remain after QC. | Count thresholds are practical filters, not biological boundaries. |
| Mitochondrial fraction filter | `MAX_PCT_MT = 20` | High mitochondrial signal can indicate stressed or damaged cells, which can create technical stress-like programs. | QC cell retention. | High mitochondrial signal can be technical or biological, so this cutoff should be reported. |
| Sample-specific upper outliers | 99.5th percentile for high gene count and high total count flags | Extreme count profiles can represent multiplets or sample-specific technical outliers. | Removes cells that may otherwise dominate summaries. | Outlier definitions depend on each sample's distribution. |
| Minimum cells per gene | `MIN_CELLS_PER_GENE = 10` | Genes detected in very few cells are hard to interpret as stable temporal programs. | Gene set available for HVG selection and CoGAPS. | Very rare but biologically meaningful genes may be excluded. |
| Normalization | Library-size normalization followed by log transform | Cells are sequenced to different depths; normalization puts expression values on a more comparable scale. | Expression values used for marker scoring, HVG selection, and downstream summaries. | Normalization reduces technical scale differences but does not remove all batch or subject effects. |
| Highly variable genes | `N_HVGS = 5000` | CoGAPS should focus on genes carrying useful cell-to-cell or time-course variation, not genes that are nearly constant. | CoGAPS input gene space. | Different HVG strategies can change the patterns; the Python and R reproduction scripts use related but not identical implementations. |
| Timepoint merge | Experimental `D14` and `D15` are merged to `D14/15` | Nearby late experimental days are combined to support a stable late-timepoint summary. | Timepoint labels used in figures and research-question tables. | The merge improves interpretability but removes a one-day distinction. |
| Broad cell-type annotation | Marker-set scores for broad immune-cell identities | PBMC identity is a major source of expression variation; broad labels help separate identity-like patterns from activity-like programs. | Composition summaries and identity/activity pattern classification. | Broad labels are useful for interpretation but are not fine-grained cell-state annotations. |
| Discovery cohort | Experimental cohort only | The experimental infection cohort has aligned days, making it the appropriate primary time course for temporal program discovery. | CoGAPS model input for the standard path. | Natural infection samples are not used as the primary discovery model here. |
| Discovery-set balancing | Up to `600` cells per sample | Balancing reduces the chance that one sample or timepoint dominates the learned patterns. | Number of discovery cells used for CoGAPS. | This is not an abundance estimate for the full source study. |
| Random seed | `seed = 42` for subset construction | Makes the balanced discovery subset reproducible. | Which cells are selected when a sample has more than 600 cells. | A different seed could select a slightly different subset. |
| Natural cohort | Reserved as an optional projection target | Keeps the primary lesson focused on the aligned experimental time course. | Optional extension, not required for the main case study. | Projection requires separate validation before being treated as a main result. |

## Full Reproduction Path: Read the Preprocessing Workflow in Pieces

The code excerpts below mirror the stages in the workflow diagram. They are shown with `eval: false`, so you can read them as tutorial code without rebuilding the dataset during the standard path. Each subsection starts with the biological or analytical reason for the step, then shows how the R and Python scripts implement that idea.

The R tab appears first because R is the default analysis path in this case study. The Python tab is shown in parallel for learners who prefer or need the Python implementation.

You have already seen several programming ideas in earlier sections, so this section does not define each one again. The table below names the recurring moves and focuses on why each move matters for preprocessing.

| Programming move | Where it appears | Why it matters |
|---|---|---|
| Start from labeled containers | imported single-cell objects | keeps counts, cell metadata, and gene metadata aligned during preprocessing |
| Use Boolean filters | quality control | makes cell and gene retention rules explicit |
| Rank and subset features | highly variable genes | defines the gene space used for pattern discovery |
| Score gene sets | marker and program scores | connects expression measurements to biological interpretation |
| Sample within groups | balanced discovery subset | prevents one sample from dominating the model input |
| Write a manifest | reproducible outputs | records enough information to explain or reproduce the data products |

### Step 1: Remove Low-Quality Cells and Rare Genes

Quality control is not about making the dataset look cleaner for its own sake. It protects the latent-factor analysis from learning patterns that are mostly low RNA recovery, damaged cells, or extreme count profiles. In a PBMC infection study, that matters because a technical stress pattern can be mistaken for an immune-response program if it is not filtered or at least reported.

The full-preprocessing script uses several filters at the same time: minimum detected genes, minimum counts, maximum mitochondrial percentage, sample-specific upper outliers, and a minimum number of cells per gene. The first small chunk names those thresholds so the choices are visible when the workflow is reported or rerun.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
MIN_GENES <- 200L
MIN_COUNTS <- 500L
MAX_PCT_MT <- 20
MIN_CELLS_PER_GENE <- 10L
```

#### Python

```{python}
#| eval: false
MIN_GENES = 200
MIN_COUNTS = 500
MAX_PCT_MT = 20.0
MIN_CELLS_PER_GENE = 10
```

:::

The next chunk calculates cell-level quality-control metrics. These metrics are not biological conclusions. They are warning signals that help identify cells whose transcriptomes may be too sparse, too extreme, or too enriched for mitochondrial reads to support reliable pattern discovery.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
qc <- scuttle::perCellQCMetrics(
  sce_all,
  subsets = list(Mito = which(is_mito))
)

sce_all$n_genes_by_counts <- qc$detected
sce_all$total_counts <- qc$sum
sce_all$pct_counts_mt <- qc$subsets_Mito_percent
```

#### Python

```{python}
#| eval: false
sc.pp.calculate_qc_metrics(
    adata_all,
    qc_vars=["mt", "ribo", "hb"],
    log1p=False,
    inplace=True,
)
```

:::

Those metrics become the filters that define which cells stay in the full-reproduction object. The Boolean mask pattern has appeared before, so the important point here is interpretive: every `TRUE` or `True` value means a cell passes all of the stated QC rules.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
keep_cells <- sce_all$n_genes_by_counts >= MIN_GENES &
  sce_all$total_counts >= MIN_COUNTS &
  sce_all$pct_counts_mt <= MAX_PCT_MT &
  !sce_all$high_genes_outlier &
  !sce_all$high_counts_outlier

sce_qc <- sce_all[, keep_cells]
```

#### Python

```{python}
#| eval: false
keep_cells = (
    (adata_all.obs["n_genes_by_counts"] >= MIN_GENES)
    & (adata_all.obs["total_counts"] >= MIN_COUNTS)
    & (adata_all.obs["pct_counts_mt"] <= MAX_PCT_MT)
    & (~adata_all.obs["high_genes_outlier"])
    & (~adata_all.obs["high_counts_outlier"])
)

adata_qc = adata_all[keep_cells].copy()
```

:::

The final QC filter removes genes detected in too few cells. This makes the downstream model less sensitive to isolated, hard-to-interpret genes, while still leaving open the limitation that very rare biology can be missed by a strict threshold.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
keep_genes <- Matrix::rowSums(counts(sce_qc) > 0) >= MIN_CELLS_PER_GENE
sce_qc <- sce_qc[keep_genes, ]
```

#### Python

```{python}
#| eval: false
sc.pp.filter_genes(adata_qc, min_cells=MIN_CELLS_PER_GENE)
```

:::

The filter values are deliberately visible because they affect interpretation. If a learner changes these thresholds during full reproduction, the resulting CoGAPS patterns may change too.

:::{.think_question}
::::{.think_question_header}
Checkpoint
::::
::::{.inner_block}
Which filter is most directly aimed at damaged or stressed cells? Which filter could remove rare but real biology if it were made too strict?

<details>
<summary>Show answer</summary>

The mitochondrial fraction filter is most directly aimed at damaged or stressed cells, because high mitochondrial signal can be a sign that a cell is compromised. The minimum-cells-per-gene filter could remove rare but real biology if it were too strict, especially genes expressed only in a small immune-cell population or transient activation state. This is why QC thresholds should be reported as analysis decisions rather than treated as invisible defaults.

</details>
::::
:::

### Step 2: Normalize Expression and Select Variable Genes

After QC, cells still differ in sequencing depth. Library-size normalization and log transformation make expression values more comparable before marker scoring and variable-gene selection. This does not make the data perfect, but it reduces a major technical scale difference.

Highly variable gene selection then focuses the model on genes that carry useful variation across cells and timepoints. CoGAPS can run on many genes, but using 5,000 highly variable genes keeps the model tractable while preserving the major transcriptional structure needed for temporal program discovery.

The first chunk names the gene-selection target. This value defines the gene space available to the selected-model workflow.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
N_HVGS <- 5000L
```

#### Python

```{python}
#| eval: false
N_HVGS = 5000
```

:::

Next, the expression values are scaled and log-transformed. This step makes cells more comparable before the workflow asks which genes vary most across the dataset.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
sce_qc <- scuttle::logNormCounts(sce_qc)
```

#### Python

```{python}
#| eval: false
sc.pp.normalize_total(adata_qc, target_sum = 1e4)
sc.pp.log1p(adata_qc)
```

:::

Finally, the workflow keeps highly variable genes. The R excerpt shows the decision with a simple variance ranking; the Python script uses Scanpy's highly variable gene routine with `sample_id` as a batch key.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
select_hvgs <- function(sce, n_hvgs) {
  logmat <- as.matrix(logcounts(sce))
  vars <- apply(logmat, 1L, stats::var)
  keep <- order(vars, decreasing = TRUE)[seq_len(min(n_hvgs, length(vars)))]
  sce[keep, ]
}

sce_hvg <- select_hvgs(sce_qc, N_HVGS)
```

#### Python

```{python}
#| eval: false
sc.pp.highly_variable_genes(
    adata_qc,
    n_top_genes=N_HVGS,
    flavor="seurat_v3",
    layer="counts",
    batch_key="sample_id",
)

adata_hvg = adata_qc[:, adata_qc.var["highly_variable"]].copy()
```

:::

The Python script uses the Scanpy highly variable gene workflow. The R script shows the same analysis decision with a simpler variance-ranking implementation. Because gene selection can affect learned patterns, the exact method should be documented whenever the full model is rerun.

:::{.think_question}
::::{.think_question_header}
Checkpoint
::::
::::{.inner_block}
Highly variable genes may include infection-response genes, but they can also emphasize stable PBMC identity differences. Why will you need cell-type and timepoint summaries before naming a CoGAPS pattern biologically?

<details>
<summary>Show answer</summary>

Highly variable genes capture strong variation, but strong variation is not automatically infection biology. In PBMC data, many high-variance signals come from stable immune-cell identities, such as T cells, B cells, monocytes, or plasmablasts. You need timepoint summaries to see whether a pattern changes across infection days, and cell-type summaries to see whether the same pattern mostly separates broad immune-cell identities.

</details>
::::
:::

### Step 3: Attach Biological Context with Marker Scores

CoGAPS learns patterns from the expression matrix, but those pattern labels are numerical at first. A pattern called `Pattern3` does not tell you whether it is related to T cells, monocytes, interferon signaling, or something else. Marker and program scores provide biological anchors before you interpret the CoGAPS patterns.

A marker score is a simple gene-set score. It asks whether each cell expresses a small group of genes associated with a broad PBMC identity, such as T cells, monocytes, natural killer cells, B cells, plasmablasts, dendritic cells, or neutrophils. A program score uses the same idea, but the gene set represents a biological activity rather than a cell identity. In this case study, the most important program score is the interferon, or IFN, program score.

These marker lists are not learned from the data. They are manually specified biological reference gene sets. The broad PBMC marker sets are canonical markers chosen to support coarse immune-cell interpretation and to match the major leukocyte populations discussed in [Waickman et al. 2021](https://doi.org/10.1371/journal.ppat.1009240){target="_blank"}; they are not copied as an exact marker table from that paper. The interferon program is also manually specified, but it is more directly anchored in the source study's dengue-response biology: Waickman et al. report interferon-induced and inflammation-associated genes such as `IFITM1`, `IFI6`, `ISG15`, `MX1`, `TRIM22`, and `LY6E` as prominent genes induced during the acute response.

| Score type | Example score | Where the genes come from | What it helps you ask | Important limitation |
|---|---|---|---|---|
| Broad marker score | `T_cell_score`, `Monocyte_score` | Manually curated canonical PBMC marker genes | Does this cell resemble a broad immune-cell identity? | It is an annotation aid, not a definitive cell-type call. |
| Program score | `ifn_program_score` | Manually specified response genes, with the IFN list anchored in Waickman et al.'s acute dengue-response genes and common interferon-stimulated genes | Does this cell express an antiviral or inflammatory response program? | It does not prove that a CoGAPS pattern is caused by interferon signaling. |

This step is not final cell typing. It gives you guideposts for asking whether a pattern is mostly stable PBMC identity, dynamic infection activity, or a mixture of both.

First, the script defines broad PBMC marker sets. These are short reference lists used for coarse interpretation, not a fine-grained annotation atlas.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
BROAD_MARKERS <- list(
  Monocyte = c("LST1", "FCN1", "CTSS", "SAT1", "TYMP", "S100A8", "S100A9"),
  T_cell = c("CD3D", "CD3E", "TRAC", "LTB", "IL7R", "MALAT1"),
  NK_cell = c("NKG7", "GNLY", "PRF1", "CTSW", "KLRD1", "TYROBP"),
  B_cell = c("MS4A1", "CD79A", "CD79B", "CD74", "HLA-DRA", "BANK1"),
  Plasmablast = c("MZB1", "JCHAIN", "XBP1", "SDC1", "IGHG1", "IGKC"),
  Dendritic = c("FCER1A", "CST3", "CLEC10A", "CD1C", "HLA-DRA"),
  Neutrophil = c("FCGR3B", "CXCR2", "CSF3R", "MNDA", "S100A8", "S100A9")
)
```

#### Python

```{python}
#| eval: false
BROAD_MARKERS = {
    "Monocyte": ["LST1", "FCN1", "CTSS", "SAT1", "TYMP", "S100A8", "S100A9"],
    "T_cell": ["CD3D", "CD3E", "TRAC", "LTB", "IL7R", "MALAT1"],
    "NK_cell": ["NKG7", "GNLY", "PRF1", "CTSW", "KLRD1", "TYROBP"],
    "B_cell": ["MS4A1", "CD79A", "CD79B", "CD74", "HLA-DRA", "BANK1"],
    "Plasmablast": ["MZB1", "JCHAIN", "XBP1", "SDC1", "IGHG1", "IGKC"],
    "Dendritic": ["FCER1A", "CST3", "CLEC10A", "CD1C", "HLA-DRA"],
    "Neutrophil": ["FCGR3B", "CXCR2", "CSF3R", "MNDA", "S100A8", "S100A9"],
}
```

:::

The response-program lists use the same scoring idea, but they point to biological activity rather than broad immune-cell identity. The IFN program is included because acute dengue responses in the source study are strongly connected to interferon-induced and inflammation-associated genes.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
PROGRAMS <- list(
  ifn_program = c(
    "IFITM1", "IFI6", "ISG15", "MX1", "IFIT1", "IFIT3",
    "IFI44L", "ISG20", "LY6E", "TRIM22", "OAS1", "OASL"
  ),
  plasmablast_program = c("MZB1", "JCHAIN", "XBP1", "SDC1", "IGHG1", "IGKC"),
  translation_program = c("RPL4", "RPL5", "RPL6", "RPS3", "RPS8", "EEF2"),
  mito_program = c("MT-CYB", "MT-ND4", "MT-ND1", "MT-CO1", "MT-CO2")
)
```

#### Python

```{python}
#| eval: false
PROGRAMS = {
    "ifn_program": [
        "IFITM1", "IFI6", "ISG15", "MX1", "IFIT1", "IFIT3",
        "IFI44L", "ISG20", "LY6E", "TRIM22", "OAS1", "OASL",
    ],
    "plasmablast_program": ["MZB1", "JCHAIN", "XBP1", "SDC1", "IGHG1", "IGKC"],
    "translation_program": ["RPL4", "RPL5", "RPL6", "RPS3", "RPS8", "EEF2"],
    "mito_program": ["MT-CYB", "MT-ND4", "MT-CO1", "MT-CO2"],
}
```

:::

The helper function repeats the same operation for each named gene set. You have already seen why functions are useful, so focus here on the input and output: the input is a named list of genes, and the output is new cell-level score columns.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
score_gene_sets <- function(sce, gene_sets) {
  logmat <- logcounts(sce)

  for (label in names(gene_sets)) {
    present <- intersect(gene_sets[[label]], rownames(sce))
    if (length(present) >= 2L) {
      colData(sce)[[paste0(label, "_score")]] <-
        Matrix::colMeans(logmat[present, , drop = FALSE])
    }
  }

  sce
}
```

#### Python

```{python}
#| eval: false
def score_gene_sets(adata, gene_sets):
    for label, genes in gene_sets.items():
        present = [gene for gene in genes if gene in adata.var_names]
        if len(present) >= 2:
            sc.tl.score_genes(
                adata,
                present,
                score_name=f"{label}_score",
                use_raw=False,
            )
```

:::

After the marker and program lists are defined, applying the scores is intentionally short. The new columns become annotation aids for later comparisons with CoGAPS cell scores and top genes.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
sce_qc <- score_gene_sets(sce_qc, BROAD_MARKERS)
sce_qc <- score_gene_sets(sce_qc, PROGRAMS)
```

#### Python

```{python}
#| eval: false
score_gene_sets(adata_qc, BROAD_MARKERS)
score_gene_sets(adata_qc, PROGRAMS)
```

:::

:::{.definition_box}
::::{.definition_box_header}
What the marker-score columns mean
::::
::::{.inner_block}
The function keeps only genes that are present in the filtered single-cell object. This matters because a marker gene may be absent after filtering or highly variable gene selection.

In the R version, the score is the average log-normalized expression of the present genes in each cell. In the Python version, `scanpy.tl.score_genes()` computes an analogous per-cell gene-set score and stores it in `adata.obs`. Either way, the output is a new cell-level metadata column, such as `Monocyte_score` or `ifn_program_score`.

The score tells you whether each cell expresses a group of genes associated with a broad PBMC identity or response program. It should be treated as an annotation aid, not as ground truth. You still need to compare these scores with CoGAPS pattern usage, top genes, time trajectories, and diagnostics before assigning a biological label to a pattern.
::::
:::

The important idea is that marker and program sets turn many individual gene measurements into interpretable columns. Those columns are later summarized by timepoint and broad cell type so you can separate PBMC composition from infection-associated activity.

After this step, the workflow has biological guideposts attached to cells. Those guideposts are not proof that a CoGAPS pattern is a T-cell program, monocyte program, or interferon-response program. They are evidence to compare against pattern scores, top genes, and temporal trajectories.

### Step 4: Construct the Balanced Experimental Discovery Set

The experimental cohort is the main discovery cohort because its timepoints are aligned across subjects. The natural infection cohort is biologically interesting, but it is less controlled for the primary temporal discovery question and is better handled as an optional projection target.

Within the experimental cohort, samples are balanced so that no sample contributes more than 600 cells. This supports fairer comparison across subjects and timepoints. The tradeoff is important: the discovery set is designed for modeling temporal programs, not for estimating the original abundance of PBMC cell types in the full source study.

The first chunk records the balancing rule and random seed. These values are part of the model input definition, so they should be easy to find in a reproduction script.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
DISCOVERY_MAX_CELLS_PER_SAMPLE <- 600L
RANDOM_SEED <- 42L
```

#### Python

```{python}
#| eval: false
DISCOVERY_MAX_CELLS_PER_SAMPLE = 600
RANDOM_SEED = 42
```

:::

Next, the full-preprocessing workflow keeps the experimental cohort for the primary time-course model. This is a biological design choice: the experimental samples have aligned days, while the natural infection samples are reserved for optional projection.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
sce_exp <- sce_hvg[, sce_hvg$cohort == "experimental"]
```

#### Python

```{python}
#| eval: false
adata_exp = adata_hvg[adata_hvg.obs["cohort"] == "experimental"].copy()
```

:::

The grouped sampling function uses a programming pattern you have already seen in summaries: split by a grouping variable, do one operation within each group, and combine the result. Here the operation is sampling cells within each sample, not calculating a table summary.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
balanced_subset <- function(sce, max_cells_per_sample, seed) {
  set.seed(seed)

  keep <- unlist(lapply(split(seq_len(ncol(sce)), sce$sample_id), function(idx) {
    if (length(idx) <= max_cells_per_sample) {
      idx
    } else {
      sample(idx, max_cells_per_sample)
    }
  }), use.names = FALSE)

  sce[, keep]
}
```

#### Python

```{python}
#| eval: false
def balanced_sample_by_group(
    adata,
    groupby="sample_id",
    max_cells_per_group=DISCOVERY_MAX_CELLS_PER_SAMPLE,
    seed=RANDOM_SEED,
):
    rng = np.random.default_rng(seed)
    keep = []

    for _, idx in adata.obs.groupby(groupby, observed=True).groups.items():
        idx_list = list(idx)
        if len(idx_list) <= max_cells_per_group:
            keep.extend(idx_list)
        else:
            keep.extend(
                rng.choice(idx_list, max_cells_per_group, replace=False).tolist()
            )

    return adata[keep].copy()
```

:::

The last chunk applies the sampling rule. After this point, the discovery object is suitable for CoGAPS model input, but it should not be used to estimate original cell-type abundance in the source study.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
sce_discovery <- balanced_subset(
  sce_exp,
  DISCOVERY_MAX_CELLS_PER_SAMPLE,
  RANDOM_SEED
)
```

#### Python

```{python}
#| eval: false
adata_discovery = balanced_sample_by_group(adata_exp)
```

:::

This grouped sampling step keeps the model from being dominated by the largest sample.

:::{.think_question}
::::{.think_question_header}
Checkpoint
::::
::::{.inner_block}
The balanced discovery set is good for modeling temporal programs. Why would it be inappropriate to use this same balanced subset to estimate the original abundance of T cells, monocytes, or B cells in the full source study?

<details>
<summary>Show answer</summary>

The subset was intentionally balanced so that each sample contributes up to the same number of cells. That helps CoGAPS learn patterns without being dominated by the largest samples, but it changes the original sample sizes. Because abundance questions depend on the original number and proportion of cells, the balanced subset is appropriate for pattern discovery but not for estimating cell-type abundance in the full source study.

</details>
::::
:::

### Step 5: Save Reproducible Outputs

The final preprocessing step records the outputs and their shapes. Large single-cell objects are useful for full reproduction, while smaller CSV and JSON files are better for the standard path. A manifest connects those two worlds by recording where files came from, how large they are, and which parameters were used.

This is not just bookkeeping. If two learners rerun preprocessing with different seeds or different HVG rules, the manifest helps explain why their model inputs may differ.

The large objects are useful when you want to rerun later full-reproduction scripts. They are not needed for the standard path because the case study includes smaller derived tables.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
saveRDS(sce_hvg, file.path(workdir, "gse154386_preprocessed_hvg.rds"))

saveRDS(
  sce_discovery,
  file.path(workdir, "gse154386_experimental_discovery_genes_x_cells.rds")
)
```

#### Python

```{python}
#| eval: false
adata_hvg.write_h5ad(workdir / "gse154386_preprocessed_hvg.h5ad")

adata_discovery_cogaps.write_h5ad(
    workdir / "gse154386_experimental_discovery_cells_x_genes.h5ad"
)
```

:::

The manifest is the compact provenance record. It connects the saved files to the source accession, object shapes, balancing rule, and seed.

::: {.panel-tabset}

#### R

```{r}
#| eval: false
manifest <- list(
  source = "GSE154386",
  import_source = "GSE154386_RAW.tar from GEO",
  hvg_shape = c(cells = ncol(sce_hvg), genes = nrow(sce_hvg)),
  experimental_discovery_shape = c(
    cells = ncol(sce_discovery),
    genes = nrow(sce_discovery)
  ),
  max_cells_per_sample = DISCOVERY_MAX_CELLS_PER_SAMPLE,
  seed = RANDOM_SEED
)

jsonlite::write_json(
  manifest,
  file.path(results_dir, "preprocessing_manifest.json"),
  pretty = TRUE,
  auto_unbox = TRUE
)
```

#### Python

```{python}
#| eval: false
write_manifest(
    results_dir / "preprocessing_manifest.json",
    {
        "source": "GSE154386",
        "import_source": "GSE154386_RAW.tar from GEO",
        "hvg_shape": [int(adata_hvg.n_obs), int(adata_hvg.n_vars)],
        "experimental_discovery_shape": [
            int(adata_discovery_cogaps.n_obs),
            int(adata_discovery_cogaps.n_vars),
        ],
        "max_cells_per_sample": args.max_cells_per_sample,
        "seed": args.seed,
    },
)
```

:::

After this step, the preprocessing workflow has two kinds of outputs. The large single-cell objects preserve the full reproduction path. The smaller CSV and JSON files are the portable evidence tables used in the rest of the case study.

## Full Reproduction Path: Complete Optional Preparation Scripts

The excerpts in **Data Import** and in this section are easier to learn from than a full script. If you want to run the optional preparation workflow, run one of the complete scripts below instead of copying the excerpts one at a time.

Each script is the executable version of the workflow you just walked through. It downloads or reuses the GEO source archive, imports the 10x files into a labeled single-cell object, applies the preprocessing decisions described above, constructs the balanced experimental discovery set, and saves large local objects plus a manifest. It prepares the inputs needed before CoGAPS is run; it does not run CoGAPS itself and it does not rerun the K-selection sweep.

You can skip this block on a first read. The full scripts are displayed as source code for reproducibility, but they do not run during the standard path.

::: {.panel-tabset}

### R

Run this from the repository root inside the case-study Docker image after downloading the large reproduction files from Zenodo, or locally in an equivalent R/Bioconductor environment:

```{bash}
#| eval: false
Rscript scripts/prepare_gse154386_subset_r.R \
  --workdir data/external/GSE154386 \
  --max-cells-per-sample 600 \
  --n-hvgs 5000 \
  --seed 42
```

```{r}
#| label: show-r-prep-script
#| echo: false
#| results: asis
cat("<details><summary>Complete R preparation script</summary>\n\n")
cat("```r\n")
cat(readLines(here::here("scripts", "prepare_gse154386_subset_r.R")), sep = "\n")
cat("\n```\n\n</details>\n")
```

### Python

Run this from the repository root inside the case-study Docker image after downloading the large reproduction files from Zenodo, or locally in an equivalent Python environment:

```{bash}
#| eval: false
python scripts/prepare_gse154386_subset_python.py \
  --workdir data/external/GSE154386 \
  --max-cells-per-sample 600 \
  --n-hvgs 5000 \
  --seed 42
```

```{r}
#| label: show-python-prep-script
#| echo: false
#| results: asis
cat("<details><summary>Complete Python preparation script</summary>\n\n")
cat("```python\n")
cat(readLines(here::here("scripts", "prepare_gse154386_subset_python.py")), sep = "\n")
cat("\n```\n\n</details>\n")
```

:::

## Standard Path: Return to the Case-Study Files

From this point forward, you return to the small files already included in `data/processed`. You are no longer preprocessing raw single-cell data. Instead, you are wrangling selected-model outputs into tables that support biological interpretation.

:::{.reference_box}
::::{.reference_box_header}
You can run these chunks now
::::
::::{.inner_block}
The code below is part of the **standard path**. It uses included selected-model files, so it should run during the ordinary case-study workflow. Full-reproduction learners can run the same checks after they regenerate equivalent files.
::::
:::

| Starting point | Programming move | Resulting evidence |
|---|---|---|
| cell-level CoGAPS scores | group by subject, timepoint, and pattern | subject-timepoint means |
| subject-timepoint means | reshape from long to wide | readable time-course tables |
| pattern summary metrics | filter, rank, and compare | identity/activity evidence table |

### Step 1: Read Subject-Level Pattern Summaries

Cell-level CoGAPS scores are useful, but cells from the same subject and timepoint are not independent people. If you treated thousands of cells as thousands of independent biological replicates, the analysis would exaggerate the amount of independent evidence. The subject-timepoint summary is the better evidence unit for temporal claims in this case study.

The first standard-path chunk reads a prepared long table. Each row represents one subject at one infection day for one CoGAPS pattern.

::: {.panel-tabset}

#### R

```{r}
#| label: subject-time-read-r
subject_time <- readr::read_csv(
  here::here("data", "processed", "interpretation", "subject_level",
             "subject_time_pattern_means_long.csv"),
  show_col_types = FALSE
)
```

#### Python

```{python}
#| label: subject-time-read-python
subject_time = pd.read_csv(
    "data/processed/interpretation/subject_level/"
    "subject_time_pattern_means_long.csv"
)
```

:::

Now inspect `Pattern3`, the candidate temporal program introduced in Data Exploration. The output keeps `n_cells` visible so you can see how many cells contributed to each subject-timepoint summary.

::: {.panel-tabset}

#### R

```{r}
#| label: subject-summary-r
subject_time |>
  filter(pattern == "Pattern3") |>
  select(subject, timepoint_merged, day_merged_numeric,
         n_cells, mean_pattern_score) |>
  arrange(subject, day_merged_numeric) |>
  head(12) |>
  knitr::kable(digits = 3)
```

#### Python

```{python}
#| label: subject-summary-python
subject_time.loc[
    subject_time["pattern"].eq("Pattern3"),
    ["subject", "timepoint_merged", "day_merged_numeric",
     "n_cells", "mean_pattern_score"]
].sort_values(["subject", "day_merged_numeric"]).head(12)
```

:::

This table shows why the case study often works at the subject-timepoint level: it keeps the time course visible while avoiding the mistake of treating every cell as an independent subject.

### Step 2: Reshape Pattern3 into a Readable Time Course

The next chunk reshapes the same information so subjects become rows and infection days become columns. You have already seen reshaping in earlier sections, so focus here on why the display changes: the wide table makes each subject's time course easier to scan.

::: {.panel-tabset}

#### R

```{r}
#| label: pattern3-subject-wide-r
pattern3_subject_wide <- subject_time |>
  filter(pattern == "Pattern3") |>
  mutate(timepoint_merged = factor(timepoint_merged, levels = timepoint_order)) |>
  select(subject, timepoint_merged, day_merged_numeric, mean_pattern_score) |>
  arrange(subject, day_merged_numeric) |>
  select(-day_merged_numeric) |>
  tidyr::pivot_wider(
    names_from = timepoint_merged,
    values_from = mean_pattern_score
  )

pattern3_subject_wide |>
  knitr::kable(digits = 3)
```

#### Python

```{python}
#| label: pattern3-subject-wide-python
pattern3_subject_wide_py = (
    subject_time
    .loc[subject_time["pattern"].eq("Pattern3"),
         ["subject", "timepoint_merged", "day_merged_numeric",
          "mean_pattern_score"]]
    .sort_values(["subject", "day_merged_numeric"])
    .pivot(index="subject",
           columns="timepoint_merged",
           values="mean_pattern_score")
    .loc[:, py_timepoint_order]
    .reset_index()
)

pattern3_subject_wide_py.round(3)
```

:::

The wide view is a quick check on the working hypothesis from Data Exploration: `Pattern3` is low near baseline and increases during the late acute window. You will still need formal summaries, diagnostics, and directionality checks before treating it as the interferon-associated program.

:::{.think_question}
::::{.think_question_header}
Checkpoint
::::
::::{.inner_block}
For a temporal claim, what is the evidence unit: one cell, one subject, or one subject at one timepoint? Why does that choice matter for how confident the figure should look?

<details>
<summary>Show answer</summary>

For the temporal claims in this case study, the evidence unit is a subject at one timepoint. Individual cells are useful measurements, but cells from the same subject and day are not independent biological replicates. Summarizing by subject-timepoint keeps the repeated-subject design visible and prevents the analysis from looking more precise than the study design supports.

</details>
::::
:::

### Step 3: Prepare the Identity/Activity Table

The final wrangling task prepares the table used for the identity/activity research question. This table asks whether a pattern changes mostly across infection days, mostly across broad immune-cell identities, or across both. That distinction matters because a pattern can look temporal if the mix of PBMC cell types changes over time, even if the pattern itself is mostly a cell-identity signal.

The heuristic compares two summaries. `eta_timepoint` measures how much pattern variation is associated with infection day. `eta_broad_cell_type` measures how much pattern variation is associated with broad PBMC identity. A pattern with stronger timepoint variation is a better candidate for dynamic infection activity, while a pattern with stronger cell-type variation is more likely to represent stable PBMC identity or composition.

The summary table already stores this classification for each selected `K = 10` pattern. Here you inspect the prepared table so the later analysis can focus on interpretation rather than table construction.

::: {.panel-tabset}

#### R

```{r}
#| label: identity-activity-table-r
rq2 <- readr::read_csv(
  here::here("data", "processed", "interpretation", "rq_tables",
             "RQ2_identity_vs_activity.csv"),
  show_col_types = FALSE
)
```

#### Python

```{python}
#| label: identity-activity-table-python
rq2_py = pd.read_csv(
    "data/processed/interpretation/rq_tables/"
    "RQ2_identity_vs_activity.csv"
)
```

:::

First, count how many selected-model patterns fall into each broad interpretation class.

::: {.panel-tabset}

#### R

```{r}
#| label: identity-activity-count-r
rq2 |>
  count(pattern_class) |>
  knitr::kable()
```

#### Python

```{python}
#| label: identity-activity-count-python
rq2_py["pattern_class"].value_counts().rename_axis(
    "pattern_class"
).reset_index(name="n")
```

:::

Next, rank patterns by the timepoint effect size. This preview prepares the identity/activity question without yet assigning a final biological label.

::: {.panel-tabset}

#### R

```{r}
#| label: identity-activity-preview-r
rq2 |>
  select(pattern, eta_timepoint, eta_broad_cell_type,
         pattern_class, dominant_broad_cell_type,
         subject_peak_timepoint) |>
  arrange(desc(eta_timepoint)) |>
  head(6) |>
  knitr::kable(digits = 3)
```

#### Python

```{python}
#| label: identity-activity-preview-python
(
    rq2_py
    .loc[:, ["pattern", "eta_timepoint", "eta_broad_cell_type",
             "pattern_class", "dominant_broad_cell_type",
             "subject_peak_timepoint"]]
    .sort_values("eta_timepoint", ascending=False)
    .head(6)
)
```

:::

In the selected `K = 10` run, most patterns are classified as identity-like. One pattern is classified as activity-like, which makes it the main candidate for the infection-response program explored later. This classification is a guide, not a substitute for biological interpretation: you still need top genes, subject trajectories, directionality, and model diagnostics.

:::{.think_question}
::::{.think_question_header}
Checkpoint
::::
::::{.inner_block}
If a pattern has high cell-type variation but low timepoint variation, what kind of biological story would be most cautious? If a pattern has high timepoint variation, what additional evidence would you want before calling it an infection-response program?

<details>
<summary>Show answer</summary>

A pattern with high cell-type variation but low timepoint variation should be described cautiously as identity-like or composition-associated, not as a dynamic infection-response program. A pattern with high timepoint variation is a stronger candidate for infection-response activity, but it still needs supporting evidence. You would want to inspect top genes, subject-level trajectories, directionality relative to baseline, and model diagnostics before assigning a biological label.

</details>
::::
:::

:::{.definition_box}
::::{.definition_box_header}
Key terms from this section
::::
::::{.inner_block}
- **Cell metadata:** Information attached to each cell, such as sample ID, subject, cohort, infection day, quality-control metrics, and broad cell-type annotation.
- **Gene metadata:** Information attached to each gene, such as gene symbol, gene ID, feature type, or whether the gene was selected as highly variable.
- **Quality control:** Filtering and diagnostic checks used to remove cells or genes that are unlikely to support reliable biological interpretation.
- **Boolean mask:** A vector of `TRUE`/`FALSE` values in R or `True`/`False` values in Python used to keep or remove cells or genes.
- **Library-size normalization:** A scaling step that reduces differences caused by cells having different total sequencing counts.
- **Log transformation:** A transformation that compresses large expression values so comparisons are less dominated by the most abundant transcripts.
- **Highly variable genes:** Genes selected because they vary substantially across cells or samples and are therefore informative for learning latent patterns.
- **Marker score:** A summary score based on a known set of marker genes. Marker scores help connect expression data to broad immune-cell identities or biological programs.
- **Program score:** A gene-set score for a biological activity rather than a broad cell identity. In this case study, `ifn_program_score` asks whether cells express an interferon-associated antiviral response gene set.
- **Canonical PBMC marker:** A gene commonly used as a reference signal for a broad peripheral blood mononuclear cell identity, such as T cells, B cells, NK cells, or monocytes. These markers support coarse annotation, but they are not definitive cell-type labels.
- **Interferon-stimulated gene:** A gene whose expression can increase when cells respond to interferon signaling. Interferon-stimulated genes are part of antiviral immune responses and help anchor the IFN-associated interpretation in this case study.
- **Annotation aid:** A supporting label or score used to guide interpretation. Annotation aids help you reason about patterns, but they should be checked against additional evidence before being treated as biological conclusions.
- **Balanced discovery subset:** The experimental subset used for CoGAPS, with up to 600 cells per sample so that one sample does not dominate the learned patterns.
- **Subject-level summary:** A table that summarizes cell-level values within a subject and timepoint, so temporal claims are tied to biological sampling units rather than to individual cells alone.
- **Manifest:** A small metadata file that records where outputs came from, which parameters were used, and how files relate to one another.
::::
:::

By the end of this section, you have followed the path from an imported single-cell object to analysis-ready model inputs, then from cell-level CoGAPS scores to subject-level summaries. The next section uses those summaries to visualize temporal programs, cell-identity structure, and the candidate IFN-associated response.

***
