In this section, you will load the packages used for the required analysis path. This path uses precomputed teaching artifacts included in the repository, so you can focus on importing, summarizing, visualizing, and interpreting the selected model. If you want to reproduce the full CoGAPS runs, the same Docker image also contains the larger R and Python packages needed for that optional workflow.

If you are using Docker, launch the runtime using the command in the **Data, Runtime, and Reproducibility** block near the top of the page. Then open RStudio Server and run either the R or Python setup tab below.

::: {.panel-tabset}

## R

```{r}
#| label: setup-r-packages
#| message: false
library(here)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(jsonlite)

if (requireNamespace("reticulate", quietly = TRUE) &&
    file.exists("/opt/cogaps-venv/bin/python")) {
  reticulate::use_python("/opt/cogaps-venv/bin/python", required = FALSE)
}
```

## Python

```{python}
#| label: setup-python-packages
from pathlib import Path
import json

import pandas as pd
import matplotlib.pyplot as plt
```

:::

The R and Python tabs are alternate ways to complete the same required analysis. You do not need to run both languages unless you want to inspect both examples.

Package | Language | Role | Use
--- | --- | --- | ---
[`here`](https://here.r-lib.org/){target="_blank"} | R | Required analysis | Build project-relative paths.
[`readr`](https://readr.tidyverse.org/){target="_blank"} | R | Required analysis | Read CSV tables used by the R code.
[`dplyr`](https://dplyr.tidyverse.org/){target="_blank"} | R | Required analysis | Select, filter, summarize, and arrange analysis tables.
[`tidyr`](https://tidyr.tidyverse.org/){target="_blank"} | R | Required analysis | Reshape pattern summaries into long and wide forms.
[`ggplot2`](https://ggplot2.tidyverse.org/){target="_blank"} | R | Required analysis | Build visualizations from tidy tables.
[`knitr`](https://yihui.org/knitr/){target="_blank"} | R | Required analysis | Render tables in the case-study document.
[`jsonlite`](https://jeroen.r-universe.dev/jsonlite){target="_blank"} | R | Required analysis | Read run manifests and diagnostic JSON files.
[`reticulate`](https://rstudio.github.io/reticulate/){target="_blank"} | R | Optional bridge | Connect R to the Python environment inside the Docker image.
[`pandas`](https://pandas.pydata.org/){target="_blank"} | Python | Required analysis | Read, filter, and summarize CSV tables.
[`matplotlib`](https://matplotlib.org/){target="_blank"} | Python | Required analysis | Build Python visualizations.
[`CoGAPS`](https://bioconductor.org/packages/CoGAPS){target="_blank"} | R | Optional full reproduction | Run the R implementation of CoGAPS.
[`PyCoGAPS`](https://github.com/FertigLab/pycogaps){target="_blank"} | Python | Optional full reproduction | Run the Python implementation of CoGAPS.
[`scanpy`](https://scanpy.readthedocs.io/){target="_blank"} | Python | Optional full reproduction | Preprocess and inspect single-cell objects.
[`anndata`](https://anndata.readthedocs.io/){target="_blank"} | Python | Optional full reproduction | Read and write H5AD single-cell objects.

You can verify the runtime with these lightweight checks:

::: {.panel-tabset}

## R

```{r}
#| label: runtime-check-r
#| message: false
#| warning: false
r_required <- c("here", "readr", "dplyr", "tidyr", "ggplot2", "knitr", "jsonlite")

r_versions <- vapply(
  r_required,
  function(pkg) as.character(utils::packageVersion(pkg)),
  character(1)
)

data.frame(
  item = c("R", r_required, "reticulate_available", "CoGAPS_available"),
  value = c(
    R.version.string,
    unname(r_versions),
    as.character(requireNamespace("reticulate", quietly = TRUE)),
    as.character(requireNamespace("CoGAPS", quietly = TRUE))
  )
) |>
  knitr::kable()
```

## Python

```{python}
#| label: runtime-check-python
import importlib.util
import sys

import matplotlib

pd.DataFrame(
    {
        "item": [
            "Python",
            "pandas",
            "matplotlib",
            "PyCoGAPS_available",
            "scanpy_available",
            "anndata_available",
        ],
        "value": [
            sys.version.split()[0],
            pd.__version__,
            matplotlib.__version__,
            str(importlib.util.find_spec("PyCoGAPS") is not None),
            str(importlib.util.find_spec("scanpy") is not None),
            str(importlib.util.find_spec("anndata") is not None),
        ],
    }
)
```

:::

The required-analysis packages should be available in whichever language you choose. The optional reproduction packages are checked here so you can confirm that the Docker image is also ready for full model reruns.

When a function is introduced for the first time, the prose uses the `package::function()` form so you can see which package provides it. Next, you will import the precomputed teaching artifacts and inspect the files that carry the selected `K = 10` model results through the rest of the case study.
