Most sections use small CSV, JSON, and figure artifacts that are already included in this repository. You can work through the required analysis in either R or Python. The case-study Docker image provides the same RStudio Server environment used to render this site and also includes the heavier packages needed for optional full reproduction.

If you are using Docker, launch the runtime using the command in the **Data, Runtime, and Reproducibility** block near the top of the page. Then open RStudio Server and run either the R or Python setup tab below.

::: {.panel-tabset}

## R

```{r}
#| label: setup-r-packages
#| message: false
library(here)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(jsonlite)

if (requireNamespace("reticulate", quietly = TRUE) &&
    file.exists("/opt/cogaps-venv/bin/python")) {
  reticulate::use_python("/opt/cogaps-venv/bin/python", required = FALSE)
}
```

## Python

```{python}
#| label: setup-python-packages
from pathlib import Path
import json

import pandas as pd
import matplotlib.pyplot as plt
```

:::

The R and Python tabs are alternate ways to complete the same required analysis. You do not need to run both languages unless you want to inspect both examples.

Package | Language | Required for required path? | Use
--- | --- | --- | ---
[`here`](https://here.r-lib.org/){target="_blank"} | R | Yes | Build project-relative paths.
[`readr`](https://readr.tidyverse.org/){target="_blank"} | R | Yes | Read CSV tables used in the rendered R code.
[`dplyr`](https://dplyr.tidyverse.org/){target="_blank"} | R | Yes | Select, filter, summarize, and arrange R analysis tables.
[`tidyr`](https://tidyr.tidyverse.org/){target="_blank"} | R | Yes | Reshape pattern summaries into long and wide forms.
[`ggplot2`](https://ggplot2.tidyverse.org/){target="_blank"} | R | Yes | Build and customize visualizations from tidy tables.
[`knitr`](https://yihui.org/knitr/){target="_blank"} | R | Yes | Render tables in the case-study document.
[`jsonlite`](https://jeroen.r-universe.dev/jsonlite){target="_blank"} | R | Yes | Read run manifests and diagnostic JSON files.
[`reticulate`](https://rstudio.github.io/reticulate/){target="_blank"} | R | Optional | Connect R to the Python environment inside the Docker image.
[`pandas`](https://pandas.pydata.org/){target="_blank"} | Python | Yes | Read, filter, and summarize CSV tables.
[`matplotlib`](https://matplotlib.org/){target="_blank"} | Python | Yes | Build Python visualizations.
[`CoGAPS`](https://bioconductor.org/packages/CoGAPS){target="_blank"} | R | Optional full reproduction | Run the R implementation of CoGAPS.
[`PyCoGAPS`](https://github.com/FertigLab/pycogaps){target="_blank"} | Python | Optional full reproduction | Run the Python implementation of CoGAPS.
[`scanpy`](https://scanpy.readthedocs.io/){target="_blank"} | Python | Optional full reproduction | Preprocess and inspect single-cell objects.
[`anndata`](https://anndata.readthedocs.io/){target="_blank"} | Python | Optional full reproduction | Read and write H5AD single-cell objects.

You can verify the rendered runtime with these lightweight checks:

::: {.panel-tabset}

## R

```{r}
#| label: runtime-check-r
#| message: false
#| warning: false
data.frame(
  item = c("R", "dplyr", "ggplot2", "CoGAPS_available"),
  value = c(
    R.version.string,
    as.character(packageVersion("dplyr")),
    as.character(packageVersion("ggplot2")),
    as.character(requireNamespace("CoGAPS", quietly = TRUE))
  )
) |>
  knitr::kable()
```

## Python

```{python}
#| label: runtime-check-python
import importlib.util
import sys

pd.DataFrame(
    {
        "item": ["Python", "pandas", "matplotlib", "PyCoGAPS_available", "scanpy_available", "anndata_available"],
        "value": [
            sys.version.split()[0],
            pd.__version__,
            __import__("matplotlib").__version__,
            str(importlib.util.find_spec("PyCoGAPS") is not None),
            str(importlib.util.find_spec("scanpy") is not None),
            str(importlib.util.find_spec("anndata") is not None),
        ],
    }
)
```

:::

When a function is introduced for the first time, the prose uses the `package::function()` form so you can see which package provides it. The **Additional Information** section records more detailed session information for the rendered environment.
